### 2020-09-30 susan_classification.Rmd
after-meeting update and then initial start...

---
title: "Initial Graph-by-Graph classification"
author: "Susan Vanderplas"
date: "9/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F, fig.path = "figure")
```

```{r data-setup, include = F}
# Download NYT covid data to create graphs which have a consistent look (removing outlet style from the equation)

# Read in the data from NYT
library(tidyverse)
library(lubridate)
library(zoo)
library(ggrepel)
library(gridExtra)
cvstate <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")

cvstate <- cvstate %>%
  group_by(state, fips) %>%
  mutate(date_100th_case = min(date[cases >= 100]),
         days_100th_case = date - date_100th_case,
         daily_cases = cases - lag(cases, 1, default = 0),
         week_avg_cases = rollmean(cases, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
         daily_avg_cases = rollmean(daily_cases, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
         daily_deaths = deaths - lag(deaths, 1, default = 0),
         week_avg_deaths = rollmean(deaths, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
         daily_avg_deaths = rollmean(daily_deaths, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T)) %>%
  ungroup() %>%
  mutate(state = factor(state))

download.file("https://www2.census.gov/programs-surveys/popest/tables/2010-2019/state/totals/nst-est2019-01.xlsx", destfile = "census_pop_state.xlsx", quiet = T, mode = "wb")
pop_data <- readxl::read_xlsx("census_pop_state.xlsx", col_names = c("state", "pop2019"), col_types = c("text", rep("skip", 11), "numeric"), skip = 9) %>%
  mutate(state = str_remove(state, "[[:punct:]]"))

cvstate <- left_join(cvstate, pop_data, by = "state")

my_theme <- theme_bw() + 
  theme(plot.background = element_rect(fill = "transparent", color = NA), 
        panel.background = element_rect(fill = "transparent", color = NA), 
        legend.background = element_rect(fill = "transparent", color = NA),
        axis.text.y = element_text(angle = 90, hjust = 0.5)) 
theme_set(my_theme)

```

# Maps

# Time Series

## Case Counts

```{r}
show_states <- c("California", "Texas", "New York", "Arizona", "Iowa", "Washington")
tmp <- cvstate %>%
  filter(state %in% show_states) %>%
  filter(date <= ymd("2020-03-15"))

tmpsum <- tmp %>% group_by(state) %>% 
  filter(daily_avg_cases == max(daily_avg_cases)) %>%
  filter(date == min(date))

ggplot() + 
  geom_line(data = tmp, 
            aes(x = date, y = daily_avg_cases, color = state)) + 
  geom_text_repel(data = tmpsum, 
                  aes(x = date, y = daily_avg_cases, 
                      label = state, color = state), 
                  hjust = 0, nudge_x = 1) + 
  scale_color_manual(values = color_palette) + 
  guides(color = F) + 
  scale_x_date(expand = expansion(mult = c(.05, .2)), 
               date_breaks = "month", date_labels = "%B") + 
  scale_y_continuous("Daily cases (7 day rolling average)") + 
  my_theme + theme(axis.title.x = element_blank())
```

### 2022-01-04 index.Rmd
Add an index file that uses jdssv format

---
documentclass: jdssv
author:
  - name: Susan Vanderplas
    affiliation: University of Nebraska, Lincoln
    # use this syntax to add text on several lines
    address: |
      | Statistics Department
      | 350 Hardin Hall North Wing
      | 3310 Holdrege St
      | University of Nebraska, Lincoln
      | Lincoln, NE 68583-0963, United States
    email: \email{svanderplas2@unl.edu}
    url: https://srvanderplas.netlify.app
  - name: Adalbert Wilhelm
    affiliation: Jacobs University
    # use this syntax to add text on several lines
    address: |
      | some address
    email: \email{a.wilhelm@jacobs-university.de}
    url: https://www.jacobs-university.de/directory/awilhelm # This link does not work...
title:
  formatted: "Covid Graphics" # We need a better title, sigh.
  # If you use tex in the formatted title, also supply version without
  plain:     "Guidelines for the Journal of Data Science, Statistics, and Visualisation Version 0.3"
  # For running headers, if needed
  short:     "Covid Graphics"
abstract: >
    Abstract text here
keywords:
  # at least one keyword must be supplied
  formatted: [COVID-19, visualization]
  plain:     [COVID-19, visualization]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{booktabs}
  \newcommand{\class}[1]{`\code{#1}'}
  \newcommand{\fct}[1]{\code{#1()}}
  \newcommand{\ma}[1]{\ensuremath{\mathbf{#1}}}
output: rticles::jss_article
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
```

\newpage
# Computational Details {-}

<!-- If necessary or useful, information about certain computational details -->
<!-- such as version numbers, operating systems, or compilers could be included -->
<!-- in an unnumbered section. Also, auxiliary packages (say, for visualizations, -->
<!-- maps, tables, \dots) that are not cited in the main text can be credited here. -->

<!-- The results in this paper were obtained using -->
<!-- \proglang{R}~3.5.1. \proglang{R} itself -->
<!-- and all packages used are available from the Comprehensive -->
<!-- \proglang{R} Archive Network (CRAN) at -->
<!-- \url{https://CRAN.R-project.org/}. -->

# Acknowledgments {-}

<!-- All acknowledgments should be collected in this -->
<!-- unnumbered section before the references. It may contain the usual information -->
<!-- about funding and feedback from colleagues/reviewers/etc. Furthermore, -->
<!-- information such as relative contributions of the authors may be added here -->
<!-- (if any). -->

\bibliography{refs}

\newpage

\setcounter{section}{0}
\renewcommand{\thesection}{\Alph{section}}

# More Technical Details {#app:technical}

<!-- Appendices can be included after the bibliography (with a page break). Each section within the appendix should have a proper section title (rather than just \emph{Appendix}). -->

### 2022-01-04 JDSSV_manuscript.Rmd
some updates from my side

---
documentclass: jdssv
author:
  - name: Susan Vanderplas
    affiliation: University of Nebraska–Lincoln
    address: |
      | Department of Statistics
      | University of Nebraska–Lincoln
      | 349A Hardin Hall 
      | Lincoln, NE 68583-0963, USA
    email: \email{susan.vanderplas@unl.edu}
    url: https://statistics.unl.edu/susan-vanderplas
  - name: Adalbert F. X. Wilhelm
    affiliation: Jacobs University Bremen
    # To add another line, use \AND at the end of the previous one as above
    address: |
      |  Department of Psychology and Methods
      |  Jacobs University Bremen gGmbH
      |  Campus Ring 1
      |  28759 Bremen
    email: \email{a.wilhelm@jacobs-university.de}
    url: http://www.jacobs-university.de/directory/wilhelm
    # use a different affiliation in adress field (differently formated here)
    # affiliation2: Universitat Autònoma de Barcelona
title:
  formatted: "Visual narratives of the Covid-19 pandemic"
  # If you use tex in the formatted title, also supply version without
  plain:     "Visual narratives of the Covid-19 pandemic"
  # For running headers, if needed
  short:     "Visual narratives of Covid-19"
abstract: >
    Covid-19 has created a world-wide interest in understanding the dynamics of a pandemic. Hence, numerous media outlets as well as researchers have produced comprehensive data visualizations to illustrate the relevant trends and figures. In this paper, we will look at an elective choice of Covid-19 data visualizations to evaluate and discuss currently established visualization tools in their capacity to provide a communication channel both within the data science team and also between data analyst, domain experts and a more general interested audience. While there is no fixed catalogue of evaluation criteria for data visualizations we will try to provide an overview on the different core aspects of visualization evaluation and their competing principles.
keywords:
  # exploratory data visualisation
  formatted: [JDSSV, style guidelines, comma-separated, not capitalized, "\\proglang{R}"]
  plain:     [JDSSV, style guidelines, comma-separated, not capitalized, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{booktabs}
  \newcommand{\class}[1]{`\code{#1}'}
  \newcommand{\fct}[1]{\code{#1()}}
  \newcommand{\ma}[1]{\ensuremath{\mathbf{#1}}}
output: rticles::jss_article
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
```

# Mission

JDSSV\footnote{This document is an adaptation of the style guide of the Journal of Statistical Software \citep{jss_style_guide}.} is an international refereed journal that creates a forum to present recent progress and ideas in the different disciplines of data science, statistics, and visualisation. It welcomes contributions to data science, statistics, and visualisation, and in particular those aspects which link and integrate these subject areas. Articles can cover topics such as machine learning and statistical learning, the visualisation and verbalisation of data, visual analytics, big data infrastructures and analytics, interactive learning, and advanced computing. Papers that discuss two or more research areas of the journal are favoured. Scientific contributions should be of a high standard. Articles should be oriented towards a wide scientific audience of statisticians, data scientists, computer scientists, data analysts, etc.

The journal welcomes original contributions that are not being considered for publication elsewhere and contain a high level of novelty. Papers with a thorough but concise review of a certain topic with the potential to provide new insights are also welcome. Manuscripts submitted to the journal generally are accompanied by supplementary material that containing software code, technical derivations or detailed explanations, additional examples, etc. All submitted material will be reviewed by the assigned associate editor and reviewers of the manuscript.

Manuscripts may have a substantial theoretical component, but it is expected that all manuscripts contain at least one application on empirical or simulated data. The journal emphasizes the reproducibility of the results presented its papers. Therefore, all data and software code that is necessary to reproduce the empirical results in the manuscript should be made available in a user friendly manner. If the empirical data cannot be released for reasons of confidentiality or otherwise, then a generated dataset with comparable properties should be provided.

# Preparing your Manuscript for Submission

All submissions to JDSSV should be written in \LaTeX\ using the JDSSV style files provided on the journal website \url{https://jdssv.org}. For initial submission it suffices providing the pdf version of the manuscript together with all the necessary supplementary material including software code to reproduce all results in the manuscript. The final version of accepted manuscripts should adhere to all the style guidelines and should incorporate all changes requested by the production editor. All \LaTeX\  source files of the final manuscript should be submitted and accepted manuscripts are only published if these files comply with all guidelines and instructions provided by the journal.

# Software

The journal expects that submissions contain accompanying
software with the aim of reproducibility of the results and application of the proposed methodology to other data by the reader.  All existing software used in the paper should be properly referenced. Code should be delivered in an easily readable manner with clear instructions on its use, and preferably is accompanied by instructive examples of its use. We highly recommend to provide code that can be used in open-source software such as \proglang{R} \citep{R}, \proglang{Python} \citep{python}, \proglang{Julia} \citep{Julia}, \proglang{Octave} \citep{octave}, etc.

To make code widely accessible, we advise making it available in a repository such as \url{https://zenodo.org} where it will receive a permanent Digital Object Identifier (DOI) which can be included in the manuscript.

[comment]: # (\todo{\tiny Alternatives: (1) We store the data/code on our JDSSV website with the article. (2) We upload it on Zenodo ourselves (not sure if that can be done).)

The provided code should at least consist of a file or set of files for the functions that execute the core of the method. For reproducibility, another necessary file is the script that creates the results (tables and figures) of the paper. For methods that run very long, provide a toy example that runs sufficiently fast and highlights the properties of the method. Nonproprietary data should be provided or a permanent link to these data should be given. The corresponding script to analyze the data should read these data.

\begin{sloppypar}
For increased readability, please apply the following naming conventions for code: Start function names with a verb, e.g., \code{set_initialisation()}, \code{compute_loss()}, \code{update_X()}, etc. when appropriate. Give  objects descriptive names or closely follow the notation in the paper.
For programming conventions (particularly in \proglang{R}), see Hadley Wickham's guidelines (\url{http://r-pkgs.had.co.nz/style.html}) and the use of the \pkg{styler} package is recommended. For \proglang{python}, consider Google's recommendations (\url{https://github.com/google/styleguide/blob/gh-pages/pyguide.md}) Sufficient comments should be added to the code to make it understandable. For \proglang{Octave} and \proglang{MatLab} \citep{matlab}, consider Richard Johnson's MatLab Style Guidelines 2.0 (\url{https://www.mathworks.com/matlabcentral/fileexchange/46056-matlab-style-guidelines-2-0}).
\end{sloppypar}

For writing about software JDSSV requires authors to use the markup
\verb|\proglang{}| (programming languages and large programmable systems), \verb|\pkg{}| (software packages), \verb|\code{}| (functions, commands, arguments, etc.). If there is such markup in (sub)section titles (as above), a plain text version has to be provided in the \LaTeX\ command as well. Below we also illustrate how abbrevations should be introduced and citation commands can be employed. See the \LaTeX\ code for more details.

# Review Process

All manuscripts should be submitted online at  \url{https://jdssv.org}. JDSSV uses a single blind review process. Upon submission of their manuscript, authors have the opportunity to provide a short list of suggested reviewers as well as a few names of researchers that preferably should not be contacted for reviewing the manuscript.  All submitted manuscript will undergo automatic checking for plagiarism and will not be considered for further review in case of plagiarism. The manuscript will be assigned to one of the editors who will make an initial screening to check the quality of the submitted work, possibly with the help of an associate editor. After positive screening, the manuscript will be assigned to an associate editor who will seek the opinion of at least two reviewers. Reviewers are asked to send their report within one month. Associate editors should typically make their recommendation one week after reception of the review reports. The initial review process should normally not take more than three months. All communication between the authors and the journal will be administered by the journal editors.

# Correspondence Analysis {#sec:models}

As an example, we provide a simple implementation of correspondence analysis \citep[see, e.g.,][]{greenacre2010biplots}. For the analysis of bivariate categorical data or other tables that contain counts, correspondence analysis can be a useful technique to visualize important relations between the categories. Consider word count data on explanations of data science by several websources provided by \citet{Lubbe2018}, see Table \ref{tab:word_counts}.

\begin{table}
\caption{Part of the table with word counts describing data science by different web sources as collected by \citet{Lubbe2018}.}
\label{tab:word_counts}
\begin{center}
\begin{tabular}{lccccccc} \hline
            & dsc.test & dsctechniques & datadiversity & dsc.tips & $\ldots$ & wikipedia \\ \hline
statistics  &       6  &           2   &          2    &    0     & $\ldots$ &  2 \\
big data    &       2  &           0   &          3    &    5     & $\ldots$ &  0 \\
analytics   &       1  &           0   &          2    &    0     & $\ldots$ &  2 \\
database    &       2  &           0   &          0    &    2     & $\ldots$ &  1 \\
insight     &       0  &           0   &          4    &    0     & $\ldots$ &  2 \\
$\vdots$    & $\vdots$ &    $\vdots$   &   $\vdots$    & $\vdots$ &$ \ddots$ & $\vdots$ \\
programming &       7  &           0   &          0    &    0     & $\ldots$ &  0 \\
\hline
\end{tabular}
\end{center}
\end{table}

Let $\ma{F}$ be the matrix containing the values in Table \ref{tab:word_counts}. The most simple model to fit such a table is the independence model in matrix $\ma{E}$ defined as 
\begin{eqnarray*}
  \ma{E} = n^{-1}\ma{D}_r \ma{11}^\top \ma{D}_c
\end{eqnarray*}
with $\ma{1}$ a vector of ones of appropriate length, $\ma{D}_r = \textrm{Diag}(\ma{F1})$ the diagonal matrix with row sums of $\ma{F}$, $\ma{D}_c = \textrm{Diag}(\ma{F^\top1})$ the diagonal matrix with column sums of $\ma{F}$, and $n = \ma{1^\top F1}$ be the sum of all elements in $\ma{F}$. The goal of correspondence analysis is to find a matrices of row and column coordinates $\ma{R}$ and $\ma{C}$ such that
\begin{eqnarray*}
  L(\ma{R}, \ma{C}) = \|\ma{D}_r^{-1/2} (\ma{F} - \ma{E} - \ma{D}_r\ma{RC}\ma{D}_c^\top) \ma{D}_c^{-1/2}\|^2
\end{eqnarray*}
is minimized \citep[see, for example,][]{vandeveldenetal2009seriation}. It may be verified that the least-squares optimal solution is obtained by computing the singular value decomposition (SVD) 
\begin{eqnarray*}
  \ma{D}_r^{-1/2} (\ma{F} - \ma{E})\ma{D}_c^{-1/2} = \ma{UDV}^\top
\end{eqnarray*}
and computing
\begin{eqnarray*}
  \ma{R} &=& n^{1/2}\ma{D}_r^{-1/2} \ma{UD}^{\alpha}\\
  \ma{C} &=& n^{1/2}\ma{D}_c^{-1/2} \ma{VD}^{1-\alpha}
\end{eqnarray*}
for some $\alpha$. Note that the least-squares optimal rank $p$ solution is obtained by taking the first $p$ columns of $\ma{R}$ and $\ma{C}$. If $\alpha = 1$ the so-called row principal solution is obtained where the row points are the weighted centroids of the column points, $\alpha = 0$ gives the column principal solution with column points being the weighted average of the row points, and $\alpha = 1/2$ results in the symmetric solution. The importance measure of the dimensions are given by the inertia, that is, the diagonal elements of $\ma{D}^2$. For more details, see, for example, \cite{greenacre2010biplots}.

A simple implementation of correspondence analysis in \proglang{R} is given by the function \code{corana()} given by

```{r corana, echo=TRUE}
corana <- function(dat, alpha = 0.5){
  # Perform correspondence analysis
  # Input: 
  ##  dat   numeric matrix dat with nonnegative entries
  ##  alpha = 1 is row principal standardisation
  ##        = 0 is column principle standardisation
  ##        = 0.5 is symmetric standardisation
  dat <- as.matrix(dat)
  Dr <- rowSums(dat)
  Dc <- colSums(dat)
  n  <- sum(Dr)
  # Compute expected values under the independence model
  E  <- outer(Dr, Dc)/n
  tt <- svd(diag(Dr^-0.5) %*% (dat - E) %*% diag(Dc^-0.5))
  ## Remove dimensions with singular value zero
  ind <- tt$d > 1e-10
  tt$d <- tt$d[ind]
  tt$u <- tt$u[, ind]
  tt$v <- tt$v[, ind]
  ## Compute row scores R and column scores C 
  R  <- n^0.5 * diag(Dr^-0.5) %*% tt$u %*% diag(tt$d^alpha)
  C  <- n^0.5 * diag(Dc^-0.5) %*% tt$v %*% diag(tt$d^(1 - alpha))
  rownames(R) <- rownames(dat)
  rownames(C) <- colnames(dat)
  ## Compute relative contribution to inertia per dimension 
  row.inert   <- outer(Dr/n, tt$d^(2*alpha),       "/") * R^2
  col.inert   <- outer(Dc/n, tt$d^(2*(1 - alpha)), "/") * C^2
  ## Compute reconstructed Chi-square distance for row and column points
  row.dist    <- R^2 %*% diag(tt$d^(2 - 2*alpha))
  row.dist    <- row.dist / outer(rowSums(row.dist), rep(1, ncol(R))) 
  col.dist    <- C^2 %*% diag(tt$d^(2 - 2*(1 - alpha)))
  col.dist    <- col.dist / outer(rowSums(col.dist), rep(1, ncol(C))) 
  ## Prepare list of results
  out <- list(R = R, C = C, sing.val = tt$d, 
              row.inert = row.inert, col.inert = col.inert,
              row.dist  = row.dist,  col.dist  = col.dist)
  class(out) <- "corana"
  return(out)
}
```
A plot is given by the \code{plot.corana()} method

```{r plot_corana, dependson = "corana", echo = TRUE}
plot.corana <- function(out, dims = 1:2, ...){
  R <- out$R[, dims]
  C <- out$C[, dims]
  ## Set up coordinate system
  coord <- rbind(R, C)
  plot(coord[, 1], coord[, 2], type = "n", asp = 1, las = 1,
       xlab = paste0("Dim ", dims[1]), ylab = paste0("Dim ", dims[2]), ...)
  abline(h = 0, v = 0, col = "gray")
  ## Use reconstructed distance as importance measure for transparency and size
  row.alpha <- rowSums(out$row.dist[, dims])
  row.alpha <- (row.alpha/max(row.alpha))^0.7
  col.alpha <- rowSums(out$col.dist[, dims])
  col.alpha <- (col.alpha/max(col.alpha))^0.7
  ## Plot row and column points 
  points(R[, 1], R[, 2], pch = 20, col = rgb(1, 0, 0, row.alpha))
  points(C[, 1], C[, 2], pch = 20, col = rgb(0, 0, 1, col.alpha))
  ## Write text labels
  text(R[, 1], R[, 2], rownames(R), pos = compute.pos(R), cex = 2*row.alpha,
       col = rgb(1, 0, 0, ifelse(row.alpha > 0.1, row.alpha, 0)))
  text(C[, 1], C[, 2], rownames(C), pos = compute.pos(R), cex = 2*col.alpha,
       col = rgb(0, 0, 1, ifelse(col.alpha > 0.1, col.alpha, 0)))
}
```

Assuming that \code{ds.word.cnt} contains the full matrix of which a part is shown in Table \ref*{tab:word_counts}, then the following code does a correspondence analysis on these data:
```{r ca_biplot_dims23, dependson = "plot_corana", echo = TRUE, fig.width = 6, fig.height = 6, out.width = "10cm", fig.align = "center", fig.cap = "Biplot of a correspondence analysis on word counts used in descriptions of data science by different web sources in dimensions 2 and 3."}
load("ds.word.cnt.RData")
source("compute.pos.R")
out.corana <- corana(ds.word.cnt)
plot(out.corana, dims = 2:3)
```
that yields the plot of Dimensions 2 and 3 in Figure \ref{fig:ca_biplot_dims23}. In this plot, the points and labels are made more transparent and the labels decrease in size as they are represented worse in these two dimensions.

\newpage
# Computational Details {-}

If necessary or useful, information about certain computational details
such as version numbers, operating systems, or compilers could be included
in an unnumbered section. Also, auxiliary packages (say, for visualizations,
maps, tables, \dots) that are not cited in the main text can be credited here.

The results in this paper were obtained using
\proglang{R}~3.5.1. \proglang{R} itself
and all packages used are available from the Comprehensive
\proglang{R} Archive Network (CRAN) at
\url{https://CRAN.R-project.org/}.

# Acknowledgments {-}

All acknowledgments should be collected in this
unnumbered section before the references. It may contain the usual information
about funding and feedback from colleagues/reviewers/etc. Furthermore,
information such as relative contributions of the authors may be added here
(if any).

\bibliography{refs}

\newpage

\setcounter{section}{0}
\renewcommand{\thesection}{\Alph{section}}

# More Technical Details {#app:technical}

Appendices can be included after the bibliography (with a page break). Each section within the appendix should have a proper section title (rather than just \emph{Appendix}).

\section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}

References need to be provided in a \textsc{Bib}\TeX\ file (\code{.bib}). All references should be made with \verb|\cite|, \verb|\citet|, \verb|\citep|, \verb|\citealp| etc.\ (and never hard-coded). This commands yield different formats of author-year citations and allow to include additional details (e.g., pages, chapters, \dots) in brackets. 
%In case you are not familiar with these commands see the JSS style FAQ for details.

Cleaning up \textsc{Bib}\TeX\ files is a somewhat tedious task -- especially when acquiring the entries automatically from mixed online sources. However, it is important that informations are complete and presented in a consistent style to avoid confusions. JDSSV requires the following format.

* Specific markup (\verb|\proglang|, \verb|\pkg|, \verb|\code|) should be used in the references.
* Titles should be inserted in title case.
* Journal titles should not be abbreviated and in title case.
* DOIs should be included where available.
* Software should be properly cited as well. For \proglang{R} packages \code{citation("pkgname")} typically provides a good starting point.

\newpage

### 2022-01-04 JDSSV_manuscript_knitr.Rnw
some updates from my side

\documentclass[article]{jdssv}

%% -- LaTeX packages and custom commands ---------------------------------------

%% recommended packages
\usepackage{thumbpdf,lmodern}
\usepackage{bm}
\usepackage{todonotes} %% Remove this package when finishing the manuscript

%% another package (only for this demo article)
\usepackage{framed}

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\newcommand{\ma}[1]{\ensuremath{\mathbf{#1}}}

%% -- Article metainformation (author, title, ...) -----------------------------

%% - \author{} with primary affiliation
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{Susan Vanderplas\\University of Nebraska–Lincoln
   \And Adalbert F.X. Wilhelm\\Jacobs University Bremen}
\Plainauthor{Susan Vanderplas, Adalbert F.X. Wilhelm}

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{Visual narratives of the Covid-19 pandemic}
\Plaintitle{Visual narratives of the Covid-19 pandemic}
\Shorttitle{Visual narratives of Covid-19}

%% - \Abstract{} almost as usual
\Abstract{
  Covid-19 has created a world-wide interest in understanding the dynamics of a pandemic. Hence, numerous media outlets as well as researchers have produced comprehensive data visualizations to illustrate the relevant trends and figures. In this paper, we will look at an elective choice of Covid-19 data visualizations to evaluate and discuss currently established visualization tools in their capacity to provide a communication channel both within the data science team and also between data analyst, domain experts and a more general interested audience. While there is no fixed catalogue of evaluation criteria for data visualizations we will try to provide an overview on the different core aspects of visualization evaluation and their competing principles.
  }

%% - \Keywords{} with LaTeX markup, at least one required
%% - \Plainkeywords{} without LaTeX markup (if necessary)
%% - Should be comma-separated and in sentence case.
\Keywords{exploratory data visualisation, logarithmic scales, visual comparisons, \proglang{R}}
\Plainkeywords{exploratory data visualisation, logarithmic scales, visual comparisons, R}

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
    email: \email{susan.vanderplas@unl.edu}
  Susan Vanderplas\\
  Department of Statistics\\
  University of Nebraska–Lincoln\\
  349A Hardin Hall \\
  Lincoln, NE 68583-0963, USA\\
  E-mail: \email{susan.vanderplas@unl.edu}\\
  URL: \url{https://statistics.unl.edu/susan-vanderplas}\\
\ \\
%}
%
%\Address{
  Adalbert F.X. Wilhelm\\
  Department of Psychology and Methods\\
  Jacobs University Bremen gGmbH\\
  Campus Ring 1\\
  28759 Bremen, Germany\\
  E-mail: \email{a.wilhelm@jacobs-university.de}\\
  URL: \url{http://www.jacobs-university.de/directory/wilhelm}
}

\begin{document}
\SweaveOpts{concordance=TRUE}
<<include=FALSE, cache=FALSE, echo=FALSE, purl=FALSE>>=
library(knitr)
opts_chunk$set(comment = NA, size = 'footnotesize', prompt = TRUE, highlight = FALSE, 
               cache = TRUE, crop = TRUE, concordance = FALSE, fig.align='center',
               fig.path='Figures_R/fig-',out.width="0.4\\textwidth")
opts_knit$set(out.format = "latex")
knit_hooks$set(crop = hook_pdfcrop)
options(width = 80, prompt = "R> ", continue = "    ", digits = 3)
thm <- knit_theme$get('default')
Sys.setenv(TEXINPUTS=getwd(),
           BIBINPUTS=getwd(),
           BSTINPUTS=getwd())
knit_theme$set(thm)
@

%% -- Introduction -------------------------------------------------------------

%% - In principle "as usual".
%% - But should typically have some discussion of both _software_ and _methods_.
%% - Use \proglang{}, \pkg{}, and \code{} markup throughout the manuscript.
%% - If such markup is in (sub)section titles, a plain text version has to be
%%   added as well.
%% - All software mentioned should be properly \cite-d.
%% - All abbreviations should be introduced.
%% - Unless the expansions of abbreviations are proper names (like "Journal
%%   of Statistical Software" above) they should be in sentence case (like
%%   "generalized linear models" below).

\section{Introduction}

Here some general introduction

Effective participation in the social dynamics of a democracy requires clear understanding of quantitative evidence and rules by the people. As the medical community, governments, business, and the public come together to fight the COVID-19 pandemic, scientists have a special role in helping to communicate what the data actually mean.
For data analysis and modeling to have an impact as a component of decision making, appropriate reporting and communication is key. There are numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reporting, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org). These standards are based on commonly accepted core quality principles and values such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility. For measures to restrain and overcome an epidemic effectively, communication among experts that follows highest professional and ethical standards is not sufficient. In a democratic society, policy measures can only be implemented if they are based on the acceptance of the wider population. This puts high demands on skills associated with communicating statistical evidence on the side of scientists, governments and media, and a citizenry able to understand statistical messages. %

In recent decades, there have been numerous publications, initiatives, and ideas to improve the communication of quantitative and statistical information, see \cite{Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}, to name only a few. Data journalism has recently taken off as an innovative component of news publishing, and COVID-19 provides numerous excellent examples, often using an interactive visual format on the Internet, such as dashboards. %Nevertheless, both the complexity of the phenomena and the "bipolarity" of statistical thinking remain challenges. 
A fundamental problem in assessing probabilities lies, for example, in the intuitive conflation of subjective risks ("how likely am I to become infected") and general risks ("how likely is it that some person will become infected"). Another issue is that of equating sensitivity of a diagnostic test and the positive predictive value \citep{Eddy1982,Gigerenzer2007,McDowell2019,Binder2020}. In particular, the prevalence (or base rate) is often neglected leading to this confusion. As \cite{Hoffrage} {\color{blue}Hoffrage et al. (2000)} point out “… statistics expressed as natural frequencies improve the statistical thinking of experts and nonexperts alike”. Smooth sliding back and forth between probabilities and natural frequencies is easy once figures are related to a unifying reference value. A well proven method to effective communication lies in representing information in accessible ways (Gigerenzer et al., 2007; Gigerenzer and Edwards, 2011) by saying, for instance, “one in 10” instead of 10\%. Using absolute in place of relative numbers, representing information in appealing graphic forms, and summarizing most relevant facts in “fact boxes” are some of the methods developed and advocated for increasing transparency in communication between stakeholders such as health providers and patients (See, for example: https:// www. hardingcenter.de)
Fact boxes combined with icon arrays are recommended for the presentation of test results. Both representations are based on natural frequencies \citep{[62],Krauss2020} and present case numbers as simply and concretely as possible. Many scientific studies show that icon arrays help people understand numbers and risks more easily (e.g. \cite{McDowell2019}). The Harding Center for Risk Literacy shows many other examples of transparent communication of risks, including COVID-19\footnote{\url{https://www.hardingcenter.de/de/mrna-schutzimpfung-gegen-covid-19-fuer-aeltere-menschen}}.

While human thinking tends towards pattern simplification and political communication also prefers a simple cause-effect relationship, real phenomena are often multivariate. Thus, when studying COVID-19 and predicting its spread, it is not only important to consider its symptomatology, the incidence and geographic distribution of diseases, population behavior patterns, government policies and impacts on the economy, on schools, on people in nursing homes and on social life as a whole, but to integrate these into the data analyses and the communication of results. Associations observed in the data can often be caused by third-party variables (confounders). In addition, much of the data comes from observational studies, which usually makes a robust causal attribution problematic. Statisticians calling out these limitations, however, face the danger that their statements might be pulled out one side in a polarized debate \citep{McConway2021}.

\section[]{The global perspective}

Visual representations take a central position in public communication and aim to represent the corresponding dynamics and contents in a quickly understandable way. Usually either time-dependent parameters or data with a spatial reference are visualised. For spatially distributed data, choropleth maps are predominantly used, in which administrative regions defined by the responsible health authorities are coloured according to the distribution density of the infection figures or variables derived from them (see Figure~\ref{fig:choro}).  Their visual perception problems - such as the visual dominance of the area of administrative regulatory frameworks that have no direct relation to infection events - are well known but still widespread.  In addition, the use of ordinance thresholds as the basis for color scaling is often at odds with color schemes that emphasize real spatial distributional differences.

\begin{figure*}
	\includegraphics[width = 0.9\textwidth]{choroplethen}
	\caption{Choropleth map of the incidence figures for Germany by district. Source: Robert-Koch-Institute https://app.23degrees.io/export/oCRP768wQ3mCswE7-choro-corona-faelle-pro-100-000/image.}
	\label{fig:choro}
\end{figure*}

For time-dependent parameters, different variants of time series diagrams are used, predominantly line and column diagrams. Classic errors of graphical representation, such as overemphasising temporal variability by reducing the value axis to a small section, have now largely disappeared from the media. Switching between line and bar charts for purely design-related reasons in order to produce corresponding graphical diversity nevertheless seems questionable. 
The use of logarithmic scales in time series diagrams should be evaluated with caution. On the one hand, they tempt superficial readers to underestimate dynamic growth processes; on the other hand, they increase the demands on the mathematical and statistical literacy of the readership without corresponding advantages of visual representation. Figure~\ref{fig:incidence} shows the time course of the 7-day incidence per 100,000 people between 24 January and 4 February 2021 for some selected countries. While the differences appear relatively small on the logarithmic scale, the linear scale shows considerable differences.

\begin{figure}
		\includegraphics[width = 0.49 \textwidth]{incidence1.png}
			\includegraphics[width = 0.49 \textwidth]{incidence2.png}
	\caption{The 7-day incidence for different countries over the course of time. On the logarithmic scale (left graphic), differences seem small. The linear scale (right graphic), however, shows considerable differences. Source: Our World in Data, https://ourworldindata.org/covid-cases?country=IND~USA~GBR~CAN~DEU~FRA.}
	\label{fig:incidence}
\end{figure}

Simulations were used particularly illustratively in the media in the course of the pandemic. An inspiring example illustrating the spread of the epidemic appeared as early as 14 March 2020 in the Washington Post\footnote{\url{https://www.washingtonpost.com/graphics/2020/world/corona-simulator/}} with the title "Why outbreaks like coronavirus spread exponentially, and how to flatten the curve". The Washington Post made this simulation available free of charge and in all major languages, which led to it being distributed worldwide, including repeatedly on German television\footnote{\url{https://web.br.de/interaktiv/corona-simulation/}}. The New York Times\footnote{\url{https://www.nytimes.com/interactive/2020/us/coronavirus-spread.html}} published a dynamic graphic entitled "How the Virus Won", which maps the spread of COVID-19 cases from February to June 2020 in the USA. It shows how an analysis of the associations between different COVID-19 strains and travel patterns can help understand the spread of the disease.

Another illustrative example is a simulation from ZEIT Online\footnote{\url{https://www.zeit.de/wissen/gesundheit/2020-11/coronavirus-aerosole-ansteckungsgefahr-infektion-hotspot-innenraeume}}, which - based on models developed by a group of researchers at the Max Planck Institute for Chemistry - estimates the probability of an infected person infecting other people in closed rooms in various scenarios. While the visualisations present the simulated infection processes in a catchy way, the dependence of the simulations on parameter assumptions and settings is usually not addressed. Simulations should also always make transparent on which model assumptions and which data basis the simulations were created.

\

\section[]{Comparisons and rankings}

\section{To log or not to log}

\section{Correspondence Analysis} \label{sec:models}

\section{Summary and discussion} \label{sec:summary}

%% -- Optional special unnumbered sections -------------------------------------
\newpage
\section*{Computational Details}

If necessary or useful, information about certain computational details
such as version numbers, operating systems, or compilers could be included
in an unnumbered section. Also, auxiliary packages (say, for visualizations,
maps, tables, \dots) that are not cited in the main text can be credited here.

The results in this paper were obtained using
\proglang{R}~3.5.1. \proglang{R} itself
and all packages used are available from the Comprehensive
\proglang{R} Archive Network (CRAN) at
\url{https://CRAN.R-project.org/}.

\section*{Acknowledgments}

All acknowledgments should be collected in this
unnumbered section before the references. It may contain the usual information
about funding and feedback from colleagues/reviewers/etc. Furthermore,
information such as relative contributions of the authors may be added here
(if any).

%% -- Bibliography -------------------------------------------------------------
%% - References need to be provided in a .bib BibTeX database.
%% - All references should be made with \cite, \citet, \citep, \citealp etc.
%%   (and never hard-coded). See the FAQ for details.
%% - JSS-specific markup (\proglang, \pkg, \code) should be used in the .bib.
%% - Titles in the .bib should be in title case.
%% - DOIs should be included where available.

\bibliography{references}

%% -- Appendix (if any) --------------------------------------------------------
%% - After the bibliography with page break.
%% - With proper section titles and _not_ just "Appendix".

\newpage

\begin{appendix}

\section{More Technical Details} \label{app:technical}

Appendices can be included after the bibliography (with a page break). Each section within the appendix should have a proper section title (rather than just \emph{Appendix}).

%For more technical style details, please check out JSS's style FAQ at
%\url{https://www.jstatsoft.org/pages/view/style#frequently-asked-questions}
%which includes the following topics:
%\begin{itemize}
%  \item All main words in titles and headers start with a capital.
%  \item Use vectorized graphics formats such as eps and pdf when possible. Graphics should be of high quality while trying to minimize the file size.
%\end{itemize}

\section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}

References need to be provided in a \textsc{Bib}{\TeX} file (\code{.bib}). All references should be made with \verb|\cite|, \verb|\citet|, \verb|\citep|, \verb|\citealp| etc.\ (and never hard-coded). This commands yield different formats of author-year citations and allow to include additional details (e.g., pages, chapters, \dots) in brackets. 
%In case you are not familiar with these commands see the JSS style FAQ for details.

Cleaning up \textsc{Bib}{\TeX} files is a somewhat tedious task -- especially when acquiring the entries automatically from mixed online sources. However, it is important that informations are complete and presented in a consistent style to avoid confusions. JDSSV requires the following format.
\begin{itemize}
  \item Specific markup (\verb|\proglang|, \verb|\pkg|, \verb|\code|) should
    be used in the references.
  \item Titles should be inserted in title case.
  \item Journal titles should not be abbreviated and in title case.
  \item DOIs should be included where available.
  \item Software should be properly cited as well. For \proglang{R} packages
    \code{citation("pkgname")} typically provides a good starting point.
\end{itemize}
\end{appendix}
\newpage
%% -----------------------------------------------------------------------------

\end{document}

### 2022-01-04 JDSSV_editorialprocess.Rmd

--- 2022-01-04 JDSSV_manuscript.Rmd
@@
 author:
-  - name: Susan Vanderplas
-    affiliation: University of Nebraska–Lincoln
+  - name: Patrick J.F. Groenen
+    affiliation: Erasmus University Rotterdam
+    # use this syntax to add text on several lines
     address: |
-      | Department of Statistics
-      | University of Nebraska–Lincoln
-      | 349A Hardin Hall 
-      | Lincoln, NE 68583-0963, USA
-    email: \email{susan.vanderplas@unl.edu}
-    url: https://statistics.unl.edu/susan-vanderplas
-  - name: Adalbert F. X. Wilhelm
-    affiliation: Jacobs University Bremen
+      | Econometric Institute
+      | Erasmus School of Economics
+      | Erasmus University Rotterdam
+      | P.O. Box  1738 
+      | 3000 DR Rotterdam, The Netherlands
+    email: \email{groenen@ese.eur.nl}
+    url: https://personal.eur.nl/groenen/
+  - name: Stefan Van Aelst
+    affiliation: KU Leuven
     # To add another line, use \AND at the end of the previous one as above
     address: |
-      |  Department of Psychology and Methods
-      |  Jacobs University Bremen gGmbH
-      |  Campus Ring 1
-      |  28759 Bremen
-    email: \email{a.wilhelm@jacobs-university.de}
-    url: http://www.jacobs-university.de/directory/wilhelm
+      |  Department of Mathematics
+      |  KU Leuven
+      |  Celestijnenlaan 200B
+      |  3001 Leuven, Belgium
+    email: \email{Stefan.VanAelst@kuleuven.be}
+    url: https://wis.kuleuven.be/statdatascience/robust
     # use a different affiliation in adress field (differently formated here)
@@
 title:
-  formatted: "Visual narratives of the Covid-19 pandemic"
+  formatted: "Guidelines for the Journal of Data Science, Statistics, and Visualisation"
   # If you use tex in the formatted title, also supply version without
-  plain:     "Visual narratives of the Covid-19 pandemic"
+  plain:     "Guidelines for the Journal of Data Science, Statistics, and Visualisation Version 0.3"
   # For running headers, if needed
-  short:     "Visual narratives of Covid-19"
+  short:     "JDSSV Guidelines, Version 0.3"
 abstract: >
-    Covid-19 has created a world-wide interest in understanding the dynamics of a pandemic.
-Hence, numerous media outlets as well as researchers have produced comprehensive data visualizations to illustrate the relevant trends and figures.
-In this paper, we will look at an elective choice of Covid-19 data visualizations to evaluate and discuss currently established visualization tools in their capacity to provide a communication channel both within the data science team and also between data analyst, domain experts and a more general interested audience.
-While there is no fixed catalogue of evaluation criteria for data visualizations we will try to provide an overview on the different core aspects of visualization evaluation and their competing principles.
+    This short article illustrates how to write a manuscript for the   \emph{Journal of Data Science, Statistics and Visualisation} (JDSSV) using its \LaTeX\ style files.
+Please follow JDSSV's style guidelines precisely.
+Also, it is recommended to keep the \LaTeX\ code as simple as possible, that is, avoid inclusion of packages/commands that are not necessary.
 keywords:
-  # exploratory data visualisation
+  # at least one keyword must be supplied
   formatted: [JDSSV, style guidelines, comma-separated, not capitalized, "\\proglang{R}"]

### 2022-01-04 JDSSV_manuscript_knitr.Rnw
Some comments and sketches

@@
 \begin{document}
-\SweaveOpts{concordance=TRUE}
+% \SweaveOpts{concordance=TRUE}
 <<include=FALSE, cache=FALSE, echo=FALSE, purl=FALSE>>=
@@
 
+% May also want to mention at the outset the limitations in the data due to low testing capacity, weekly effects due to reporting, and so on.
+We must be able to represent the data that exist visually while acknowledging the limitations associated with that data.
+
 \section[]{The global perspective}
@@
 
-\
+\section{Time series}
+% Line plots, frequency domain, NY times tile plots w/ cases represented in color intensity
 
 \section[]{Comparisons and rankings}
+% https://twitter.com/AndToddsaid/status/1314231055096971265 is a horrible set of graphs and a great example of the snark associated with covid management policy and geographic comparisons...
 
@@
 
-\section{Correspondence Analysis} \label{sec:models}
+% Discussion of how this section's conclusions change with the pandemic
+
+% Focus primarily on time-series graphs (as opposed to log scales in color)
+
+% Problems with log scales
+
+% Problems with linear scales
+
+% Discussion of sites which allow the user to switch back and forth between scales
 

### 2022-01-05 JDSSV_manuscript_knitr.Rnw
Add reference. Probably need a few more?

@@
 \section{To log or not to log}
-
+As COVID cases grow quasi-exponentially while there are susceptible members of the population^[subject to the effectiveness of mitigation measures and testing availability], it seems natural to use log scales to allow for more effective comparisons of slight changes in case counts over time.
+In addition, log scales make it possible to compare regions with different populations or infection rates in the same chart.
+As noted previously, however, interpreting log scales requires levels of numerical sophistication that may not be appropriate for the general public.
+Even researchers do not always read and interpret log scales correctly\citep{mengeLogarithmicScalesEcological2018}; expecting the general public to do so with emotionally charged content is difficult.
 % Discussion of how this section's conclusions change with the pandemic

### 2022-01-12 JDSSV_manuscript_knitr.Rnw

@@
 
+<<get-nyt-data, include = F>>=
+
+# Read in the data from NYT
+library(tidyverse)
+library(lubridate)
+library(zoo)
+library(ggrepel)
+library(gridExtra)
+cvstate <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")
+
+cvstate <- cvstate %>%
+  group_by(state, fips) %>%
+  mutate(date_100th_case = min(date[cases >= 100]),
+         days_100th_case = date - date_100th_case,
+         daily_cases = cases - lag(cases, 1, default = 0),
+         week_avg_cases = rollmean(cases, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
+         daily_avg_cases = rollmean(daily_cases, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
+         daily_deaths = deaths - lag(deaths, 1, default = 0),
+         week_avg_deaths = rollmean(deaths, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
+         daily_avg_deaths = rollmean(daily_deaths, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T)) %>%
+  ungroup() %>%
+  mutate(state = factor(state))
+
+download.file("https://www2.census.gov/programs-surveys/popest/tables/2010-2019/state/totals/nst-est2019-01.xlsx", destfile = "census_pop_state.xlsx", quiet = T, mode = "wb")
+pop_data <- readxl::read_xlsx("census_pop_state.xlsx", col_names = c("state", "pop2019"), col_types = c("text", rep("skip", 11), "numeric"), skip = 9) %>%
+  mutate(state = str_remove(state, "[[:punct:]]"))
+
+cvstate <- left_join(cvstate, pop_data, by = "state")
+
+my_theme <- theme_bw() + 
+  theme(plot.background = element_rect(fill = "transparent", color = NA), 
+        panel.background = element_rect(fill = "transparent", color = NA), 
+        legend.background = element_rect(fill = "transparent", color = NA),
+        axis.text.y = element_text(angle = 90, hjust = 0.5)) 
+theme_set(my_theme)
+@
+
+<<colorpal, include = F>>=
+# Including this for temporary compatibility - we may want to use *shudder* viridis?
+main_colors <- c("#d00000", "#f5f1e7", "#c7c8ca")
+
+# use as accents
+secondary_colors <- c("#001226", "#249ab5")
+
+# These are only to be used for infographics/charts
+tertiary_colors <- c("#bccb2a", "#f58a1f", "#005d84", "#ffd74f", "#a5228d")
+
+color_palette <- c("#d00000", "#249ab5",  "#1B8500", "#f58a1f", "#a5228d", "#001226", "#B1EB66", "#005d84", "#c25700")
+
+heat_color_palette <- c("#ffd74f", "#f58a1f", "#c25700", "#d00000", "#9D0000", "#6A0000")
+
+blueseqpal <- c("001226", "06293E", "0C3F56", "12566E", "186D85", "1E839D", 
+                "249AB5", 
+                "39ABC1", "4EBCCE", "64CDDA", "79DDE6", "8EEEF3", "A3FFFF") %>% 
+  paste0("#", .) %>% rev()
+
+blueseqpal_lt <- rev(c("#B2FFFF", "#C2FFFF", "#D1FFFF", "#E0FFFF", "#F0FFFF"))
+
+darkblueseqpal <- colorspace::darken(blueseqpal, .2)
+lightblueseqpal <- colorspace::lighten(blueseqpal, .2)
+@
 %% -- Introduction -------------------------------------------------------------
@@
 
+<<freq-domain-plot, include = F, eval = F>>
+show_states <- c("California", "Texas", "New York", "Arizona", "Iowa", "Washington")
+
+n <- 50000
+tmp <- cvstate %>%
+  filter(state %in% show_states) %>%
+  select(state, date, cases) %>%
+  mutate(mod_num = cases %% n,
+         whole_num = floor(cases/n)) %>%
+  group_by(state, whole_num) %>%
+  mutate(whole_num_days = n()) %>%
+  group_by(state) %>%
+  arrange(date)
+
+tmp2 <- bind_rows(
+  # Day before first case detected:
+  filter(tmp, whole_num == 0) %>%
+    group_by(state) %>%
+    summarize(date = min(date) - lubridate::days(1), y = 0, whole_num = 0, 
+              whole_num_days = unique(whole_num_days), mod_num = 0) %>%
+    unique() %>%
+    ungroup(),
+  # Date before the count rolls over
+  tmp %>%
+    group_by(state, whole_num) %>%
+    filter(date == max(date)) %>%
+    group_by(state) %>%
+    mutate(y = ifelse(whole_num == max(whole_num), mod_num, n),
+           whole_num_days = ifelse(whole_num == max(whole_num), whole_num_days/(y/10000), whole_num_days)),
+  # Date after the count rolls over
+  tmp %>%
+    group_by(state, whole_num) %>%
+    filter(date == min(date)) %>%
+    mutate(y = 0, date = date - lubridate::days(1))
+  ) %>%
+  group_by(state, whole_num) %>%
+  mutate(whole_num_days = max(whole_num_days))
+
+ggplot(data = tmp2, aes(x = date, ymin = 0, ymax = y, 
+                        color = whole_num_days, 
+                        fill = whole_num_days, 
+                        group = whole_num)) + 
+  geom_ribbon() + 
+  facet_grid(state~.) + 
+  scale_x_date(date_breaks = "month", date_labels = "%B") +
+  scale_y_continuous(breaks = c(n/2, n), labels = paste0(c(n/2000, n/1000), "k"))  + 
+  scale_color_gradientn(guide = 'none', paste0("Days to ", n/1000, "k\nmore cases"), space = "Lab", 
+    trans = "log10",
+    colours = darkblueseqpal
+    ) + 
+  scale_fill_gradientn(paste0("Days to ", n/1000, "k\nmore cases"), space = "Lab", 
+    trans = "log10",
+    colours = blueseqpal) + 
+  my_theme + 
+  theme(axis.title = element_blank(), 
+        legend.background = element_rect(fill = "transparent", color = NA), 
+        legend.position = c(0, .66), legend.justification = c(0, 1), legend.direction = "horizontal")
+@
+
 \section[]{Comparisons and rankings}
@@
 % Problems with log scales
+
+<<log-scale-failures>>=
+# ----- Set up labeller to tweak things ----------------------------------------
+rlistmap <- function(x, myfun) {
+  if ("list" %in% class(x)) return(purrr::map(x, ~rlistmap(., myfun)))
+
+  return(myfun(x))
+}
+fix_labels <- function(x) {
+  x  %>%
+    str_replace('scale: (.*)', "\\1 scale") %>%
+    str_replace("baseline: (\\d{1,})", "Baseline:\n\\1 cases")
+}
+label_custom <- function(labels, multi_line = T, sep = ": ") {
+  out <- label_both(labels, multi_line = multi_line, sep = sep)
+
+  out <- rlistmap(out, fix_labels)
+
+  out
+}
+class(label_custom) <- "labeller"
+
+# ----- Set up data ------------------------------------------------------------
+growth <- rpois(30, 15*exp((1:30 - 20)/5))
+
+tmp <- tibble(baseline = 15,
+       x = 1:30,
+       y = rpois(30, 15),
+       y1 = growth) %>%
+  bind_rows(
+    tibble(baseline = 50,
+       x = 1:30,
+       y = rpois(30, 50),
+       y1 = growth) 
+  ) %>%
+  bind_rows(
+    tibble(baseline = 200,
+       x = 1:30,
+       y = rpois(30, 200),
+       y1 = growth) 
+  ) %>%
+  pivot_longer(matches("y")) %>%
+  mutate(name = str_replace_all(name, c("y$" = "Background", "y1" = "Exponential")))  %>%
+  group_by(baseline, x) %>%
+  mutate(Total = sum(value)) %>%
+  pivot_wider(names_from = "name", values_from = "value") %>%
+  pivot_longer(matches("Background|Exponential|Total")) %>%
+  mutate(name = factor(name, levels = c("Total", "Exponential", "Background")))
+
+tmp <- bind_rows(mutate(tmp, scale = "Log"), mutate(tmp, scale = "Linear")) %>%
+  mutate(Baseline = reorder(paste0("Baseline: ", baseline, " cases"), baseline)) %>%
+  mutate(scale = factor(scale, levels = c("Log", "Linear")))
+
+# ----- Set up scales ----------------------------------------------------------
+library(facetscales)
+scales_y <- list(
+  Log = scale_y_log10(),
+  Linear = scale_y_continuous()
+)
+
+tmp %>%
+ggplot(aes(x = x, y = value, color = name, size = name)) + 
+  geom_line(alpha = .9) +
+  facet_grid_sc(rows = vars(scale), cols = vars(Baseline), 
+                scales = list(x = "fixed", y = scales_y)) +
+  scale_size_manual(values = c(1.75, 1, 1)) +
+  scale_color_manual(values = c("#249ab5", "#d00000", "#001226")) +
+  scale_fill_discrete(name = "") + 
+  theme(legend.position = "bottom", axis.title = element_blank(),
+        legend.title = element_blank())
+@
 

### 2022-01-12 JDSSV_manuscript_knitr.Rnw
a few more paragraphs

@@
 
+One issue with assessing the use of log scales is that their effectiveness changes with the stage of the pandemic and the amount (and varieties) of data shown.
+Initially, log scales were incredibly useful at showing case counts, because minimal mitigation measures were in place and the growth of case counts (or presumptive positive cases, in absence of available testing) was fairly close to exponential.
+In addition, the use of log scales allowed for the comparison of nominal cases across entities with large population differences: in the US, we could compare cases in New York and California with cases in Michigan and Washington, even though the population of Michigan and Washington are much lower than the population of either New York or California.
+
+However, after the first wave of COVID, the issues with log scales became more apparent; \Cref{fig:log-scale-failures} demonstrates the problem.
+With a baseline number of cases that occur either because charts show a cumulative case count or because there is some baseline spread level that is not exponential because of mitigation measures ($R_{eff} \leq 1$), a log scale makes it difficult to detect slight increases in case counts that might indicate a new source of uncontrolled spread.
+
+\Cref{fig:log-scale-failures} shows
+
 % Focus primarily on time-series graphs (as opposed to log scales in color)

### 2022-01-15 JDSSV_manuscript_knitr.Rnw
Some more log stuff

@@
 \usepackage{framed}
+
+\usepackage{cleveref}
 
@@
                cache = TRUE, crop = TRUE, concordance = FALSE, fig.align='center',
-               fig.path='Figures_R/fig-',out.width="0.4\\textwidth")
+               fig.path='Figures_R/fig-',out.width="0.4\\textwidth", 
+               message = F, error = F, warning = F, echo = F)
 opts_knit$set(out.format = "latex")
@@
 
-<<freq-domain-plot, include = F, eval = F>>
+<<freq-domain-plot, include = F, eval = F>>=
 show_states <- c("California", "Texas", "New York", "Arizona", "Iowa", "Washington")
@@
 \section{To log or not to log}
-As COVID cases grow quasi-exponentially while there are susceptible members of the population^[subject to the effectiveness of mitigation measures and testing availability], it seems natural to use log scales to allow for more effective comparisons of slight changes in case counts over time.
+As COVID cases grow quasi-exponentially while there are susceptible members of the population (subject to the effectiveness of mitigation measures and testing availability), it seems natural to use log scales to allow for more effective comparisons of slight changes in case counts over time.
 In addition, log scales make it possible to compare regions with different populations or infection rates in the same chart.
 As noted previously, however, interpreting log scales requires levels of numerical sophistication that may not be appropriate for the general public.
-Even researchers do not always read and interpret log scales correctly\citep{mengeLogarithmicScalesEcological2018}; expecting the general public to do so with emotionally charged content is difficult.
+Even researchers do not always read and interpret log scales correctly\citep{mengeLogarithmicScalesEcological2018}; expecting the general public to do so is difficult under normal circumstances\citep{hecklerStudentAccuracyReading2013} is difficult.
+When panic, fear, uncertainty, and doubt about the situation are added to the mix,
 % Discussion of how this section's conclusions change with the pandemic
@@
 
-However, after the first wave of COVID, the issues with log scales became more apparent; \Cref{fig:log-scale-failures} demonstrates the problem.
-With a baseline number of cases that occur either because charts show a cumulative case count or because there is some baseline spread level that is not exponential because of mitigation measures ($R_{eff} \leq 1$), a log scale makes it difficult to detect slight increases in case counts that might indicate a new source of uncontrolled spread.
+<<log-scale-initial, fig.width = 8, fig.height = 4, out.width = "\\linewidth", fig.cap = "In the early stages of the pandemic, log scales allowed the comparison of raw case counts in locations with vastly different population and case counts.">>=
+# remotes::install_github("AllanCameron/geomtextpath")
+library(geomtextpath)
 
-\Cref{fig:log-scale-failures} shows
+library(facetscales)
+scales_y <- list(
+  Log = scale_y_log10(),
+  Linear = scale_y_continuous()
+)
+tmp <- cvstate %>% 
+  filter(date < ymd("2020-06-01")) %>%
+  filter(state %in% c("Michigan", "Washington", "New York", "California"))
 
-% Focus primarily on time-series graphs (as opposed to log scales in color)
+p1 <-  tmp %>%
+  ggplot(aes(x = date, y = daily_avg_cases, group = state)) +
+  geom_textline(aes(color = state, label = state, vjust = -0.1, hjust =.93) , text_smoothing = 50) + 
+  ylab("Daily average cases (7 day average)") + 
+  xlab("Date") + 
+  guides(color = "none") + 
+  scale_y_log10() + 
+  facet_grid(~"Log") + 
+  theme_bw()
+
+p2 <-  tmp %>%
+  ggplot(aes(x = date, y = daily_avg_cases, group = state)) +
+  geom_textline(aes(color = state, label = state, vjust = -0.25, hjust = .85) , text_smoothing = 60) + 
+  ylab("Daily average cases (7 day average)") + 
+  xlab("Date") + 
+  guides(color = "none") + 
+  facet_grid(~"Linear") + 
+  theme_bw() +
+  theme(axis.title.y = element_blank())
+
+grid.arrange(p1, p2, nrow = 1)
+@
+
+While log scales are not necessarily intuitive, many outlets tried to make the graphs more intuitive by adding reference lines, as shown in \Cref{fig:diag-ref-lines}.
+
+\begin{figure}
+\centering
+\includegraphics[width=.8\linewidth]{ft-covid-19-deaths}
+\caption{Reference lines to compare exponential growth rates of deaths in different countries.
+This provides some additional context that may help individuals use log scale data more successfully.
+This approach was first featured in the Financial Times, but was quickly adopted by the New York Times, 91-DIVOC, and other outlets.
+Graph from the Financial Times (March 23, 2020), image from \citet{kosaraPraiseDiagonalReference2020}.}
+\label{fig:diag-ref-lines}
+\end{figure}
 
 % Problems with log scales
+However, after the first wave of COVID, the issues with log scales became more apparent: it was difficult to detect slight increases in case counts that indicated the beginning of a new wave amid a background level of spread, as demonstrated in  \Cref{fig:log-scale-failures}.
+Diagonal reference lines from the origin were also less helpful, as the growth of cases or deaths was no longer approximately exponential and varied over time; for these reference lines to be effective there would need to be a clear idea of when the case counts started to increase exponentially, which is difficult to determine whilst in the thick of a potential COVID wave.
 
-<<log-scale-failures>>=
-# ----- Set up labeller to tweak things ----------------------------------------
+<<log-scale-failures, echo = F, fig.width = 8, fig.height = 4, out.width = ".95\\linewidth", message = F, error = F, warning = F, fig.cap = "One problem with log scales is that if there is a background level of spread, it can be hard to notice the introduction of an additional source of exponential spread.
+Linear scales do not have this problem - the exponential source is noticeable very quickly in the total line, but on the log scale it is much harder to discern when the exponential source causes the total line to diverge from the background.
+In the top-right corner, it is difficult to identify that there is an exponential increase in cases amid the baseline, even though the exponential source makes up approximately 50\\% of the cases at the end of the time period shown." >>=
+# Set up labeller to tweak things 
 rlistmap <- function(x, myfun) {
@@
 
-# ----- Set up data ------------------------------------------------------------
+# Set up data 
 growth <- rpois(30, 15*exp((1:30 - 20)/5))
@@
 
-# ----- Set up scales ----------------------------------------------------------
-library(facetscales)
+# Set up scales 
 scales_y <- list(
@@
   scale_fill_discrete(name = "") + 
+  theme_bw() + 
   theme(legend.position = "bottom", axis.title = element_blank(),
@@
 
+% Focus primarily on time-series graphs (as opposed to log scales in color)
+
 % Problems with linear scales
+While log scales have their problems, linear scales are not immune from issues either.
+It can be very difficult to adequately compare to past situations when looking at the full time series of case counts.
+For example, in \Cref{fig:linear-scales-ref-lines}, it is difficult to tell whether the first wave of COVID cases in March 2020 had an increase as fast as that in January of 2021; it is even more difficult to compare the order-of-magnitude of change in case rate growth of January 2021 relative to January 2022 when the more contagious omicron variant became prevalent.
+
+<<linear-scales-ref-lines, fig.width = 8, fig.height = 4, out.width = ".8\\linewidth", fig.cap = "Reported COVID cases in New York State, 2020-2022.
+The linear scale makes it difficult to compare the trajectory of different waves to determine how severe the current status is relative to the past, because the primary contrast is the height of the relative peaks, rather than the growth \\emph{rate}.
+A similar graph on the log scale would have the peaks at much more similar heights (though there would still be a difference), allowing the reader to focus on other information.">>=
+cvstate %>%
+  filter(state == "New York") %>%
+  ggplot(aes(x = date, y = daily_avg_cases)) + 
+  geom_line() + 
+  ggtitle("Reported COVID cases in New York State, 2020-2022") +
+  ylab("Daily Cases reported (7-day average)") + 
+  xlab("Date") + theme_bw()
+@
+
+<<log-scales-ref-lines, fig.width = 8, fig.height = 4, out.width = ".8\\linewidth", fig.cap = "Reported COVID cases in New York State, 2020-2022.
+The linear scale makes it difficult to compare the trajectory of different waves to determine how severe the current status is relative to the past.">>=
+cvstate %>%
+  filter(state == "New York") %>%
+  ggplot(aes(x = date, y = daily_avg_cases)) + 
+  geom_line() + 
+  ggtitle("Reported COVID cases in New York State, 2020-2022") +
+  ylab("Daily Cases reported (7-day average)") + 
+  xlab("Date") + theme_bw() + 
+  scale_y_log10()
+@
 
 % Discussion of sites which allow the user to switch back and forth between scales
+
+It is not clear that the use of log or linear scales during the COVID-19 pandemic had a large effect on public opinion.
+Several studies were conducted in the early stages of the pandemic \citep{romanoScaleCOVID19Graphs2020, seviLogarithmicLinearVisualizations2020, ryanGraphsLogarithmicAxes2020} and results seem to suggest that while individuals have difficulty understanding log scale graphs, these issues do not tend to affect their support for intervention measures (perhaps in part because COVID-related news saturated the news and opinions were set outside of information provided in the experiments).
 
@@
 
-\bibliography{references}
+\bibliography{refs}
 

### 2022-01-17 JDSSV_editorialprocess_knitr.Rnw
update on paper and references

--- 2022-01-04 JDSSV_editorialprocess.Rmd
@@
----
-documentclass: jdssv
-author:
-  - name: Patrick J.F. Groenen
-    affiliation: Erasmus University Rotterdam
-    # use this syntax to add text on several lines
-    address: |
-      | Econometric Institute
-      | Erasmus School of Economics
-      | Erasmus University Rotterdam
-      | P.O. Box  1738 
-      | 3000 DR Rotterdam, The Netherlands
-    email: \email{groenen@ese.eur.nl}
-    url: https://personal.eur.nl/groenen/
-  - name: Stefan Van Aelst
-    affiliation: KU Leuven
-    # To add another line, use \AND at the end of the previous one as above
-    address: |
-      |  Department of Mathematics
-      |  KU Leuven
-      |  Celestijnenlaan 200B
-      |  3001 Leuven, Belgium
-    email: \email{Stefan.VanAelst@kuleuven.be}
-    url: https://wis.kuleuven.be/statdatascience/robust
-    # use a different affiliation in adress field (differently formated here)
-    # affiliation2: Universitat Autònoma de Barcelona
-title:
-  formatted: "Guidelines for the Journal of Data Science, Statistics, and Visualisation"
-  # If you use tex in the formatted title, also supply version without
-  plain:     "Guidelines for the Journal of Data Science, Statistics, and Visualisation Version 0.3"
-  # For running headers, if needed
-  short:     "JDSSV Guidelines, Version 0.3"
-abstract: >
-    This short article illustrates how to write a manuscript for the   \emph{Journal of Data Science, Statistics and Visualisation} (JDSSV) using its \LaTeX\ style files.
-Please follow JDSSV's style guidelines precisely.
-Also, it is recommended to keep the \LaTeX\ code as simple as possible, that is, avoid inclusion of packages/commands that are not necessary.
-keywords:
-  # at least one keyword must be supplied
-  formatted: [JDSSV, style guidelines, comma-separated, not capitalized, "\\proglang{R}"]
-  plain:     [JDSSV, style guidelines, comma-separated, not capitalized, R]
-preamble: >
-  \usepackage{amsmath}
-  \usepackage{longtable}
-  \usepackage{booktabs}
+\documentclass[article]{jdssv}
+
+%% -- LaTeX packages and custom commands ---------------------------------------
+
+%% recommended packages
+\usepackage{thumbpdf,lmodern}
+\usepackage{bm}
+\usepackage{todonotes} %% Remove this package when finishing the manuscript
+
+%% another package (only for this demo article)
+\usepackage{framed}
+
+%% new custom commands
 \newcommand{\class}[1]{`\code{#1}'}
@@
 \newcommand{\ma}[1]{\ensuremath{\mathbf{#1}}}
-output: rticles::jss_article
----
 
-```{r, setup, include=FALSE}
-options(prompt = 'R> ', continue = '+ ')
-```
+%% -- Article metainformation (author, title, ...) -----------------------------
 
-# Mission
+%% - \author{} with primary affiliation
+%% - \Plainauthor{} without affiliations
+%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
+%% - \AND starts a new line, \And does not.
+\author{Patrick J.F. Groenen\\Erasmus University Rotterdam
+   \And Stefan Van Aelst\\KU Leuven}
+\Plainauthor{Patrick J.F. Groenen, Stefan Van Aelst}
+
+%% - \title{} in title case
+%% - \Plaintitle{} without LaTeX markup (if any)
+%% - \Shorttitle{} with LaTeX markup (if any), used as running title
+\title{Guidelines for the Journal of Data Science, Statistics, and Visualisation}
+\Plaintitle{Guidelines for the Journal of Data Science, Statistics, and Visualisation Version 0.1}
+\Shorttitle{JDSSV Guidelines, Version 0.1}
+
+%% - \Abstract{} almost as usual
+\Abstract{
+  This short article illustrates how to write a manuscript for the   \emph{Journal of Data Science, Statistics and Visualisation} (JDSSV) using its {\LaTeX} style files.
+Please follow JDSSV's style guidelines precisely.
+Also, it is recommended to keep the {\LaTeX} code as simple as possible, that is, avoid inclusion of packages/commands that are not necessary.
+}
+
+%% - \Keywords{} with LaTeX markup, at least one required
+%% - \Plainkeywords{} without LaTeX markup (if necessary)
+%% - Should be comma-separated and in sentence case.
+\Keywords{JDSSV, style guidelines, comma-separated, not capitalized, \proglang{R}}
+\Plainkeywords{JDSSV, style guidelines, comma-separated, not capitalized, R}
+
+%% - \Address{} of at least one author
+%% - May contain multiple affiliations for each author
+%%   (in extra lines, separated by \emph{and}\\).
+%% - May contain multiple authors for the same affiliation
+%%   (in the same first line, separated by comma).
+\Address{
+  Patrick J.F. Groenen\\
+  Econometric Institute\\
+  Erasmus School of Economics\\
+  Erasmus University Rotterdam\\
+  P.O. Box  1738 \\
+  3000 DR Rotterdam, The Netherlands\\
+  E-mail: \email{groenen@ese.eur.nl}\\
+  URL: \url{https://personal.eur.nl/groenen/}\\
+\ \\
+%}
+%
+%\Address{
+  Stefan Van Aelst\\
+  Department of Mathematics\\
+  KU Leuven\\
+  Celestijnenlaan 200B\\
+  3001 Leuven, Belgium\\
+  E-mail: \email{Stefan.VanAelst@kuleuven.be}\\
+  URL: \url{https://wis.kuleuven.be/statdatascience/robust}
+}
+
+\begin{document}
+\SweaveOpts{concordance=TRUE}
+<<include=FALSE, cache=FALSE, echo=FALSE, purl=FALSE>>=
+library(knitr)
+opts_chunk$set(comment = NA, size = 'footnotesize', prompt = TRUE, highlight = FALSE, 
+               cache = TRUE, crop = TRUE, concordance = FALSE, fig.align='center',
+               fig.path='Figures_R/fig-',out.width="0.4\\textwidth")
+opts_knit$set(out.format = "latex")
+knit_hooks$set(crop = hook_pdfcrop)
+options(width = 80, prompt = "R> ", continue = "    ", digits = 3)
+thm <- knit_theme$get('default')
+Sys.setenv(TEXINPUTS=getwd(),
+           BIBINPUTS=getwd(),
+           BSTINPUTS=getwd())
+knit_theme$set(thm)
+@
+
+%% -- Introduction -------------------------------------------------------------
+
+%% - In principle "as usual".
+%% - But should typically have some discussion of both _software_ and _methods_.
+%% - Use \proglang{}, \pkg{}, and \code{} markup throughout the manuscript.
+%% - If such markup is in (sub)section titles, a plain text version has to be
+%%   added as well.
+%% - All software mentioned should be properly \cite-d.
+%% - All abbreviations should be introduced.
+%% - Unless the expansions of abbreviations are proper names (like "Journal
+%%   of Statistical Software" above) they should be in sentence case (like
+%%   "generalized linear models" below).
+
+\section{Mission}
 
@@
 
-# Preparing your Manuscript for Submission
+\section[]{Preparing your Manuscript for Submission}
 
-All submissions to JDSSV should be written in \LaTeX\ using the JDSSV style files provided on the journal website \url{https://jdssv.org}.
+All submissions to JDSSV should be written in {\LaTeX} using the JDSSV style files provided on the journal website \url{https://jdssv.org}.
 For initial submission it suffices providing the pdf version of the manuscript together with all the necessary supplementary material including software code to reproduce all results in the manuscript.
 The final version of accepted manuscripts should adhere to all the style guidelines and should incorporate all changes requested by the production editor.
-All \LaTeX\  source files of the final manuscript should be submitted and accepted manuscripts are only published if these files comply with all guidelines and instructions provided by the journal.
+All {\LaTeX}  source files of the final manuscript should be submitted and accepted manuscripts are only published if these files comply with all guidelines and instructions provided by the journal.
 
-# Software
+\section[]{Software}
 
@@
 To make code widely accessible, we advise making it available in a repository such as \url{https://zenodo.org} where it will receive a permanent Digital Object Identifier (DOI) which can be included in the manuscript.
-
-[comment]: # (\todo{\tiny Alternatives: (1) We store the data/code on our JDSSV website with the article. (2) We upload it on Zenodo ourselves (not sure if that can be done).)
+%\todo{\tiny Alternatives: (1) We store the data/code on our JDSSV website with the article. (2) We upload it on Zenodo ourselves (not sure if that can be done).}
 
@@
 \verb|\proglang{}| (programming languages and large programmable systems), \verb|\pkg{}| (software packages), \verb|\code{}| (functions, commands, arguments, etc.).
-If there is such markup in (sub)section titles (as above), a plain text version has to be provided in the \LaTeX\ command as well.
+If there is such markup in (sub)section titles (as above), a plain text version has to be provided in the {\LaTeX} command as well.
 Below we also illustrate how abbrevations should be introduced and citation commands can be employed.
-See the \LaTeX\ code for more details.
+See the {\LaTeX} code for more details.
 
-# Review Process
+\section{Review Process}
 
@@
 
-# Correspondence Analysis {#sec:models}
+%% -- Manuscript ---------------------------------------------------------------
+
+%% - In principle "as usual" again.
+%% - When using equations (e.g., {equation}, {eqnarray}, {align}, etc.
+%%   avoid empty lines before and after the equation (which would signal a new
+%%   paragraph.
+%% - When describing longer chunks of code that are _not_ meant for execution
+%%   (e.g., a function synopsis or list of arguments), the environment {Code}
+%%   is recommended. Alternatively, a plain {verbatim} can also be used.
+%%   (For executed code see the next section.)
+
+\section{Correspondence Analysis} \label{sec:models}
 
@@
 For the analysis of bivariate categorical data or other tables that contain counts, correspondence analysis can be a useful technique to visualize important relations between the categories.
-Consider word count data on explanations of data science by several websources provided by \citet{Lubbe2018}, see Table \ref{tab:word_counts}.
+Consider word count data on explanations of data science by several websources provided by \citet{Lubbe2018}, see Table~\ref{tab:word_counts}.
 
@@
 
-Let $\ma{F}$ be the matrix containing the values in Table \ref{tab:word_counts}. The most simple model to fit such a table is the independence model in matrix $\ma{E}$ defined as 
+Let $\ma{F}$ be the matrix containing the values in Table~\ref{tab:word_counts}. The most simple model to fit such a table is the independence model in matrix $\ma{E}$ defined as 
 \begin{eqnarray*}
@@
 A simple implementation of correspondence analysis in \proglang{R} is given by the function \code{corana()} given by
-
-```{r corana, echo=TRUE}
+<<corana, echo=TRUE>>=
 corana <- function(dat, alpha = 0.5){
@@
 }
-```
+@
 A plot is given by the \code{plot.corana()} method
-
-```{r plot_corana, dependson = "corana", echo = TRUE}
+<<plot_corana, dependson = "corana", echo = TRUE>>=
 plot.corana <- function(out, dims = 1:2, ...){
@@
 }
-```
-
-Assuming that \code{ds.word.cnt} contains the full matrix of which a part is shown in Table \ref*{tab:word_counts}, then the following code does a correspondence analysis on these data:
-```{r ca_biplot_dims23, dependson = "plot_corana", echo = TRUE, fig.width = 6, fig.height = 6, out.width = "10cm", fig.align = "center", fig.cap = "Biplot of a correspondence analysis on word counts used in descriptions of data science by different web sources in dimensions 2 and 3."}
+@
+Assuming that \code{ds.word.cnt} contains the full matrix of which a part is shown in Table~\ref*{tab:word_counts}, then the following code does a correspondence analysis on these data:
+<<ca_biplot_dims23, dependson = "plot_corana", echo = TRUE, fig.width = 6, fig.height = 6, out.width = "10cm", fig.align = "center", fig.cap = "Biplot of a correspondence analysis on word counts used in descriptions of data science by different web sources in dimensions 2 and 3.">>=
 load("ds.word.cnt.RData")
@@
 plot(out.corana, dims = 2:3)
-```
-that yields the plot of Dimensions 2 and 3 in Figure \ref{fig:ca_biplot_dims23}.
+@
+that yields the plot of Dimensions 2 and 3 in Figure~\ref{fig:ca_biplot_dims23}.
 In this plot, the points and labels are made more transparent and the labels decrease in size as they are represented worse in these two dimensions.
 
+%\begin{figure}
+%\begin{center}
+%\includegraphics[width=10cm]{ca_example.pdf}
+%\end{center}
+%\caption{Biplot of a correspondence analysis on word counts used in descriptions of data science by different web sources in dimensions 2 and 3.} \label{fig:ca_biplot_dims23}
+%\end{figure}
+
+%% -- Summary/conclusions/discussion -------------------------------------------
+
+%\section{Summary and discussion} \label{sec:summary}
+%
+%\begin{leftbar}
+%As usual \dots
+%\end{leftbar}
+
+%% -- Optional special unnumbered sections -------------------------------------
 \newpage
-# Computational Details {-}
+\section*{Computational Details}
 
@@
 
-# Acknowledgments {-}
+\section*{Acknowledgments}
 
@@
 
+%% -- Bibliography -------------------------------------------------------------
+%% - References need to be provided in a .bib BibTeX database.
+%% - All references should be made with \cite, \citet, \citep, \citealp etc.
+%%   (and never hard-coded). See the FAQ for details.
+%% - JSS-specific markup (\proglang, \pkg, \code) should be used in the .bib.
+%% - Titles in the .bib should be in title case.
+%% - DOIs should be included where available.
+
 \bibliography{refs}
+
+%% -- Appendix (if any) --------------------------------------------------------
+%% - After the bibliography with page break.
+%% - With proper section titles and _not_ just "Appendix".
 
@@
 
-\setcounter{section}{0}
-\renewcommand{\thesection}{\Alph{section}}
+\begin{appendix}
 
-# More Technical Details {#app:technical}
+\section{More Technical Details} \label{app:technical}
 
@@
 
+%For more technical style details, please check out JSS's style FAQ at
+%\url{https://www.jstatsoft.org/pages/view/style#frequently-asked-questions}
+%which includes the following topics:
+%\begin{itemize}
+%  \item All main words in titles and headers start with a capital.
+%  \item Use vectorized graphics formats such as eps and pdf when possible. Graphics should be of high quality while trying to minimize the file size.
+%\end{itemize}
+
 \section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}
 
-References need to be provided in a \textsc{Bib}\TeX\ file (\code{.bib}).
+References need to be provided in a \textsc{Bib}{\TeX} file (\code{.bib}).
 All references should be made with \verb|\cite|, \verb|\citet|, \verb|\citep|, \verb|\citealp| etc.\ (and never hard-coded).
@@
 
-Cleaning up \textsc{Bib}\TeX\ files is a somewhat tedious task -- especially when acquiring the entries automatically from mixed online sources.
+Cleaning up \textsc{Bib}{\TeX} files is a somewhat tedious task -- especially when acquiring the entries automatically from mixed online sources.
 However, it is important that informations are complete and presented in a consistent style to avoid confusions.
 JDSSV requires the following format.
+\begin{itemize}
+  \item Specific markup (\verb|\proglang|, \verb|\pkg|, \verb|\code|) should
+    be used in the references.
+  \item Titles should be inserted in title case.
+  \item Journal titles should not be abbreviated and in title case.
+  \item DOIs should be included where available.
+  \item Software should be properly cited as well. For \proglang{R} packages
+    \code{citation("pkgname")} typically provides a good starting point.
+\end{itemize}
+\end{appendix}
+\newpage
+%% -----------------------------------------------------------------------------
 
-* Specific markup (\verb|\proglang|, \verb|\pkg|, \verb|\code|) should be used in the references.
-* Titles should be inserted in title case.
-* Journal titles should not be abbreviated and in title case.
-* DOIs should be included where available.
-* Software should be properly cited as well. For \proglang{R} packages \code{citation("pkgname")} typically provides a good starting point.
-
-\newpage
+\end{document}

### 2022-01-17 JDSSV_manuscript_knitr.Rnw
update on paper and references

@@
 \Abstract{
-  Covid-19 has created a world-wide interest in understanding the dynamics of a pandemic.
-Hence, numerous media outlets as well as researchers have produced comprehensive data visualizations to illustrate the relevant trends and figures.
-In this paper, we will look at an elective choice of Covid-19 data visualizations to evaluate and discuss currently established visualization tools in their capacity to provide a communication channel both within the data science team and also between data analyst, domain experts and a more general interested audience.
-While there is no fixed catalogue of evaluation criteria for data visualizations we will try to provide an overview on the different core aspects of visualization evaluation and their competing principles.
+  Covid-19 has sparked a worldwide interest in understanding the dynamic evolution of a pandemic and tracking the effectiveness of preventive measures and rules.
+For this reason, numerous media and research groups have produced comprehensive data visualisations to illustrate the relevant trends and figures.
+In this paper, we will look at a selection of Covid 19 data visualisations to evaluate and discuss the currently established visualisation tools in terms of their ability to provide a communication channel both within the data science team and between data analysts, domain experts and a general interested audience.
+Although there is no set catalogue of evaluation criteria for data visualisations, we will try to give an overview of the different core aspects of visualisation evaluation and their competing principles.
   }
@@
 
-Here some general introduction
+Over the past two years, several waves of Covid-19 infections with different mutants of the SARS-CoV-2 virus have swept across the globe, claiming many lives, causing numerous health damages and affecting our personal lives in many ways.
+According to the WHO, over 304 million confirmed cases and over 5.4 million deaths have been reported as of early January 2022 \footnote{\url{https://www.who.int/publications/m/item/weekly-epidemiological-update-on-covid-19---11-january-2022}}.
+The pandemic has generated enormous interest in epidemiological data, its analysis and visualisation.
+From the beginning of the pandemic, data on the number of infections and covid-related deaths has been published daily and made available to the public.
+Media, politicians and individuals use this data to build their narratives about the pandemic, discuss its evolution, justify the measures taken and discuss different prevention strategies against the spread of the virus.
+So there are several goals to be achieved by visualising Covid-19 data.
+These goals and their priorities were adapted as dynamically as the virus mutated and the pandemic changed pace.
+But as with any other data visualisation two general principles remain the same: ensuring clear vision by optimising the data-to-ink ratio \citep{Tufte2001} (or signal-to noise ratio) and ensuring clear understanding by organising the graphics in such a way that the story of the data is told most effectively.
 
-Effective participation in the social dynamics of a democracy requires clear understanding of quantitative evidence and rules by the people.
-As the medical community, governments, business, and the public come together to fight the COVID-19 pandemic, scientists have a special role in helping to communicate what the data actually mean.
-For data analysis and modeling to have an impact as a component of decision making, appropriate reporting and communication is key.
-There are numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reporting, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org).
-These standards are based on commonly accepted core quality principles and values such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility.
-For measures to restrain and overcome an epidemic effectively, communication among experts that follows highest professional and ethical standards is not sufficient.
-In a democratic society, policy measures can only be implemented if they are based on the acceptance of the wider population.
-This puts high demands on skills associated with communicating statistical evidence on the side of scientists, governments and media, and a citizenry able to understand statistical messages.
-%
+In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic as they have been showcasing their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods.
+Data journalism will certainly be one of the beneficiaries of the Covid-19 pandemic and it has become an innovative part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards.
 
-In recent decades, there have been numerous publications, initiatives, and ideas to improve the communication of quantitative and statistical information, see \cite{Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}, to name only a few.
-Data journalism has recently taken off as an innovative component of news publishing, and COVID-19 provides numerous excellent examples, often using an interactive visual format on the Internet, such as dashboards.
-%Nevertheless, both the complexity of the phenomena and the "bipolarity" of statistical thinking remain challenges.
-A fundamental problem in assessing probabilities lies, for example, in the intuitive conflation of subjective risks ("how likely am I to become infected") and general risks ("how likely is it that some person will become infected").
-Another issue is that of equating sensitivity of a diagnostic test and the positive predictive value \citep{Eddy1982,Gigerenzer2007,McDowell2019,Binder2020}.
-In particular, the prevalence (or base rate) is often neglected leading to this confusion.
-As \cite{Hoffrage} {\color{blue}Hoffrage et al.
-(2000)} point out “… statistics expressed as natural frequencies improve the statistical thinking of experts and nonexperts alike”.
-Smooth sliding back and forth between probabilities and natural frequencies is easy once figures are related to a unifying reference value.
+At the same time we still see a lot of defective graphics disseminated and shared: some that violate fundamental visualisation and statistical reporting principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility.
+These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reporting, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org).
+Numerous publications, initiatives, and ideas to improve the communication of quantitative and statistical information have been prepared, see for example \cite{Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.
+
+The Covid-19 pandemic has provided ample evidence that policy measures are only effectively implemented and adhered to by the people if they are accepted by a vast majority of the population.
+To achieve this goal, effective communication of quantitative evidence and statistical results among scientists, governments, media, and the citizens is essential to justify the adequacy, usefulness, and relevance of the measures.
+
 A well proven method to effective communication lies in representing information in accessible ways (Gigerenzer et al., 2007; Gigerenzer and Edwards, 2011) by saying, for instance, “one in 10” instead of 10\%.
@@
 
+Nevertheless, both the complexity of the phenomena and the "bipolarity" of statistical thinking remain challenges.
+
 While human thinking tends towards pattern simplification and political communication also prefers a simple cause-effect relationship, real phenomena are often multivariate.
@@
 
-% May also want to mention at the outset the limitations in the data due to low testing capacity, weekly effects due to reporting, and so on.
-We must be able to represent the data that exist visually while acknowledging the limitations associated with that data.
-
 \section[]{The global perspective}
 
-Visual representations take a central position in public communication and aim to represent the corresponding dynamics and contents in a quickly understandable way.
-Usually either time-dependent parameters or data with a spatial reference are visualised.
-For spatially distributed data, choropleth maps are predominantly used, in which administrative regions defined by the responsible health authorities are coloured according to the distribution density of the infection figures or variables derived from them (see Figure~\ref{fig:choro}).
-Their visual perception problems - such as the visual dominance of the area of administrative regulatory frameworks that have no direct relation to infection events - are well known but still widespread.
-In addition, the use of ordinance thresholds as the basis for color scaling is often at odds with color schemes that emphasize real spatial distributional differences.
+On 11 March 2020, WHO declared the outbreak of the novel coronavirus disease (COVID-19) a pandemic, and since that date at the latest, the global perspective of the disease has been in the public eye.
+The spatial spread of the virus and the resulting cases and deaths are commonly visualized by choropleth maps, see for example Figure~\ref{fig:choro1} showing the total number of infections reported in each country as of January 14, 2022.
+For the pandemic perspective a central element of the narrative is the ubiquity of the disease and the accompanying global impact.
+Choropleth maps based on raw numbers of cases might look convincing and fit to the purpose, but neglect a number of well-know caveats for statistical reporting and visualisation:
+\begin{enumerate}
+\item Absolute value unsuitability: As explained in \citep{monmonier2005, slocum2008, speckmann2010} among others, choropleth maps are fundamentally unsuitable for the representation of absolute numbers.
+Especially in the case of similarly coloured areas of the regions, viewers tend to integrate them unconsciously and perceive choropleths as representations of density.
+They also do not help to convey the desired message as the absolute numbers of Covid-19 cases are strongly influenced by the population size of the country, but also by the number of tests performed and the accuracy of the recording and reporting system.
+\item The area-bias: The visual impression is determined more by the colour and the geographical area of the individual countries than by the number of Covid cases.
+Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
+\item Color-scheme obstructions: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997}.
+The choice of a continuous scale or a categorical scale, the choice of a scale that promotes the recognition of patterns or a scale that supports the filtering out of specific map details, influences the quality of choropleth maps.
+\end{enumerate}
 
 \begin{figure*}
-	\includegraphics[width = 0.9\textwidth]{choroplethen}
-	\caption{Choropleth map of the incidence figures for Germany by district. Source: Robert-Koch-Institute https://app.23degrees.io/export/oCRP768wQ3mCswE7-choro-corona-faelle-pro-100-000/image.}
-	\label{fig:choro}
+	\includegraphics[width = 0.98\textwidth]{who_totalcases_choro.png}
+	\caption{Choropleth map of the Covid-19 cases world-wide by country. Source: WHO \url{https://covid19.who.int}}
+	\label{fig:choro1}
 \end{figure*}
-
-For time-dependent parameters, different variants of time series diagrams are used, predominantly line and column diagrams.
-Classic errors of graphical representation, such as overemphasising temporal variability by reducing the value axis to a small section, have now largely disappeared from the media.
-Switching between line and bar charts for purely design-related reasons in order to produce corresponding graphical diversity nevertheless seems questionable.
-The use of logarithmic scales in time series diagrams should be evaluated with caution.
-On the one hand, they tempt superficial readers to underestimate dynamic growth processes; on the other hand, they increase the demands on the mathematical and statistical literacy of the readership without corresponding advantages of visual representation.
-Figure~\ref{fig:incidence} shows the time course of the 7-day incidence per 100,000 people between 24 January and 4 February 2021 for some selected countries.
-While the differences appear relatively small on the logarithmic scale, the linear scale shows considerable differences.
-
-\begin{figure}
-		\includegraphics[width = 0.49 \textwidth]{incidence1.png}
-			\includegraphics[width = 0.49 \textwidth]{incidence2.png}
-	\caption{The 7-day incidence for different countries over the course of time.
-On the logarithmic scale (left graphic), differences seem small.
-The linear scale (right graphic), however, shows considerable differences.
-Source: Our World in Data, https://ourworldindata.org/covid-cases?country=IND~USA~GBR~CAN~DEU~FRA.}
-	\label{fig:incidence}
-\end{figure}
 

### 2022-01-17 JDSSV_manuscript_knitr.Rnw
update Jan 17

@@
 
+%shall we include a list of newspapers/media outlets that we aim at covering?
+
 At the same time we still see a lot of defective graphics disseminated and shared: some that violate fundamental visualisation and statistical reporting principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility.
-These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reporting, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org).
+These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org).
 Numerous publications, initiatives, and ideas to improve the communication of quantitative and statistical information have been prepared, see for example \cite{Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.
 
-The Covid-19 pandemic has provided ample evidence that policy measures are only effectively implemented and adhered to by the people if they are accepted by a vast majority of the population.
-To achieve this goal, effective communication of quantitative evidence and statistical results among scientists, governments, media, and the citizens is essential to justify the adequacy, usefulness, and relevance of the measures.
+The Covid 19 pandemic has amply demonstrated that policies are only effectively implemented and followed by the population if they are accepted by a large majority of the population.
+To achieve this goal, effective communication of quantitative evidence and statistical results between scientists, governments, the media and citizens is essential to justify the appropriateness, usefulness and relevance of the measures.
 
-A well proven method to effective communication lies in representing information in accessible ways (Gigerenzer et al., 2007; Gigerenzer and Edwards, 2011) by saying, for instance, “one in 10” instead of 10\%.
-Using absolute in place of relative numbers, representing information in appealing graphic forms, and summarizing most relevant facts in “fact boxes” are some of the methods developed and advocated for increasing transparency in communication between stakeholders such as health providers and patients (See, for example: https:// www.
-hardingcenter.de)
-Fact boxes combined with icon arrays are recommended for the presentation of test results.
-Both representations are based on natural frequencies \citep{[62],Krauss2020} and present case numbers as simply and concretely as possible.
-Many scientific studies show that icon arrays help people understand numbers and risks more easily (e.g.
+A best practice for effective communication is to present information in an understandable way (Gigerenzer et al., 2007; Gigerenzer and Edwards, 2011), for example by saying "one in ten" instead of 10\%.
+Using absolute rather than relative numbers, presenting information in an appealing graphic form and summarising the most important facts in "fact boxes" are some of the methods that have been developed and advocated to increase transparency in communication between stakeholders such as healthcare providers and patients (see for example: https:// www.hardingcenter.de).
+Fact boxes in combination with icon arrays are recommended for the presentation of test results.
+Both representations are based on natural frequencies \citep{[62],Krauss2020} and represent case numbers as simply and concretely as possible.
+Many scientific studies show that icon arrays help people to understand numbers and risks more easily (e.g.
 \cite{McDowell2019}).
@@
 
-Nevertheless, both the complexity of the phenomena and the "bipolarity" of statistical thinking remain challenges.
+Nevertheless, both the complexity of phenomena and the "bipolarity" of statistical thinking remain a challenge.
 
-While human thinking tends towards pattern simplification and political communication also prefers a simple cause-effect relationship, real phenomena are often multivariate.
-Thus, when studying COVID-19 and predicting its spread, it is not only important to consider its symptomatology, the incidence and geographic distribution of diseases, population behavior patterns, government policies and impacts on the economy, on schools, on people in nursing homes and on social life as a whole, but to integrate these into the data analyses and the communication of results.
-Associations observed in the data can often be caused by third-party variables (confounders).
-In addition, much of the data comes from observational studies, which usually makes a robust causal attribution problematic.
-Statisticians calling out these limitations, however, face the danger that their statements might be pulled out one side in a polarized debate \citep{McConway2021}.
+While human thinking tends to simplify patterns and political communication also prefers a simple cause-effect relationship, real phenomena are often multivariate.
+Thus, in studying COVID-19 and predicting its spread, it is not only important to consider symptomatology, disease incidence and geographic distribution, population behaviour patterns, government policies, and impacts on the economy, schools, people in nursing homes, and society at large, but also to incorporate these into data analyses and communication of results.
+Associations observed in the data can often be caused by third party variables (confounders).
+In addition, much of the data comes from observational studies, which usually makes robust causal attribution problematic.
+However, statisticians who point out these limitations are at risk of having their statements pulled out one-sidedly in a polarised debate \citep{McConway2021}.
 

### 2022-01-17 JDSSV_manuscript_knitr.Rnw
Summarize log scale vs. linear scale issues

--- 2022-01-15 JDSSV_manuscript_knitr.Rnw
@@
 The linear scale makes it difficult to compare the trajectory of different waves to determine how severe the current status is relative to the past, because the primary contrast is the height of the relative peaks, rather than the growth \\emph{rate}.
-A similar graph on the log scale would have the peaks at much more similar heights (though there would still be a difference), allowing the reader to focus on other information.">>=
+A similar graph on the log scale would have the peaks at much more similar heights (though there would still be a difference), allowing the reader to focus on other information, such as the slope of the relative lines.">>=
 cvstate %>%
@@
 
-<<log-scales-ref-lines, fig.width = 8, fig.height = 4, out.width = ".8\\linewidth", fig.cap = "Reported COVID cases in New York State, 2020-2022.
-The linear scale makes it difficult to compare the trajectory of different waves to determine how severe the current status is relative to the past.">>=
-cvstate %>%
-  filter(state == "New York") %>%
-  ggplot(aes(x = date, y = daily_avg_cases)) + 
-  geom_line() + 
-  ggtitle("Reported COVID cases in New York State, 2020-2022") +
-  ylab("Daily Cases reported (7-day average)") + 
-  xlab("Date") + theme_bw() + 
-  scale_y_log10()
-@
-
 % Discussion of sites which allow the user to switch back and forth between scales
@@
 It is not clear that the use of log or linear scales during the COVID-19 pandemic had a large effect on public opinion.
-Several studies were conducted in the early stages of the pandemic \citep{romanoScaleCOVID19Graphs2020, seviLogarithmicLinearVisualizations2020, ryanGraphsLogarithmicAxes2020} and results seem to suggest that while individuals have difficulty understanding log scale graphs, these issues do not tend to affect their support for intervention measures (perhaps in part because COVID-related news saturated the news and opinions were set outside of information provided in the experiments).
+Several studies were conducted in the early stages of the pandemic \citep{romanoScaleCOVID19Graphs2020, seviLogarithmicLinearVisualizations2020, ryanGraphsLogarithmicAxes2020} and results seem to suggest that while individuals have difficulty understanding log scale graphs, these issues do not tend to affect their support for intervention measures, perhaps in part because COVID-related news saturated the news and opinions were set outside of information provided in the experiments.
+This saturation makes it difficult to study the graphical influences while removing effects of popular opinion, political leaning, and emotional sentiment relating to the pandemic.
+
+If we evaluate the use of log and linear scales under the more general context of exponential (or near-exponential) growth rates, however, we can gain some clarity as to their use in this particular context.
+We are abysmal at forecasting exponential growth, vastly under-estimating future growth by using linear or quadratic approximations \citep{wagenaarExtrapolationExponentialTime1978a,lawrenceExploringJudgementalForecasting1992,timmersInverseStatisticsMisperception1977a}.
+If the goal of a chart of COVID case counts is to allow individuals to forecast the trend and make decisions accordingly, then using a log scale (with all of its pitfalls in understanding) may be the best option, as it at least replaces the need to forecast along an exponential curve with the need to forecast along a straight line.
+However, it is not clear that individuals can transfer that prediction back to a linear scale.
+If the goal of presenting a chart of case counts is to tell the story of what has happened (rather than supporting future decision-making), then it is undoubtedly better to present the data on a more familiar linear scale that requires less cognitive load on the part of the viewer.
+
+Unfortunately, there are not a lot of good options for supporting forecasting decisions in mathematically novice viewers.
+Guide lines, like those used in the Financial Times, were helpful in the initial phases of the pandemic, but no outlets that we are aware of shifted the lines to show the growth rate of the current peak (as opposed to the initial peak), and due to the issues shown in \Cref{fig:log-scale-failures}, these guide lines may not have been all that successful in any case.
+Those hoping to create successful time-series charts that allow forecasting are thus left with two options: deal with our inability to forecast exponential growth, or deal with most individuals' inability to translate log scales into practical reality.
 

### 2022-01-17 JDSSV_manuscript_knitr.Rnw
color -> colour to standardize, also visualisation

@@
 my_theme <- theme_bw() + 
-  theme(plot.background = element_rect(fill = "transparent", color = NA), 
-        panel.background = element_rect(fill = "transparent", color = NA), 
-        legend.background = element_rect(fill = "transparent", color = NA),
+  theme(plot.background = element_rect(fill = "transparent", colour = NA), 
+        panel.background = element_rect(fill = "transparent", colour = NA), 
+        legend.background = element_rect(fill = "transparent", colour = NA),
         axis.text.y = element_text(angle = 90, hjust = 0.5)) 
@@
 
-<<colorpal, include = F>>=
+<<colourpal, include = F>>=
 # Including this for temporary compatibility - we may want to use *shudder* viridis?
-main_colors <- c("#d00000", "#f5f1e7", "#c7c8ca")
+main_colours <- c("#d00000", "#f5f1e7", "#c7c8ca")
 
 # use as accents
-secondary_colors <- c("#001226", "#249ab5")
+secondary_colours <- c("#001226", "#249ab5")
 
 # These are only to be used for infographics/charts
-tertiary_colors <- c("#bccb2a", "#f58a1f", "#005d84", "#ffd74f", "#a5228d")
+tertiary_colours <- c("#bccb2a", "#f58a1f", "#005d84", "#ffd74f", "#a5228d")
 
-color_palette <- c("#d00000", "#249ab5",  "#1B8500", "#f58a1f", "#a5228d", "#001226", "#B1EB66", "#005d84", "#c25700")
+colour_palette <- c("#d00000", "#249ab5",  "#1B8500", "#f58a1f", "#a5228d", "#001226", "#B1EB66", "#005d84", "#c25700")
 
-heat_color_palette <- c("#ffd74f", "#f58a1f", "#c25700", "#d00000", "#9D0000", "#6A0000")
+heat_colour_palette <- c("#ffd74f", "#f58a1f", "#c25700", "#d00000", "#9D0000", "#6A0000")
 
@@
   ggplot(aes(x = date, y = daily_avg_cases, group = state)) +
-  geom_textline(aes(color = state, label = state, vjust = -0.1, hjust =.93) , text_smoothing = 50) + 
+  geom_textline(aes(colour = state, label = state, vjust = -0.1, hjust =.93) , text_smoothing = 50) + 
   ylab("Daily average cases (7 day average)") + 
   xlab("Date") + 
-  guides(color = "none") + 
+  guides(colour = "none") + 
   scale_y_log10() + 
@@
   ggplot(aes(x = date, y = daily_avg_cases, group = state)) +
-  geom_textline(aes(color = state, label = state, vjust = -0.25, hjust = .85) , text_smoothing = 60) + 
+  geom_textline(aes(colour = state, label = state, vjust = -0.25, hjust = .85) , text_smoothing = 60) + 
   ylab("Daily average cases (7 day average)") + 
   xlab("Date") + 
-  guides(color = "none") + 
+  guides(colour = "none") + 
   facet_grid(~"Linear") + 
@@
 tmp %>%
-ggplot(aes(x = x, y = value, color = name, size = name)) + 
+ggplot(aes(x = x, y = value, colour = name, size = name)) + 
   geom_line(alpha = .9) +
@@
   scale_size_manual(values = c(1.75, 1, 1)) +
-  scale_color_manual(values = c("#249ab5", "#d00000", "#001226")) +
+  scale_colour_manual(values = c("#249ab5", "#d00000", "#001226")) +
   scale_fill_discrete(name = "") + 
@@
 
-% Focus primarily on time-series graphs (as opposed to log scales in color)
+% Focus primarily on time-series graphs (as opposed to log scales in colour)
+
+\begin{figure}
+\centering
+\includegraphics[width=.8\linewidth]{ft-reference-lines}
+\caption{An annotated time series from the Financial Times which appeared on Twitter on November 12, 2020.
+This chart has annotations which show the decreasing $R_{eff}$ in successive peaks of the pandemic, with periods of $R_{eff}<1$ between surges in cases.
+These annotations assist the reader with drawing complex conclusions from the chart, but are difficult to automate and thus tend to be manually curated.
+Source: Mark Gubrud (@mgubrud), \url{https://twitter.com/mgubrud/status/1326784082399944704}.}
+\label{fig:log-scale-ref-multiple}
+\end{figure}
 
@@
 such as version numbers, operating systems, or compilers could be included
-in an unnumbered section. Also, auxiliary packages (say, for visualizations,
+in an unnumbered section. Also, auxiliary packages (say, for visualisations,
 maps, tables, \dots) that are not cited in the main text can be credited here.

### 2022-01-17 JDSSV_manuscript_knitr.Rnw
Additional figures

@@
 Especially in the case of similarly coloured areas of the regions, viewers tend to integrate them unconsciously and perceive choropleths as representations of density.
-They also do not help to convey the desired message as the absolute numbers of Covid-19 cases are strongly influenced by the population size of the country, but also by the number of tests performed and the accuracy of the recording and reporting system.
+They also do not help to convey the desired message as the absolute numbers of Covid-19 cases are strongly influenced by the population size of the country, number of tests performed, and the accuracy of the recording and reporting system.
+\Cref{fig:scaled-choro} shows case load scaled by population, but is still subject to data quality and comparability issues related to testing and case reporting; these biases are hidden as the chart presents all data as equivalent and comparable when in fact it is likely that direct comparisons may be unwise in many cases.
 \item The area-bias: The visual impression is determined more by the colour and the geographical area of the individual countries than by the number of Covid cases.
 Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
-\item Color-scheme obstructions: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997}.
+Cartograms, like that shown in \Cref{fig:hex-cartogram}, can mitigate the area bias by sacrificing the viewer's ease of geographic recognition.
+\item colour-scheme obstructions: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997}.
 The choice of a continuous scale or a categorical scale, the choice of a scale that promotes the recognition of patterns or a scale that supports the filtering out of specific map details, influences the quality of choropleth maps.
@@
 \end{figure*}
+
+\begin{figure}
+\includegraphics[width=.99\linewidth]{covid-tracking-hex-cartogram}
+\caption{Hexagonal Cartogram of Covid-19 cases in the United States by state, per million people.
+This somewhat mitigates the area bias issue, but is still subject to other perceptual biases.
+Source: The COVID Tracking Project at The Atlantic \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=.99\linewidth]{nyt-covid-scaled-choro}
+\caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
+\end{figure}
+
+Throughout the pandemic, both in choropleth maps and time-series graphs, creators have had to grapple with scale.
+When scales automatically adjust, mental comparisons across time can become misleading, because e.g.
+the colour scale from yesterday does not mean the same thing as it did today, as the range has expanded and the colours have stayed the same.
+This problem is more noticeable in choropleths than in time-series charts, where we are more likely to expect and notice a scale change, but it is still a fundamental issue in many different types of charts.
+Some publications leaned into this by arranging their graphics to highlight the fact that cases were ``off the charts", as in \Cref{fig:jbm-tweet}, implicitly cuing readers that the situation was exceptional and worthy of their attention.
+
+\begin{figure}
+\includegraphics[width=.99\linewidth]{ft-off-charts}
+\caption{Chart showing the increase in cases as a result of the B.1.617.2 variant in regions of the UK.
+Notably, this chart arranges areas to accommodate certain regions whose cases are ``off the chart" in order to emphasize the exceptional situation.
+Source: Tweet by John Burn-Murdoch (@jburnmurdoch), designer for the Financial Times.
+\url{https://twitter.com/jburnmurdoch/status/1397995388267810818}.}
+\label{fig:jbm-tweet}
+\end{figure}
+
+% Is it worth discussing e.g.
+https://www.nytimes.com/interactive/2021/us/covid-risk-map.html that does not make this distinction and runs into the "boy who cried wolf" problem?
+When everything is always extremely high risk we just tend to tune things out.
+I don't want to distract from the overview, though...
 
@@
 \section{Time series}
-% Line plots, frequency domain, NY times tile plots w/ cases represented in color intensity
+% Line plots, frequency domain, NY times tile plots w/ cases represented in colour intensity
+
+Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time.
+This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state.
+Time series plots are the most natural choice for these goals, and over the course of the pandemic, many different attempts have been made to show time series information using different graphical forms.
+
+\begin{figure}
+\centering
+\includegraphics[width=.99\linewidth]{wapo-covid-monitoring}
+\caption{COVID monitoring summary charts on the Washington Post's Coronavirus page. Source: \url{https://www.washingtonpost.com/coronavirus/}, 2022-01-17.}
+\label{fig:wapo-covid-time-series}
+\end{figure}
+
+The most obvious time-series plot is also the most common: plotting cases against date, often with a 7 or 14-day smooth to handle periodicity in case reporting due to the work week, as shown in \Cref{fig:wapo-covid-time-series}.
+This allows for individuals to assess the current situation and easily compare the current number of cases to past times for which the individual has a direct reference for what to expect in impact to daily life.
+Often, as in \Cref{fig:wapo-covid-time-series}, these time series show multiple variables: cases, deaths, and hospitalizations (and sometimes vaccinations) to provide a more comprehensive view of the situation; this is particularly useful as there has been some decoupling of case numbers and hospitalizations/deaths with the emergence of effective vaccinations and strains which are more contagious but potentially also less severe.
+Many outlets also provide additional charts designed to provide some immediate context to readers, as shown in \Cref{fig:wapo-context}.
+
+\begin{figure}
+\centering
+\includegraphics[width=.5\linewidth]{wapo-covid-context}
+\caption{Some outlets provide contextual information to assist readers with interpreting time-series plots.
+Source: Washington Post Coronavirus Tracking, \url{https://www.washingtonpost.com/graphics/2020/national/coronavirus-us-cases-deaths/}.}
+\label{fig:wapo-context}
+\end{figure}
 
@@
 ggplot(data = tmp2, aes(x = date, ymin = 0, ymax = y, 
-                        color = whole_num_days, 
+                        colour = whole_num_days, 
                         fill = whole_num_days, 
@@
   scale_y_continuous(breaks = c(n/2, n), labels = paste0(c(n/2000, n/1000), "k"))  + 
-  scale_color_gradientn(guide = 'none', paste0("Days to ", n/1000, "k\nmore cases"), space = "Lab", 
+  scale_colour_gradientn(guide = 'none', paste0("Days to ", n/1000, "k\nmore cases"), space = "Lab", 
     trans = "log10",
@@
   theme(axis.title = element_blank(), 
-        legend.background = element_rect(fill = "transparent", color = NA), 
+        legend.background = element_rect(fill = "transparent", colour = NA), 
         legend.position = c(0, .66), legend.justification = c(0, 1), legend.direction = "horizontal")
@@
 Even researchers do not always read and interpret log scales correctly\citep{mengeLogarithmicScalesEcological2018}; expecting the general public to do so is difficult under normal circumstances\citep{hecklerStudentAccuracyReading2013} is difficult.
-When panic, fear, uncertainty, and doubt about the situation are added to the mix,
+When panic, fear, uncertainty, and doubt about the situation are added to the mix, it is easy to imagine that we become even worse when interpreting graphics.
 % Discussion of how this section's conclusions change with the pandemic
@@
 Diagonal reference lines from the origin were also less helpful, as the growth of cases or deaths was no longer approximately exponential and varied over time; for these reference lines to be effective there would need to be a clear idea of when the case counts started to increase exponentially, which is difficult to determine whilst in the thick of a potential COVID wave.
+Occasionally graphs with manually drawn reference lines were made available, but usually only in a retrospective manner, as in \Cref{fig:log-scale-ref-multiple}.
 

### 2022-01-17 JDSSV_manuscript_knitr.Rnw
phrasing
Frequency domain and nyt sparklines charts

@@
 \usepackage{cleveref}
+
+\usepackage{subcaption}
 
@@
 \begin{figure*}
-	\includegraphics[width = 0.98\textwidth]{who_totalcases_choro.png}
+	\includegraphics[width = 0.98\textwidth]{who_totalcases_choro}
 	\caption{Choropleth map of the Covid-19 cases world-wide by country. Source: WHO \url{https://covid19.who.int}}
@@
 \begin{figure}
-\includegraphics[width=.99\linewidth]{covid-tracking-hex-cartogram}
+\centering
+\begin{subfigure}[t]{.45\textwidth}
+\includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
 \caption{Hexagonal Cartogram of Covid-19 cases in the United States by state, per million people.
@@
 Source: The COVID Tracking Project at The Atlantic \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=.99\linewidth]{nyt-covid-scaled-choro}
+\end{subfigure}\hfill
+\begin{subfigure}[t]{.45\textwidth}
+\includegraphics[width=\textwidth]{nyt-covid-scaled-choro}
 \caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
+\end{subfigure}
 \end{figure}
@@
 I don't want to distract from the overview, though...
+https://twitter.com/rypan/status/1282020934044401664 is a good example
 
@@
 
-Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time.
+Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time, as there are new developments on a daily basis.
+Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis.
 This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state.
@@
 
-<<freq-domain-plot, include = F, eval = F>>=
-show_states <- c("California", "Texas", "New York", "Arizona", "Iowa", "Washington")
+However, there are some limitations with this presentation: if we are interested in showing the change in counts over time, viewers must assess the slope of the line, rather than its position.
+We know that slope judgments are much less accurate than position \citep{clevelandGraphicalPerceptionVisual1987}, even in relatively simple situations; when we add in our ability to assess exponential growth, our perceptual accuracy is even more suspect.
 
-n <- 50000
-tmp <- cvstate %>%
-  filter(state %in% show_states) %>%
-  select(state, date, cases) %>%
-  mutate(mod_num = cases %% n,
-         whole_num = floor(cases/n)) %>%
-  group_by(state, whole_num) %>%
-  mutate(whole_num_days = n()) %>%
-  group_by(state) %>%
-  arrange(date)
+Other time-series representations were more abstract, sacrificing exact data representations for a higher-level summary overview that allowed viewers to make comparisons between regions and points in time without the burden of numerical calculations.
+Several versions from the New York Times are shown in \Cref{fig:sparklines-heatmap-nyt}.
+The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data, however, they do not present precise numerical information and even the color scale is fairly opaque -- it would be extremely difficult to translate the visualization into any sort of accurate numerical estimate.
+In addition, by hiding these details from the user, it is very difficult to identify when the representation changes over time, as it did several times during the summer of 2020.
+The use of a similar color scheme as well as similar representations made it extremely difficult for users to identify that the value represented had changed.
+Clearly, the purpose of these charts is not to provide numerical precision, but rather a comparative assessment of the status of one region relative to others as well as relative to previous time points.
 
-tmp2 <- bind_rows(
-  # Day before first case detected:
-  filter(tmp, whole_num == 0) %>%
-    group_by(state) %>%
-    summarize(date = min(date) - lubridate::days(1), y = 0, whole_num = 0, 
-              whole_num_days = unique(whole_num_days), mod_num = 0) %>%
-    unique() %>%
-    ungroup(),
-  # Date before the count rolls over
-  tmp %>%
-    group_by(state, whole_num) %>%
-    filter(date == max(date)) %>%
-    group_by(state) %>%
-    mutate(y = ifelse(whole_num == max(whole_num), mod_num, n),
-           whole_num_days = ifelse(whole_num == max(whole_num), whole_num_days/(y/10000), whole_num_days)),
-  # Date after the count rolls over
-  tmp %>%
-    group_by(state, whole_num) %>%
-    filter(date == min(date)) %>%
-    mutate(y = 0, date = date - lubridate::days(1))
-  ) %>%
-  group_by(state, whole_num) %>%
-  mutate(whole_num_days = max(whole_num_days))
+\begin{figure}
+\centering
+\includegraphics[width=.3\linewidth]{NYT-heatmap-crop}
+\includegraphics[width=.3\linewidth]{NYT-heatmap-crop3}
+\includegraphics[width=.3\linewidth]{NYT-heatmap-crop2}
+\caption{Sparklines style heatmaps embedded in a table of coronavirus data by state.
+Over the course of Summer 2020, the specific measures shown changed, even though the form of the data remained highly similar.
+Source: \url{https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html} (May 6, 2020; June 15, 2020; and July 24, 2020, respectively) }
 
-ggplot(data = tmp2, aes(x = date, ymin = 0, ymax = y, 
-                        colour = whole_num_days, 
-                        fill = whole_num_days, 
-                        group = whole_num)) + 
-  geom_ribbon() + 
-  facet_grid(state~.) + 
-  scale_x_date(date_breaks = "month", date_labels = "%B") +
-  scale_y_continuous(breaks = c(n/2, n), labels = paste0(c(n/2000, n/1000), "k"))  + 
-  scale_colour_gradientn(guide = 'none', paste0("Days to ", n/1000, "k\nmore cases"), space = "Lab", 
-    trans = "log10",
-    colours = darkblueseqpal
-    ) + 
-  scale_fill_gradientn(paste0("Days to ", n/1000, "k\nmore cases"), space = "Lab", 
-    trans = "log10",
-    colours = blueseqpal) + 
-  my_theme + 
-  theme(axis.title = element_blank(), 
-        legend.background = element_rect(fill = "transparent", colour = NA), 
-        legend.position = c(0, .66), legend.justification = c(0, 1), legend.direction = "horizontal")
-@
+\label{fig:sparklines-heatmap-nyt}
+\end{figure}
+
+One unique time-series plot created a representation of the case counts in the frequency domain: instead of plotting cases as they occurred, the chart instead shows line segments from 0 to N in $y$, where $N$ is a predefined number of cases; then the chart resets to show the next angle.
+This produces a sense of intensity of the waves of the pandemic over time as shown in \Cref{fig:freq-domain}.
+Interestingly, this frequency representation can also be easily transferred into an audio domain, providing access for visually impaired users as well as a multimodal representation of the data for sighted users.
+
+\begin{figure}
+\centering
+\includegraphics[width=.75\linewidth]{frequency-domain}
+\caption{Frequency-domain case count graphics.
+Here, each 1000 deaths are represented by a single line segment whose slope represents the time taken to reach the next 1000 deaths.
+This provides viewers with a sense of the pace of the epidemic, rather than the raw case counts in more standard time-series representations.
+This representation can also be shown in the auditory domain, providing access to those who are typically excluded from visualizations due to vision loss.
+Source: Romain Vuillemot, \url{https://observablehq.com/@romsson/sonification-of-covid-19-data}.}
+\label{fig:freq-domain}
+\end{figure}
 
@@
 
-Unfortunately, there are not a lot of good options for supporting forecasting decisions in mathematically novice viewers.
+Unfortunately, there are not a lot of good options for supporting forecasting decisions in mathematically unsophisticated viewers.
 Guide lines, like those used in the Financial Times, were helpful in the initial phases of the pandemic, but no outlets that we are aware of shifted the lines to show the growth rate of the current peak (as opposed to the initial peak), and due to the issues shown in \Cref{fig:log-scale-failures}, these guide lines may not have been all that successful in any case.

### 2022-01-17 JDSSV_manuscript_knitr.Rnw
Adding in the spiral and lining up cases by wave

@@
 
+There are many different ways that designers can use time-series data to provide additional contextual information and facilitate comparisons.
+\Cref{fig:ft-waves} shows covid cases, positivity rate, and hospital admissions from South Africa's Gauteng province, which was one of the first areas to experience a large wave of the Omicron variant.
+This chart provides context for the Omicron wave, showing that while infections are occurring at a much faster rate than in previous waves, hospital admissions seem to be approximately following previous trends.
+
+\begin{figure}\centering
+\includegraphics[width=\linewidth]{ft-waves}\\
+\includegraphics[width=\linewidth]{ft-waves-linear}
+\caption{Covid Cases, positivity rate, and hospital admissions from South Africa's Gauteng province, on a log scale (top) and linear scale (bottom).
+Successive waves are overlaid, showing that the increase in cases due to the Omicron variant was sharper than any previously measured increase, while hospitalizations were much more comparable to previous waves.
+Source: John Burn-Murdoch, \url{https://twitter.com/jburnmurdoch/status/1466480113487392769}}
+\label{fig:ft-waves}
+\end{figure}
+
+While traditionally, time-series information has been limited to charts with time on the x-axis, there have been many attempts to make use of the capabilities of web graphics and animation in order to convey temporal information in less traditional time-series forms.
+One controversial chart appeared in the New York Times; the x-axis was wrapped around an Archimedean spiral to provide a sense of periodicity in the year-by-year evolution of case counts, as shown in \Cref{fig:nyt-spiral}.
+The original form of this chart is prone to the line-width illusion \citep{vanderplasSignsSineIllusion2015}, in addition to the well-known problems with polar charts \citep{hofmannGraphicalTestsPower2012,waldnerComparisonRadialLinear2020}.
+The re-envisioned version, which alignes counts on the outside of the spiral, and displays them as radial lines, mitigates some of the issues with the line-width illusion by directly showing each line as its own entity, facilitating direct comparisons.
+Still, it is more work to interpret and compare peaks on this chart than on a similar linear time series chart with the same data, as the reader must assess line width and then do a mental rotation and shift operation in order to compare with any other time period.
+
+\begin{figure}\centering
+\includegraphics[width=.45\linewidth]{nyt-spiral}\hfill\includegraphics[width=.45\linewidth]{spiral-rework}
+\caption{Covid cases in the United States, 2020-2022.
+The original graph from the New York Times is on the left, and a reenvisioned version which is more perceptually friendly is on the right.
+Source: (Left) New York Times.
+Here's When We Expect Omicron to Peak \url{https://www.nytimes.com/2022/01/06/opinion/omicron-covid-us.html}, (Right) Covid Spiral, by Sourya Shrestha, \url{https://github.com/souryashrestha/covid_spiral}.}
+\label{fig:nyt-spiral}
+\end{figure}
+
 \section[]{Comparisons and rankings}

### 2022-01-17 JDSSV_manuscript_knitr.Rnw
Update graph with legend (screenshot)

@@
 The original form of this chart is prone to the line-width illusion \citep{vanderplasSignsSineIllusion2015}, in addition to the well-known problems with polar charts \citep{hofmannGraphicalTestsPower2012,waldnerComparisonRadialLinear2020}.
-The re-envisioned version, which alignes counts on the outside of the spiral, and displays them as radial lines, mitigates some of the issues with the line-width illusion by directly showing each line as its own entity, facilitating direct comparisons.
+The re-envisioned version, which aligns counts on the outside of the spiral, and displays them as radial lines, mitigates some of the issues with the line-width illusion by directly showing each line as its own entity, facilitating direct comparisons.
+In addition, the re-envisioned chart includes a reference line at 100K cases/day, which allows the reader to compare the severity of different waves directly.
 Still, it is more work to interpret and compare peaks on this chart than on a similar linear time series chart with the same data, as the reader must assess line width and then do a mental rotation and shift operation in order to compare with any other time period.
@@
 The original graph from the New York Times is on the left, and a reenvisioned version which is more perceptually friendly is on the right.
-Source: (Left) New York Times.
-Here's When We Expect Omicron to Peak \url{https://www.nytimes.com/2022/01/06/opinion/omicron-covid-us.html}, (Right) Covid Spiral, by Sourya Shrestha, \url{https://github.com/souryashrestha/covid_spiral}.}
+Source: (Left) New York Times \url{https://www.nytimes.com/2022/01/06/opinion/omicron-covid-us.html}, (Right) Covid Spiral, by Sourya Shrestha, \url{https://github.com/souryashrestha/covid_spiral}.}
 \label{fig:nyt-spiral}

### 2022-01-17 JDSSV_manuscript_knitr.Rnw
spacing
additional thoughts

@@
 Especially in the case of similarly coloured areas of the regions, viewers tend to integrate them unconsciously and perceive choropleths as representations of density.
-They also do not help to convey the desired message as the absolute numbers of Covid-19 cases are strongly influenced by the population size of the country, number of tests performed, and the accuracy of the recording and reporting system.
-\Cref{fig:scaled-choro} shows case load scaled by population, but is still subject to data quality and comparability issues related to testing and case reporting; these biases are hidden as the chart presents all data as equivalent and comparable when in fact it is likely that direct comparisons may be unwise in many cases.
+They also do not help to convey the desired message as the absolute numbers of Covid-19 cases are strongly influenced by the population size of the country, but also by the number of tests performed and the accuracy of the recording and reporting system.
 \item The area-bias: The visual impression is determined more by the colour and the geographical area of the individual countries than by the number of Covid cases.
 Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
-Cartograms, like that shown in \Cref{fig:hex-cartogram}, can mitigate the area bias by sacrificing the viewer's ease of geographic recognition.
-\item colour-scheme obstructions: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997}.
+\item Color-scheme obstructions: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997}.
 The choice of a continuous scale or a categorical scale, the choice of a scale that promotes the recognition of patterns or a scale that supports the filtering out of specific map details, influences the quality of choropleth maps.
@@
 \begin{figure*}
-	\includegraphics[width = 0.98\textwidth]{who_totalcases_choro}
+	\includegraphics[width = 0.98\textwidth]{Figures_Web/who_totalcases_choro.png}
 	\caption{Choropleth map of the Covid-19 cases world-wide by country. Source: WHO \url{https://covid19.who.int}}
@@
 \end{figure*}
+
+Choropleth maps are quite commonly used by media companies and governmental organisations and it is easy to find good and bad examples of their usage.
+
+Proportional symbol maps, see Fig.~\ref{fig:propsymb} or graduated symbol maps place scaled symbols or diagrams directly on the input map, often on the centroid of the regions.
+The symbol, most commonly a disk or a square, is scaled such that its area corresponds to the data value of the region.
+
+\begin{figure*}
+	\includegraphics[width = 0.98\textwidth]{Figures_Web/wp_totaldeaths_propsymb.png}
+	\caption{Proportional symbol map showing the cumulated reported Covid-19 related deaths adjusted for population world-wide by country.
+Source: Washington Post \url{https://www.washingtonpost.com/graphics/2020/world/mapping-spread-new-coronavirus/}}
+	\label{fig:propsymb}
+\end{figure*}
+
+And clearly, the visual impression of any map based illustration depends on the projection chosen for the underlying map as the examples quite often exercies a western-centric viewpoint.
+
+The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe.
+None of the above viusalisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them.
+The use of maps in this context more often focusses on the comparative aspects: how are we doing as opposed to our neighbors, which startegy to fight Covid-19 is better?
+Topics that we will look closer into in Section~\ref{sec:rankings}
 
@@
 \centering
-\includegraphics[width=.3\linewidth]{NYT-heatmap-crop}
-\includegraphics[width=.3\linewidth]{NYT-heatmap-crop3}
+\includegraphics[width=.3\linewidth]{NYT-heatmap-crop}\hfill
+\includegraphics[width=.3\linewidth]{NYT-heatmap-crop3}\hfill
 \includegraphics[width=.3\linewidth]{NYT-heatmap-crop2}
@@
 
+% bar-chart race https://www.youtube.com/watch?v=8WH-bTH6GqI
+
+% BBC flowers w/ sonification https://www.bbc.co.uk/news/resources/idt-7464500a-6368-4029-aa41-ab94e0ee09fb
+
+% Other animations?
+
 \section[]{Comparisons and rankings}
+\label{sec:rankings}
+
+The plague is always the others: This is the short formula for dealing with infectious diseases from a historical perspective \citep{thiessen2021}.
+Sociologists have coined the term "othering" for such attributions of the others \citep{mountz2009}.
+This refers to the observation that above all new, unknown threats are projected onto "strangers" and "the others".
+Closing the borders and restricting access to the country became a popular means of controlling the spread of the disease.
+Unsurprisingly, the prevailing visual narrative focused on comparisons often fuelled by political rivalry, historical dependencies, recent withdrawal from supranational institutions or regional competitions.
+An ongoing debate about the true extent of the dangers of covid-19 and how best to combat it, combined with daily availability and public access to data at all types of adminsitratory levels, fostered ongoing competition and the derivation of leaderboards.
 % https://twitter.com/AndToddsaid/status/1314231055096971265 is a horrible set of graphs and a great example of the snark associated with covid management policy and geographic comparisons...

### 2022-01-17 JDSSV_manuscript_knitr.tex
more thoughts

--- 2022-01-17 JDSSV_manuscript_knitr.Rnw
@@
-\documentclass[article]{jdssv}
+\documentclass[article]{jdssv}\usepackage[]{graphicx}\usepackage[]{color}
+% maxwidth is the original width if it is less than linewidth
+% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 
@@
 \usepackage{cleveref}
-
-\usepackage{subcaption}
 
@@
 }
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
 % \SweaveOpts{concordance=TRUE}
-<<include=FALSE, cache=FALSE, echo=FALSE, purl=FALSE>>=
-library(knitr)
-opts_chunk$set(comment = NA, size = 'footnotesize', prompt = TRUE, highlight = FALSE, 
-               cache = TRUE, crop = TRUE, concordance = FALSE, fig.align='center',
-               fig.path='Figures_R/fig-',out.width="0.4\\textwidth", 
-               message = F, error = F, warning = F, echo = F)
-opts_knit$set(out.format = "latex")
-knit_hooks$set(crop = hook_pdfcrop)
-options(width = 80, prompt = "R> ", continue = "    ", digits = 3)
-thm <- knit_theme$get('default')
-Sys.setenv(TEXINPUTS=getwd(),
-           BIBINPUTS=getwd(),
-           BSTINPUTS=getwd())
-knit_theme$set(thm)
-@
 
-<<get-nyt-data, include = F>>=
-
-# Read in the data from NYT
-library(tidyverse)
-library(lubridate)
-library(zoo)
-library(ggrepel)
-library(gridExtra)
-cvstate <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")
-
-cvstate <- cvstate %>%
-  group_by(state, fips) %>%
-  mutate(date_100th_case = min(date[cases >= 100]),
-         days_100th_case = date - date_100th_case,
-         daily_cases = cases - lag(cases, 1, default = 0),
-         week_avg_cases = rollmean(cases, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
-         daily_avg_cases = rollmean(daily_cases, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
-         daily_deaths = deaths - lag(deaths, 1, default = 0),
-         week_avg_deaths = rollmean(deaths, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
-         daily_avg_deaths = rollmean(daily_deaths, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T)) %>%
-  ungroup() %>%
-  mutate(state = factor(state))
-
-download.file("https://www2.census.gov/programs-surveys/popest/tables/2010-2019/state/totals/nst-est2019-01.xlsx", destfile = "census_pop_state.xlsx", quiet = T, mode = "wb")
-pop_data <- readxl::read_xlsx("census_pop_state.xlsx", col_names = c("state", "pop2019"), col_types = c("text", rep("skip", 11), "numeric"), skip = 9) %>%
-  mutate(state = str_remove(state, "[[:punct:]]"))
-
-cvstate <- left_join(cvstate, pop_data, by = "state")
-
-my_theme <- theme_bw() + 
-  theme(plot.background = element_rect(fill = "transparent", colour = NA), 
-        panel.background = element_rect(fill = "transparent", colour = NA), 
-        legend.background = element_rect(fill = "transparent", colour = NA),
-        axis.text.y = element_text(angle = 90, hjust = 0.5)) 
-theme_set(my_theme)
-@
-
-<<colourpal, include = F>>=
-# Including this for temporary compatibility - we may want to use *shudder* viridis?
-main_colours <- c("#d00000", "#f5f1e7", "#c7c8ca")
-
-# use as accents
-secondary_colours <- c("#001226", "#249ab5")
-
-# These are only to be used for infographics/charts
-tertiary_colours <- c("#bccb2a", "#f58a1f", "#005d84", "#ffd74f", "#a5228d")
-
-colour_palette <- c("#d00000", "#249ab5",  "#1B8500", "#f58a1f", "#a5228d", "#001226", "#B1EB66", "#005d84", "#c25700")
-
-heat_colour_palette <- c("#ffd74f", "#f58a1f", "#c25700", "#d00000", "#9D0000", "#6A0000")
-
-blueseqpal <- c("001226", "06293E", "0C3F56", "12566E", "186D85", "1E839D", 
-                "249AB5", 
-                "39ABC1", "4EBCCE", "64CDDA", "79DDE6", "8EEEF3", "A3FFFF") %>% 
-  paste0("#", .) %>% rev()
-
-blueseqpal_lt <- rev(c("#B2FFFF", "#C2FFFF", "#D1FFFF", "#E0FFFF", "#F0FFFF"))
-
-darkblueseqpal <- colorspace::darken(blueseqpal, .2)
-lightblueseqpal <- colorspace::lighten(blueseqpal, .2)
-@
 %% -- Introduction -------------------------------------------------------------
@@
 
-\begin{figure}
-\centering
-\begin{subfigure}[t]{.45\textwidth}
-\includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
-\caption{Hexagonal Cartogram of Covid-19 cases in the United States by state, per million people.
-This somewhat mitigates the area bias issue, but is still subject to other perceptual biases.
-Source: The COVID Tracking Project at The Atlantic \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
-\end{subfigure}\hfill
-\begin{subfigure}[t]{.45\textwidth}
-\includegraphics[width=\textwidth]{nyt-covid-scaled-choro}
-\caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
-\end{subfigure}
-\end{figure}
+Another illustrative example is a simulation from ZEIT Online\footnote{\url{https://www.zeit.de/wissen/gesundheit/2020-11/coronavirus-aerosole-ansteckungsgefahr-infektion-hotspot-innenraeume}}, which - based on models developed by a group of researchers at the Max Planck Institute for Chemistry - estimates the probability of an infected person infecting other people in closed rooms in various scenarios.
+While the visualisations present the simulated infection processes in a catchy way, the dependence of the simulations on parameter assumptions and settings is usually not addressed.
+Simulations should also always make transparent on which model assumptions and which data basis the simulations were created.
 
-Throughout the pandemic, both in choropleth maps and time-series graphs, creators have had to grapple with scale.
-When scales automatically adjust, mental comparisons across time can become misleading, because e.g.
-the colour scale from yesterday does not mean the same thing as it did today, as the range has expanded and the colours have stayed the same.
-This problem is more noticeable in choropleths than in time-series charts, where we are more likely to expect and notice a scale change, but it is still a fundamental issue in many different types of charts.
-Some publications leaned into this by arranging their graphics to highlight the fact that cases were ``off the charts", as in \Cref{fig:jbm-tweet}, implicitly cuing readers that the situation was exceptional and worthy of their attention.
-
-\begin{figure}
-\includegraphics[width=.99\linewidth]{ft-off-charts}
-\caption{Chart showing the increase in cases as a result of the B.1.617.2 variant in regions of the UK.
-Notably, this chart arranges areas to accommodate certain regions whose cases are ``off the chart" in order to emphasize the exceptional situation.
-Source: Tweet by John Burn-Murdoch (@jburnmurdoch), designer for the Financial Times.
-\url{https://twitter.com/jburnmurdoch/status/1397995388267810818}.}
-\label{fig:jbm-tweet}
-\end{figure}
-
-% Is it worth discussing e.g.
-https://www.nytimes.com/interactive/2021/us/covid-risk-map.html that does not make this distinction and runs into the "boy who cried wolf" problem?
-When everything is always extremely high risk we just tend to tune things out.
-I don't want to distract from the overview, though...
-https://twitter.com/rypan/status/1282020934044401664 is a good example
+\section{Time series and dynamics of the outbreak}
+% Line plots, frequency domain, NY times tile plots w/ cases represented in color intensity
 
@@
 It shows how an analysis of the associations between different COVID-19 strains and travel patterns can help understand the spread of the disease.
-
-Another illustrative example is a simulation from ZEIT Online\footnote{\url{https://www.zeit.de/wissen/gesundheit/2020-11/coronavirus-aerosole-ansteckungsgefahr-infektion-hotspot-innenraeume}}, which - based on models developed by a group of researchers at the Max Planck Institute for Chemistry - estimates the probability of an infected person infecting other people in closed rooms in various scenarios.
-While the visualisations present the simulated infection processes in a catchy way, the dependence of the simulations on parameter assumptions and settings is usually not addressed.
-Simulations should also always make transparent on which model assumptions and which data basis the simulations were created.
-
-\section{Time series}
-% Line plots, frequency domain, NY times tile plots w/ cases represented in colour intensity
-
-Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time, as there are new developments on a daily basis.
-Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis.
-This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state.
-Time series plots are the most natural choice for these goals, and over the course of the pandemic, many different attempts have been made to show time series information using different graphical forms.
-
-\begin{figure}
-\centering
-\includegraphics[width=.99\linewidth]{wapo-covid-monitoring}
-\caption{COVID monitoring summary charts on the Washington Post's Coronavirus page. Source: \url{https://www.washingtonpost.com/coronavirus/}, 2022-01-17.}
-\label{fig:wapo-covid-time-series}
-\end{figure}
-
-The most obvious time-series plot is also the most common: plotting cases against date, often with a 7 or 14-day smooth to handle periodicity in case reporting due to the work week, as shown in \Cref{fig:wapo-covid-time-series}.
-This allows for individuals to assess the current situation and easily compare the current number of cases to past times for which the individual has a direct reference for what to expect in impact to daily life.
-Often, as in \Cref{fig:wapo-covid-time-series}, these time series show multiple variables: cases, deaths, and hospitalizations (and sometimes vaccinations) to provide a more comprehensive view of the situation; this is particularly useful as there has been some decoupling of case numbers and hospitalizations/deaths with the emergence of effective vaccinations and strains which are more contagious but potentially also less severe.
-Many outlets also provide additional charts designed to provide some immediate context to readers, as shown in \Cref{fig:wapo-context}.
-
-\begin{figure}
-\centering
-\includegraphics[width=.5\linewidth]{wapo-covid-context}
-\caption{Some outlets provide contextual information to assist readers with interpreting time-series plots.
-Source: Washington Post Coronavirus Tracking, \url{https://www.washingtonpost.com/graphics/2020/national/coronavirus-us-cases-deaths/}.}
-\label{fig:wapo-context}
-\end{figure}
-
-However, there are some limitations with this presentation: if we are interested in showing the change in counts over time, viewers must assess the slope of the line, rather than its position.
-We know that slope judgments are much less accurate than position \citep{clevelandGraphicalPerceptionVisual1987}, even in relatively simple situations; when we add in our ability to assess exponential growth, our perceptual accuracy is even more suspect.
-
-Other time-series representations were more abstract, sacrificing exact data representations for a higher-level summary overview that allowed viewers to make comparisons between regions and points in time without the burden of numerical calculations.
-Several versions from the New York Times are shown in \Cref{fig:sparklines-heatmap-nyt}.
-The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data, however, they do not present precise numerical information and even the color scale is fairly opaque -- it would be extremely difficult to translate the visualization into any sort of accurate numerical estimate.
-In addition, by hiding these details from the user, it is very difficult to identify when the representation changes over time, as it did several times during the summer of 2020.
-The use of a similar color scheme as well as similar representations made it extremely difficult for users to identify that the value represented had changed.
-Clearly, the purpose of these charts is not to provide numerical precision, but rather a comparative assessment of the status of one region relative to others as well as relative to previous time points.
-
-\begin{figure}
-\centering
-\includegraphics[width=.3\linewidth]{NYT-heatmap-crop}\hfill
-\includegraphics[width=.3\linewidth]{NYT-heatmap-crop3}\hfill
-\includegraphics[width=.3\linewidth]{NYT-heatmap-crop2}
-\caption{Sparklines style heatmaps embedded in a table of coronavirus data by state.
-Over the course of Summer 2020, the specific measures shown changed, even though the form of the data remained highly similar.
-Source: \url{https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html} (May 6, 2020; June 15, 2020; and July 24, 2020, respectively) }
-
-\label{fig:sparklines-heatmap-nyt}
-\end{figure}
-
-One unique time-series plot created a representation of the case counts in the frequency domain: instead of plotting cases as they occurred, the chart instead shows line segments from 0 to N in $y$, where $N$ is a predefined number of cases; then the chart resets to show the next angle.
-This produces a sense of intensity of the waves of the pandemic over time as shown in \Cref{fig:freq-domain}.
-Interestingly, this frequency representation can also be easily transferred into an audio domain, providing access for visually impaired users as well as a multimodal representation of the data for sighted users.
-
-\begin{figure}
-\centering
-\includegraphics[width=.75\linewidth]{frequency-domain}
-\caption{Frequency-domain case count graphics.
-Here, each 1000 deaths are represented by a single line segment whose slope represents the time taken to reach the next 1000 deaths.
-This provides viewers with a sense of the pace of the epidemic, rather than the raw case counts in more standard time-series representations.
-This representation can also be shown in the auditory domain, providing access to those who are typically excluded from visualizations due to vision loss.
-Source: Romain Vuillemot, \url{https://observablehq.com/@romsson/sonification-of-covid-19-data}.}
-\label{fig:freq-domain}
-\end{figure}
-
-There are many different ways that designers can use time-series data to provide additional contextual information and facilitate comparisons.
-\Cref{fig:ft-waves} shows covid cases, positivity rate, and hospital admissions from South Africa's Gauteng province, which was one of the first areas to experience a large wave of the Omicron variant.
-This chart provides context for the Omicron wave, showing that while infections are occurring at a much faster rate than in previous waves, hospital admissions seem to be approximately following previous trends.
-
-\begin{figure}\centering
-\includegraphics[width=\linewidth]{ft-waves}\\
-\includegraphics[width=\linewidth]{ft-waves-linear}
-\caption{Covid Cases, positivity rate, and hospital admissions from South Africa's Gauteng province, on a log scale (top) and linear scale (bottom).
-Successive waves are overlaid, showing that the increase in cases due to the Omicron variant was sharper than any previously measured increase, while hospitalizations were much more comparable to previous waves.
-Source: John Burn-Murdoch, \url{https://twitter.com/jburnmurdoch/status/1466480113487392769}}
-\label{fig:ft-waves}
-\end{figure}
-
-While traditionally, time-series information has been limited to charts with time on the x-axis, there have been many attempts to make use of the capabilities of web graphics and animation in order to convey temporal information in less traditional time-series forms.
-One controversial chart appeared in the New York Times; the x-axis was wrapped around an Archimedean spiral to provide a sense of periodicity in the year-by-year evolution of case counts, as shown in \Cref{fig:nyt-spiral}.
-The original form of this chart is prone to the line-width illusion \citep{vanderplasSignsSineIllusion2015}, in addition to the well-known problems with polar charts \citep{hofmannGraphicalTestsPower2012,waldnerComparisonRadialLinear2020}.
-The re-envisioned version, which aligns counts on the outside of the spiral, and displays them as radial lines, mitigates some of the issues with the line-width illusion by directly showing each line as its own entity, facilitating direct comparisons.
-In addition, the re-envisioned chart includes a reference line at 100K cases/day, which allows the reader to compare the severity of different waves directly.
-Still, it is more work to interpret and compare peaks on this chart than on a similar linear time series chart with the same data, as the reader must assess line width and then do a mental rotation and shift operation in order to compare with any other time period.
-
-\begin{figure}\centering
-\includegraphics[width=.45\linewidth]{nyt-spiral}\hfill\includegraphics[width=.45\linewidth]{spiral-rework}
-\caption{Covid cases in the United States, 2020-2022.
-The original graph from the New York Times is on the left, and a reenvisioned version which is more perceptually friendly is on the right.
-Source: (Left) New York Times \url{https://www.nytimes.com/2022/01/06/opinion/omicron-covid-us.html}, (Right) Covid Spiral, by Sourya Shrestha, \url{https://github.com/souryashrestha/covid_spiral}.}
-\label{fig:nyt-spiral}
-\end{figure}
-
-% bar-chart race https://www.youtube.com/watch?v=8WH-bTH6GqI
-
-% BBC flowers w/ sonification https://www.bbc.co.uk/news/resources/idt-7464500a-6368-4029-aa41-ab94e0ee09fb
-
-% Other animations?
 
@@
 Even researchers do not always read and interpret log scales correctly\citep{mengeLogarithmicScalesEcological2018}; expecting the general public to do so is difficult under normal circumstances\citep{hecklerStudentAccuracyReading2013} is difficult.
-When panic, fear, uncertainty, and doubt about the situation are added to the mix, it is easy to imagine that we become even worse when interpreting graphics.
+When panic, fear, uncertainty, and doubt about the situation are added to the mix,
 % Discussion of how this section's conclusions change with the pandemic
@@
 
-<<log-scale-initial, fig.width = 8, fig.height = 4, out.width = "\\linewidth", fig.cap = "In the early stages of the pandemic, log scales allowed the comparison of raw case counts in locations with vastly different population and case counts.">>=
-# remotes::install_github("AllanCameron/geomtextpath")
-library(geomtextpath)
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{0.961, 0.961, 0.961}\color{fgcolor}\begin{figure}
 
-library(facetscales)
-scales_y <- list(
-  Log = scale_y_log10(),
-  Linear = scale_y_continuous()
-)
-tmp <- cvstate %>% 
-  filter(date < ymd("2020-06-01")) %>%
-  filter(state %in% c("Michigan", "Washington", "New York", "California"))
+{\centering \includegraphics[width=\linewidth]{Figures_R/fig-log-scale-initial-1}
 
-p1 <-  tmp %>%
-  ggplot(aes(x = date, y = daily_avg_cases, group = state)) +
-  geom_textline(aes(colour = state, label = state, vjust = -0.1, hjust =.93) , text_smoothing = 50) + 
-  ylab("Daily average cases (7 day average)") + 
-  xlab("Date") + 
-  guides(colour = "none") + 
-  scale_y_log10() + 
-  facet_grid(~"Log") + 
-  theme_bw()
+}
 
-p2 <-  tmp %>%
-  ggplot(aes(x = date, y = daily_avg_cases, group = state)) +
-  geom_textline(aes(colour = state, label = state, vjust = -0.25, hjust = .85) , text_smoothing = 60) + 
-  ylab("Daily average cases (7 day average)") + 
-  xlab("Date") + 
-  guides(colour = "none") + 
-  facet_grid(~"Linear") + 
-  theme_bw() +
-  theme(axis.title.y = element_blank())
+\caption[In the early stages of the pandemic, log scales allowed the comparison of raw case counts in locations with vastly different population and case counts]{In the early stages of the pandemic, log scales allowed the comparison of raw case counts in locations with vastly different population and case counts.}\label{fig:log-scale-initial}
+\end{figure}
 
-grid.arrange(p1, p2, nrow = 1)
-@
+\end{knitrout}
 
@@
 Diagonal reference lines from the origin were also less helpful, as the growth of cases or deaths was no longer approximately exponential and varied over time; for these reference lines to be effective there would need to be a clear idea of when the case counts started to increase exponentially, which is difficult to determine whilst in the thick of a potential COVID wave.
-Occasionally graphs with manually drawn reference lines were made available, but usually only in a retrospective manner, as in \Cref{fig:log-scale-ref-multiple}.
 
-<<log-scale-failures, echo = F, fig.width = 8, fig.height = 4, out.width = ".95\\linewidth", message = F, error = F, warning = F, fig.cap = "One problem with log scales is that if there is a background level of spread, it can be hard to notice the introduction of an additional source of exponential spread.
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{0.961, 0.961, 0.961}\color{fgcolor}\begin{figure}
+
+{\centering \includegraphics[width=.95\linewidth]{Figures_R/fig-log-scale-failures-1}
+
+}
+
+\caption[One problem with log scales is that if there is a background level of spread, it can be hard to notice the introduction of an additional source of exponential spread]{One problem with log scales is that if there is a background level of spread, it can be hard to notice the introduction of an additional source of exponential spread.
 Linear scales do not have this problem - the exponential source is noticeable very quickly in the total line, but on the log scale it is much harder to discern when the exponential source causes the total line to diverge from the background.
-In the top-right corner, it is difficult to identify that there is an exponential increase in cases amid the baseline, even though the exponential source makes up approximately 50\\% of the cases at the end of the time period shown." >>=
-# Set up labeller to tweak things 
-rlistmap <- function(x, myfun) {
-  if ("list" %in% class(x)) return(purrr::map(x, ~rlistmap(., myfun)))
+In the top-right corner, it is difficult to identify that there is an exponential increase in cases amid the baseline, even though the exponential source makes up approximately 50\% of the cases at the end of the time period shown.}\label{fig:log-scale-failures}
+\end{figure}
 
-  return(myfun(x))
-}
-fix_labels <- function(x) {
-  x  %>%
-    str_replace('scale: (.*)', "\\1 scale") %>%
-    str_replace("baseline: (\\d{1,})", "Baseline:\n\\1 cases")
-}
-label_custom <- function(labels, multi_line = T, sep = ": ") {
-  out <- label_both(labels, multi_line = multi_line, sep = sep)
+\end{knitrout}
 
-  out <- rlistmap(out, fix_labels)
-
-  out
-}
-class(label_custom) <- "labeller"
-
-# Set up data 
-growth <- rpois(30, 15*exp((1:30 - 20)/5))
-
-tmp <- tibble(baseline = 15,
-       x = 1:30,
-       y = rpois(30, 15),
-       y1 = growth) %>%
-  bind_rows(
-    tibble(baseline = 50,
-       x = 1:30,
-       y = rpois(30, 50),
-       y1 = growth) 
-  ) %>%
-  bind_rows(
-    tibble(baseline = 200,
-       x = 1:30,
-       y = rpois(30, 200),
-       y1 = growth) 
-  ) %>%
-  pivot_longer(matches("y")) %>%
-  mutate(name = str_replace_all(name, c("y$" = "Background", "y1" = "Exponential")))  %>%
-  group_by(baseline, x) %>%
-  mutate(Total = sum(value)) %>%
-  pivot_wider(names_from = "name", values_from = "value") %>%
-  pivot_longer(matches("Background|Exponential|Total")) %>%
-  mutate(name = factor(name, levels = c("Total", "Exponential", "Background")))
-
-tmp <- bind_rows(mutate(tmp, scale = "Log"), mutate(tmp, scale = "Linear")) %>%
-  mutate(Baseline = reorder(paste0("Baseline: ", baseline, " cases"), baseline)) %>%
-  mutate(scale = factor(scale, levels = c("Log", "Linear")))
-
-# Set up scales 
-scales_y <- list(
-  Log = scale_y_log10(),
-  Linear = scale_y_continuous()
-)
-
-tmp %>%
-ggplot(aes(x = x, y = value, colour = name, size = name)) + 
-  geom_line(alpha = .9) +
-  facet_grid_sc(rows = vars(scale), cols = vars(Baseline), 
-                scales = list(x = "fixed", y = scales_y)) +
-  scale_size_manual(values = c(1.75, 1, 1)) +
-  scale_colour_manual(values = c("#249ab5", "#d00000", "#001226")) +
-  scale_fill_discrete(name = "") + 
-  theme_bw() + 
-  theme(legend.position = "bottom", axis.title = element_blank(),
-        legend.title = element_blank())
-@
-
-% Focus primarily on time-series graphs (as opposed to log scales in colour)
-
-\begin{figure}
-\centering
-\includegraphics[width=.8\linewidth]{ft-reference-lines}
-\caption{An annotated time series from the Financial Times which appeared on Twitter on November 12, 2020.
-This chart has annotations which show the decreasing $R_{eff}$ in successive peaks of the pandemic, with periods of $R_{eff}<1$ between surges in cases.
-These annotations assist the reader with drawing complex conclusions from the chart, but are difficult to automate and thus tend to be manually curated.
-Source: Mark Gubrud (@mgubrud), \url{https://twitter.com/mgubrud/status/1326784082399944704}.}
-\label{fig:log-scale-ref-multiple}
-\end{figure}
+% Focus primarily on time-series graphs (as opposed to log scales in color)
 
@@
 
-<<linear-scales-ref-lines, fig.width = 8, fig.height = 4, out.width = ".8\\linewidth", fig.cap = "Reported COVID cases in New York State, 2020-2022.
-The linear scale makes it difficult to compare the trajectory of different waves to determine how severe the current status is relative to the past, because the primary contrast is the height of the relative peaks, rather than the growth \\emph{rate}.
-A similar graph on the log scale would have the peaks at much more similar heights (though there would still be a difference), allowing the reader to focus on other information, such as the slope of the relative lines.">>=
-cvstate %>%
-  filter(state == "New York") %>%
-  ggplot(aes(x = date, y = daily_avg_cases)) + 
-  geom_line() + 
-  ggtitle("Reported COVID cases in New York State, 2020-2022") +
-  ylab("Daily Cases reported (7-day average)") + 
-  xlab("Date") + theme_bw()
-@
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{0.961, 0.961, 0.961}\color{fgcolor}\begin{figure}
+
+{\centering \includegraphics[width=.8\linewidth]{Figures_R/fig-linear-scales-ref-lines-1}
+
+}
+
+\caption[Reported COVID cases in New York State, 2020-2022]{Reported COVID cases in New York State, 2020-2022.
+The linear scale makes it difficult to compare the trajectory of different waves to determine how severe the current status is relative to the past, because the primary contrast is the height of the relative peaks, rather than the growth \emph{rate}.
+A similar graph on the log scale would have the peaks at much more similar heights (though there would still be a difference), allowing the reader to focus on other information.}\label{fig:linear-scales-ref-lines}
+\end{figure}
+
+\end{knitrout}
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{0.961, 0.961, 0.961}\color{fgcolor}\begin{figure}
+
+{\centering \includegraphics[width=.8\linewidth]{Figures_R/fig-log-scales-ref-lines-1}
+
+}
+
+\caption[Reported COVID cases in New York State, 2020-2022]{Reported COVID cases in New York State, 2020-2022.
+The linear scale makes it difficult to compare the trajectory of different waves to determine how severe the current status is relative to the past.}\label{fig:log-scales-ref-lines}
+\end{figure}
+
+\end{knitrout}
 
@@
 It is not clear that the use of log or linear scales during the COVID-19 pandemic had a large effect on public opinion.
-Several studies were conducted in the early stages of the pandemic \citep{romanoScaleCOVID19Graphs2020, seviLogarithmicLinearVisualizations2020, ryanGraphsLogarithmicAxes2020} and results seem to suggest that while individuals have difficulty understanding log scale graphs, these issues do not tend to affect their support for intervention measures, perhaps in part because COVID-related news saturated the news and opinions were set outside of information provided in the experiments.
-This saturation makes it difficult to study the graphical influences while removing effects of popular opinion, political leaning, and emotional sentiment relating to the pandemic.
-
-If we evaluate the use of log and linear scales under the more general context of exponential (or near-exponential) growth rates, however, we can gain some clarity as to their use in this particular context.
-We are abysmal at forecasting exponential growth, vastly under-estimating future growth by using linear or quadratic approximations \citep{wagenaarExtrapolationExponentialTime1978a,lawrenceExploringJudgementalForecasting1992,timmersInverseStatisticsMisperception1977a}.
-If the goal of a chart of COVID case counts is to allow individuals to forecast the trend and make decisions accordingly, then using a log scale (with all of its pitfalls in understanding) may be the best option, as it at least replaces the need to forecast along an exponential curve with the need to forecast along a straight line.
-However, it is not clear that individuals can transfer that prediction back to a linear scale.
-If the goal of presenting a chart of case counts is to tell the story of what has happened (rather than supporting future decision-making), then it is undoubtedly better to present the data on a more familiar linear scale that requires less cognitive load on the part of the viewer.
-
-Unfortunately, there are not a lot of good options for supporting forecasting decisions in mathematically unsophisticated viewers.
-Guide lines, like those used in the Financial Times, were helpful in the initial phases of the pandemic, but no outlets that we are aware of shifted the lines to show the growth rate of the current peak (as opposed to the initial peak), and due to the issues shown in \Cref{fig:log-scale-failures}, these guide lines may not have been all that successful in any case.
-Those hoping to create successful time-series charts that allow forecasting are thus left with two options: deal with our inability to forecast exponential growth, or deal with most individuals' inability to translate log scales into practical reality.
+Several studies were conducted in the early stages of the pandemic \citep{romanoScaleCOVID19Graphs2020, seviLogarithmicLinearVisualizations2020, ryanGraphsLogarithmicAxes2020} and results seem to suggest that while individuals have difficulty understanding log scale graphs, these issues do not tend to affect their support for intervention measures (perhaps in part because COVID-related news saturated the news and opinions were set outside of information provided in the experiments).
 
@@
 such as version numbers, operating systems, or compilers could be included
-in an unnumbered section. Also, auxiliary packages (say, for visualisations,
+in an unnumbered section. Also, auxiliary packages (say, for visualizations,
 maps, tables, \dots) that are not cited in the main text can be credited here.

### 2022-01-17 JDSSV_manuscript_knitr.tex
a few more thoughts

--- 2022-01-17 JDSSV_manuscript_knitr.Rnw
@@
 
-<<log-scale-failures, echo = F, fig.width = 8, fig.height = 4, out.width = ".95\\linewidth", message = F, error = F, warning = F, fig.cap = "One problem with log scales is that if there is a background level of spread, it can be hard to notice the introduction of an additional source of exponential spread.
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{0.961, 0.961, 0.961}\color{fgcolor}\begin{figure}
+
+{\centering \includegraphics[width=.95\linewidth]{Figures_R/fig-log-scale-failures-1}
+
+}
+
+\caption[One problem with log scales is that if there is a background level of spread, it can be hard to notice the introduction of an additional source of exponential spread]{One problem with log scales is that if there is a background level of spread, it can be hard to notice the introduction of an additional source of exponential spread.
 Linear scales do not have this problem - the exponential source is noticeable very quickly in the total line, but on the log scale it is much harder to discern when the exponential source causes the total line to diverge from the background.
-In the top-right corner, it is difficult to identify that there is an exponential increase in cases amid the baseline, even though the exponential source makes up approximately 50\\% of the cases at the end of the time period shown." >>=
-# Set up labeller to tweak things 
-rlistmap <- function(x, myfun) {
-  if ("list" %in% class(x)) return(purrr::map(x, ~rlistmap(., myfun)))
+In the top-right corner, it is difficult to identify that there is an exponential increase in cases amid the baseline, even though the exponential source makes up approximately 50\% of the cases at the end of the time period shown.}\label{fig:log-scale-failures}
+\end{figure}
 
-  return(myfun(x))
-}
-fix_labels <- function(x) {
-  x  %>%
-    str_replace('scale: (.*)', "\\1 scale") %>%
-    str_replace("baseline: (\\d{1,})", "Baseline:\n\\1 cases")
-}
-label_custom <- function(labels, multi_line = T, sep = ": ") {
-  out <- label_both(labels, multi_line = multi_line, sep = sep)
-
-  out <- rlistmap(out, fix_labels)
-
-  out
-}
-class(label_custom) <- "labeller"
-
-# Set up data 
-growth <- rpois(30, 15*exp((1:30 - 20)/5))
-
-tmp <- tibble(baseline = 15,
-       x = 1:30,
-       y = rpois(30, 15),
-       y1 = growth) %>%
-  bind_rows(
-    tibble(baseline = 50,
-       x = 1:30,
-       y = rpois(30, 50),
-       y1 = growth) 
-  ) %>%
-  bind_rows(
-    tibble(baseline = 200,
-       x = 1:30,
-       y = rpois(30, 200),
-       y1 = growth) 
-  ) %>%
-  pivot_longer(matches("y")) %>%
-  mutate(name = str_replace_all(name, c("y$" = "Background", "y1" = "Exponential")))  %>%
-  group_by(baseline, x) %>%
-  mutate(Total = sum(value)) %>%
-  pivot_wider(names_from = "name", values_from = "value") %>%
-  pivot_longer(matches("Background|Exponential|Total")) %>%
-  mutate(name = factor(name, levels = c("Total", "Exponential", "Background")))
-
-tmp <- bind_rows(mutate(tmp, scale = "Log"), mutate(tmp, scale = "Linear")) %>%
-  mutate(Baseline = reorder(paste0("Baseline: ", baseline, " cases"), baseline)) %>%
-  mutate(scale = factor(scale, levels = c("Log", "Linear")))
-
-# Set up scales 
-scales_y <- list(
-  Log = scale_y_log10(),
-  Linear = scale_y_continuous()
-)
-
-tmp %>%
-ggplot(aes(x = x, y = value, colour = name, size = name)) + 
-  geom_line(alpha = .9) +
-  facet_grid_sc(rows = vars(scale), cols = vars(Baseline), 
-                scales = list(x = "fixed", y = scales_y)) +
-  scale_size_manual(values = c(1.75, 1, 1)) +
-  scale_colour_manual(values = c("#249ab5", "#d00000", "#001226")) +
-  scale_fill_discrete(name = "") + 
-  theme_bw() + 
-  theme(legend.position = "bottom", axis.title = element_blank(),
-        legend.title = element_blank())
-@
+\end{knitrout}
 
@@
 
-<<linear-scales-ref-lines, fig.width = 8, fig.height = 4, out.width = ".8\\linewidth", fig.cap = "Reported COVID cases in New York State, 2020-2022.
-The linear scale makes it difficult to compare the trajectory of different waves to determine how severe the current status is relative to the past, because the primary contrast is the height of the relative peaks, rather than the growth \\emph{rate}.
-A similar graph on the log scale would have the peaks at much more similar heights (though there would still be a difference), allowing the reader to focus on other information, such as the slope of the relative lines.">>=
-cvstate %>%
-  filter(state == "New York") %>%
-  ggplot(aes(x = date, y = daily_avg_cases)) + 
-  geom_line() + 
-  ggtitle("Reported COVID cases in New York State, 2020-2022") +
-  ylab("Daily Cases reported (7-day average)") + 
-  xlab("Date") + theme_bw()
-@
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{0.961, 0.961, 0.961}\color{fgcolor}\begin{figure}
+
+{\centering \includegraphics[width=.8\linewidth]{Figures_R/fig-linear-scales-ref-lines-1}
+
+}
+
+\caption[Reported COVID cases in New York State, 2020-2022]{Reported COVID cases in New York State, 2020-2022.
+The linear scale makes it difficult to compare the trajectory of different waves to determine how severe the current status is relative to the past, because the primary contrast is the height of the relative peaks, rather than the growth \emph{rate}.
+A similar graph on the log scale would have the peaks at much more similar heights (though there would still be a difference), allowing the reader to focus on other information, such as the slope of the relative lines.}\label{fig:linear-scales-ref-lines}
+\end{figure}
+
+\end{knitrout}
 

### 2022-01-18 JDSSV_manuscript_knitr.Rnw
a few more thoughts
a bit on comparison and ranking

@@
 An ongoing debate about the true extent of the dangers of covid-19 and how best to combat it, combined with daily availability and public access to data at all types of adminsitratory levels, fostered ongoing competition and the derivation of leaderboards.
+
+\begin{figure}\centering
+\includegraphics[width=\linewidth]{Figures_Web/ft_deathrates_bar}\\
+\includegraphics[width=\linewidth]{Figures_Web/ft_deathcounts_area}
+\caption{Ordered death rates in countries and the temporal evolution of the death toll share in major world regions.
+Source: Financial Times, \url{https://www.ft.com/content/a2901ce8-5eb7-4633-b89c-cbdf5b386938}}
+\label{fig:ft-death-ranking}
+\end{figure}
+
+The bar charts used for this purpose (see top image in Fig.~\ref{fig:ft-death-ranking}) are straigthforward and provide typically a good way to show the "leading" countries or regions.
+Emphasis on specific countries is either preconfigured by the creator of the chart or can be modified interactively in the online version by check boxes and drop-down menus.
+These bar charts also allow easy adjustment of the count data to population size or other meaningful standard.
+Streamgraphs (see lower image in Fig.~\ref{fig:ft-death-ranking}) are a special type of stacked area graphs and became quite popular around 2008, being used by the New York Times to visualise box office results of movies \citep{streamgraph}.
+Their usefulness depends heavily on the clarity of the pattern.
+Various design choices such as the ordering of the different groups have a strong impact on the visual impression and message conveyed by the graph.
+Small variations in the proportions of the groups are almost impossible to detect, but streamgraphs give a good indication of which group is predominant at any given time.
+
 % https://twitter.com/AndToddsaid/status/1314231055096971265 is a horrible set of graphs and a great example of the snark associated with covid management policy and geographic comparisons...

### 2022-01-22 JDSSV_manuscript_knitr.Rnw
Additional information on comparisons and misinformation as well as some tweaks to the english in the first couple of sections

@@
 
-At the same time we still see a lot of defective graphics disseminated and shared: some that violate fundamental visualisation and statistical reporting principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility.
+At the same time, we still see a lot of defective graphics disseminated and shared: some that violate fundamental visualisation and statistical reporting principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility.
 These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org).
@@
 Nevertheless, both the complexity of phenomena and the "bipolarity" of statistical thinking remain a challenge.
-
 While human thinking tends to simplify patterns and political communication also prefers a simple cause-effect relationship, real phenomena are often multivariate.
@@
 
-\section[]{The global perspective}
+\section{The global perspective}
 
-On 11 March 2020, WHO declared the outbreak of the novel coronavirus disease (COVID-19) a pandemic, and since that date at the latest, the global perspective of the disease has been in the public eye.
-The spatial spread of the virus and the resulting cases and deaths are commonly visualized by choropleth maps, see for example Figure~\ref{fig:choro1} showing the total number of infections reported in each country as of January 14, 2022.
-For the pandemic perspective a central element of the narrative is the ubiquity of the disease and the accompanying global impact.
+On 11 March 2020, WHO declared the outbreak of the novel coronavirus disease (COVID-19) a pandemic, and since that date at the latest, the global distribution of the disease has been in the public eye.
+The spatial spread of the virus and the resulting cases and deaths are commonly visualized by choropleth maps, see for example \Cref{fig:choro1} showing the total number of infections reported in each country as of January 14, 2022.
+For the pandemic perspective, a central element of the narrative is the ubiquity of the disease and the accompanying global impact.
 Choropleth maps based on raw numbers of cases might look convincing and fit to the purpose, but neglect a number of well-know caveats for statistical reporting and visualisation:
@@
 Choropleth maps are quite commonly used by media companies and governmental organisations and it is easy to find good and bad examples of their usage.
-
 Proportional symbol maps, see Fig.~\ref{fig:propsymb} or graduated symbol maps place scaled symbols or diagrams directly on the input map, often on the centroid of the regions.
@@
 
-And clearly, the visual impression of any map based illustration depends on the projection chosen for the underlying map as the examples quite often exercies a western-centric viewpoint.
+And clearly, the visual impression of any map based illustration depends on the projection chosen for the underlying map as the examples quite often demonstrate a western-centric viewpoint.
 
@@
 None of the above viusalisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them.
-The use of maps in this context more often focusses on the comparative aspects: how are we doing as opposed to our neighbors, which startegy to fight Covid-19 is better?
-Topics that we will look closer into in Section~\ref{sec:rankings}
+The use of maps in this context more often focuses on the comparative aspects: how are we doing as opposed to our neighbors, which strategy to fight Covid-19 is better?
+We will examine these topics more closely in Section~\ref{sec:rankings}.
 
@@
 Still, it is more work to interpret and compare peaks on this chart than on a similar linear time series chart with the same data, as the reader must assess line width and then do a mental rotation and shift operation in order to compare with any other time period.
+In addition, this type of chart devotes less area to previous years of the pandemic, which may decrease the amount of focus given to that data; whether this is an advantage or disadvantage depends on your perspective and goals when using the chart.
 
@@
 
-% bar-chart race https://www.youtube.com/watch?v=8WH-bTH6GqI
+Another attempt to creatively show time-series data that has been extremely common online throughout the pandemic is the bar chart race: an animated series of bar charts over time that show how case counts in each country are increasing.
+A typical example can be found on YouTube at \url{https://www.youtube.com/watch?v=8WH-bTH6GqI}; a screenshot is shown in \Cref{fig:barchart-race}.
+This type of chart is effective in showing instantaneous changes in counts over days for countries with an extremely high number of cases, but may make it difficult for the viewer to process how cases are changing in countries that do not dominate case counts.
+By focusing viewer attention on changes in relative totals of case counts, the graphic tends to hide steady, proportional increases in cases across the globe, making it somewhat difficult to perceive large global trends relative to small changes in comparative case counts.
+
+\begin{figure}
+\centering
+\includegraphics[width=.8\linewidth]{Figures_Web/Bar-chart-race-screenshot}
+\caption{A screenshot from Stats on Clock's bar chart race showing counts of covid cases over time throughout the pandemic.
+Source: \url{https://www.youtube.com/watch?v=8WH-bTH6GqI}}\label{fig:barchart-race}
+\end{figure}
 
@@
 
-\section[]{Comparisons and rankings}
+\section{Comparisons and rankings}
 \label{sec:rankings}
@@
 The plague is always the others: This is the short formula for dealing with infectious diseases from a historical perspective \citep{thiessen2021}.
-Sociologists have coined the term "othering" for such attributions of the others \citep{mountz2009}.
-This refers to the observation that above all new, unknown threats are projected onto "strangers" and "the others".
-Closing the borders and restricting access to the country became a popular means of controlling the spread of the disease.
-Unsurprisingly, the prevailing visual narrative focused on comparisons often fuelled by political rivalry, historical dependencies, recent withdrawal from supranational institutions or regional competitions.
-An ongoing debate about the true extent of the dangers of covid-19 and how best to combat it, combined with daily availability and public access to data at all types of adminsitratory levels, fostered ongoing competition and the derivation of leaderboards.
+Sociologists have coined the term ``othering'' for such attributions of the others \citep{mountz2009}.
+This refers to the observation that above all new, unknown threats are projected onto ``strangers'' and ``the others''.
+Closing the borders and restricting access to the country became a popular means of controlling the spread of the disease at several different points during the pandemic.
+Unsurprisingly, the prevailing visual narrative focused on comparisons often fueled by political rivalry, historical dependencies, recent withdrawal from supranational institutions or regional competitions.
+An ongoing debate about the true extent of the dangers of covid-19 and how best to combat it, combined with daily availability and public access to data across administrative levels, fostered ongoing competition and the use of leaderboards to show how bad case counts were somewhere else.
 
@@
 
-The bar charts used for this purpose (see top image in Fig.~\ref{fig:ft-death-ranking}) are straigthforward and provide typically a good way to show the "leading" countries or regions.
+The bar charts used for this purpose (see top image in Fig.~\ref{fig:ft-death-ranking}) are straightforward and provide a good way to show the ``leading'' countries or regions.
 Emphasis on specific countries is either preconfigured by the creator of the chart or can be modified interactively in the online version by check boxes and drop-down menus.
 These bar charts also allow easy adjustment of the count data to population size or other meaningful standard.
-Streamgraphs (see lower image in Fig.~\ref{fig:ft-death-ranking}) are a special type of stacked area graphs and became quite popular around 2008, being used by the New York Times to visualise box office results of movies \citep{streamgraph}.
+Streamgraphs (see lower image in Fig.~\ref{fig:ft-death-ranking}) are a special type of stacked area graphs and became quite popular around 2008 when they were used by the New York Times to visualise box office results of movies \citep{streamgraph}.
 Their usefulness depends heavily on the clarity of the pattern.
 Various design choices such as the ordering of the different groups have a strong impact on the visual impression and message conveyed by the graph.
-Small variations in the proportions of the groups are almost impossible to detect, but streamgraphs give a good indication of which group is predominant at any given time.
+Small variations in the proportions of the groups are almost impossible to detect, but streamgraphs give a good indication of which group is predominant at any given time, even though they are subject to the line-width or sine illusion \citep{vanderplasSignsSineIllusion2015}.
 
-% https://twitter.com/AndToddsaid/status/1314231055096971265 is a horrible set of graphs and a great example of the snark associated with covid management policy and geographic comparisons...
+Often, these comparative charts were combined with policy discussions, with individuals challenged to spot certain policy interventions on the case-count graphs of different localities with different policies\citep{weissThese12Graphs2020}.
+\Cref{fig:policy-comparison} shows a chart from an article which appeared in The Federalist, a conservative outlet in the United States, suggesting that mask mandates are ineffective because it is difficult to spot the impact of the mask mandate on the overall trend of cases.
+Of course, mask mandates are but one component of a much broader pandemic management strategy, but the goal of the chart is clear: the reader is supposed to conclude that cases and masks are not associated.
+These charts were so common on social media that media outlets ran stories identifying them as misinformation \citep{reutersstaffFactCheckMask2020}; clearly, comparative charts were effectively deployed to misinform as well as to inform.
+
+\begin{figure}
+\centering
+\includegraphics[width=.8\linewidth]{Figures_Web/Deaths-NY-TX-GA-Sweden-Federalist.jpeg}
+\caption{Deaths per million people in New York, Sweden, Texas, and Georgia, as shown in the Federalist \citep{weissThese12Graphs2020}.
+The goal of this comparison chart is to lead the reader to conclude that masks are ineffective, as there is not a clear difference between Texas and Georgia.
+New York is also shown (though not labeled as requiring or not requiring masks), and Sweden is shown as well, despite not being a US state at all, leading the critical reader to conclude that the time series shown in this chart have been cherry-picked to make a rhetorical point.}\label{fig:policy-comparison}
+\end{figure}
 

### 2022-01-22 JDSSV_manuscript_knitr.tex
Additional information on comparisons and misinformation as well as some tweaks to the english in the first couple of sections

@@
 \definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\makeatletter
+\@ifundefined{AddToHook}{}{\AddToHook{package/xcolor/after}{\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}}}
+\makeatother
 \newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
@@
 \definecolor{errorcolor}{rgb}{1, 0, 0}
+\makeatletter
+\@ifundefined{AddToHook}{}{\AddToHook{package/xcolor/after}{
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+}}
+\makeatother
 \newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

### 2022-01-22 JDSSV_manuscript_knitr.Rnw
Fixing up time series, references and guessing on FT screenshots

@@
 
-A best practice for effective communication is to present information in an understandable way (Gigerenzer et al., 2007; Gigerenzer and Edwards, 2011), for example by saying "one in ten" instead of 10\%.
-Using absolute rather than relative numbers, presenting information in an appealing graphic form and summarising the most important facts in "fact boxes" are some of the methods that have been developed and advocated to increase transparency in communication between stakeholders such as healthcare providers and patients (see for example: https:// www.hardingcenter.de).
+A best practice for effective communication is to present information in an understandable way \citep{gigerenzerHelpingDoctorsPatients2007,gigerenzerWhatAreNatural2011}, for example by saying "one in ten" instead of 10\%.
+Using absolute rather than relative numbers, presenting information in an appealing graphic form and summarising the most important facts in "fact boxes" are some of the methods that have been developed and advocated to increase transparency in communication between stakeholders such as healthcare providers and patients (see for example: \url{https:// www.hardingcenter.de}).
 Fact boxes in combination with icon arrays are recommended for the presentation of test results.
-Both representations are based on natural frequencies \citep{[62],Krauss2020} and represent case numbers as simply and concretely as possible.
+Both representations are based on natural frequencies \citep{gigerenzerWhatAreNatural2011,Krauss2020} and represent case numbers as simply and concretely as possible.
 Many scientific studies show that icon arrays help people to understand numbers and risks more easily (e.g.
@@
 
+As people grappled with the scale of the pandemic and the lives lost as a result, a different set of data displays attempted to provide context to this loss over time.
+In the New York times, this took the form of a story showing the accumulation of the first 100,000 deaths due to covid, with occasional short quotes from individuals' obituaries to humanize the names\citep{barryRemembering1000002020}.
+While this is not a typical time-series chart, as the user scrolled through the page the pace of the names increases, showing the scale of the pandemic in a visceral way.
+
+A slightly different approach by the BBC's Visual and Data Journalism team, shown in \Cref{fig:bbc-flower}, displayed the Covid cases and deaths as a growing flower, with the stem proportional to the number of cases and the flower petals showing the deaths over time\citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}.
+This visualization is intended to evoke emotion, with the flower representation as an explicitly recognized symbol of grief; it is clearly not about showing specific case counts and deaths numerically.
+The visualization comes with sound as well, an attempt to make the pandemic an auditory experience as well as a visual one.
+
+\begin{figure}
+\centering
+\includegraphics[width=.8\linewidth]{Figures_Web/BBC_flower_time_series}
+\caption{The BBC Visual and Data Journalism Team's Covid flower, showing the cases and deaths over time.
+Source: \url{https://www.bbc.co.uk/news/resources/idt-7464500a-6368-4029-aa41-ab94e0ee09fb}}\label{fig:bbc-flower}
+\end{figure}
 % BBC flowers w/ sonification https://www.bbc.co.uk/news/resources/idt-7464500a-6368-4029-aa41-ab94e0ee09fb
@@
 
-\newpage
-
-\begin{appendix}
-
-\section{More Technical Details} \label{app:technical}
-
-Appendices can be included after the bibliography (with a page break). Each section within the appendix should have a proper section title (rather than just \emph{Appendix}).
+% \newpage
+% 
+% \begin{appendix}
+% 
+% \section{More Technical Details} \label{app:technical}
+% 
+% 
+% Appendices can be included after the bibliography (with a page break). Each section within the appendix should have a proper section title (rather than just \emph{Appendix}).
 
@@
 
-\section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}
-
-References need to be provided in a \textsc{Bib}{\TeX} file (\code{.bib}).
+% 
+% \section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}
+% 
+% References need to be provided in a \textsc{Bib}{\TeX} file (\code{.bib}).
 All references should be made with \verb|\cite|, \verb|\citet|, \verb|\citep|, \verb|\citealp| etc.\ (and never hard-coded).
 This commands yield different formats of author-year citations and allow to include additional details (e.g., pages, chapters, \dots) in brackets.
-%In case you are not familiar with these commands see the JSS style FAQ for details.
-
-Cleaning up \textsc{Bib}{\TeX} files is a somewhat tedious task -- especially when acquiring the entries automatically from mixed online sources.
+% %In case you are not familiar with these commands see the JSS style FAQ for details.
+% 
+% Cleaning up \textsc{Bib}{\TeX} files is a somewhat tedious task -- especially when acquiring the entries automatically from mixed online sources.
 However, it is important that informations are complete and presented in a consistent style to avoid confusions.
 JDSSV requires the following format.
-\begin{itemize}
-  \item Specific markup (\verb|\proglang|, \verb|\pkg|, \verb|\code|) should
-    be used in the references.
-  \item Titles should be inserted in title case.
-  \item Journal titles should not be abbreviated and in title case.
-  \item DOIs should be included where available.
-  \item Software should be properly cited as well. For \proglang{R} packages
-    \code{citation("pkgname")} typically provides a good starting point.
-\end{itemize}
-\end{appendix}
-\newpage
-%% -----------------------------------------------------------------------------
+% \begin{itemize}
+%   \item Specific markup (\verb|\proglang|, \verb|\pkg|, \verb|\code|) should
+%     be used in the references.
+%   \item Titles should be inserted in title case.
+%   \item Journal titles should not be abbreviated and in title case.
+%   \item DOIs should be included where available.
+%   \item Software should be properly cited as well. For \proglang{R} packages
+%     \code{citation("pkgname")} typically provides a good starting point.
+% \end{itemize}
+% \end{appendix}
+% \newpage
+% %% -----------------------------------------------------------------------------
 

### 2022-01-22 JDSSV_manuscript_knitr.Rnw

@@
 \section{Summary and discussion} \label{sec:summary}
+Throughout the pandemic (thus far), visual representations of data have been an integral part of scientific communication.
+While not always optimal, these graphics have attempted to provide meaningful context, to encourage individual and collective action, and to help individuals grapple with the scale of the pandemic in cases, deaths, vaccinations, and interventions.
+As in any developing situation, choices made at some points in the pandemic did not always persist - many news outlets refined their charts and approach to the design of graphics over the course of the pandemic.
+
+%Broad conclusions: 
+% importance of scientific communication
+% graphs are weaponized for misinformation as well as information?
+% Need to improve broad guidelines for communication, underpinning with studies on perception and graphics that
 

### 2022-01-24 JDSSV_manuscript_knitr.Rnw
updated section titles; shortened intro

@@
 Using absolute rather than relative numbers, presenting information in an appealing graphic form and summarising the most important facts in "fact boxes" are some of the methods that have been developed and advocated to increase transparency in communication between stakeholders such as healthcare providers and patients (see for example: \url{https:// www.hardingcenter.de}).
-Fact boxes in combination with icon arrays are recommended for the presentation of test results.
-Both representations are based on natural frequencies \citep{gigerenzerWhatAreNatural2011,Krauss2020} and represent case numbers as simply and concretely as possible.
-Many scientific studies show that icon arrays help people to understand numbers and risks more easily (e.g.
-\cite{McDowell2019}).
-The Harding Center for Risk Literacy shows many other examples of transparent communication of risks, including COVID-19\footnote{\url{https://www.hardingcenter.de/de/mrna-schutzimpfung-gegen-covid-19-fuer-aeltere-menschen}}.
-
+Such "fact boxes" combined with the dynamic layer of the internet allowed for illustrative simulations of the spread of the epidemic, for example as presented already on 14 March 2020 in the Washington Post\footnote{\url{https://www.washingtonpost.com/graphics/2020/world/corona-simulator/}} with the title "Why outbreaks like coronavirus spread exponentially, and how to flatten the curve".
 Nevertheless, both the complexity of phenomena and the "bipolarity" of statistical thinking remain a challenge.
@@
 
-\section{The global perspective}
+\section{The global narrative}
 
@@
 
-Simulations were used particularly illustratively in the media in the course of the pandemic.
-An inspiring example illustrating the spread of the epidemic appeared as early as 14 March 2020 in the Washington Post\footnote{\url{https://www.washingtonpost.com/graphics/2020/world/corona-simulator/}} with the title "Why outbreaks like coronavirus spread exponentially, and how to flatten the curve".
-The Washington Post made this simulation available free of charge and in all major languages, which led to it being distributed worldwide, including repeatedly on German television\footnote{\url{https://web.br.de/interaktiv/corona-simulation/}}.
-The New York Times\footnote{\url{https://www.nytimes.com/interactive/2020/us/coronavirus-spread.html}} published a dynamic graphic entitled "How the Virus Won", which maps the spread of COVID-19 cases from February to June 2020 in the USA.
-It shows how an analysis of the associations between different COVID-19 strains and travel patterns can help understand the spread of the disease.
-
-Another illustrative example is a simulation from ZEIT Online\footnote{\url{https://www.zeit.de/wissen/gesundheit/2020-11/coronavirus-aerosole-ansteckungsgefahr-infektion-hotspot-innenraeume}}, which - based on models developed by a group of researchers at the Max Planck Institute for Chemistry - estimates the probability of an infected person infecting other people in closed rooms in various scenarios.
-While the visualisations present the simulated infection processes in a catchy way, the dependence of the simulations on parameter assumptions and settings is usually not addressed.
-Simulations should also always make transparent on which model assumptions and which data basis the simulations were created.
-
-\section{Time series}
+\section{The temporal narrative}
 % Line plots, frequency domain, NY times tile plots w/ cases represented in colour intensity
@@
 
-\section{Comparisons and rankings}
+\section{The ranking narrative}
 \label{sec:rankings}
@@
 
+Ubiquity of the data, persisting general interest on the topic, the rise of data journalism, and the opportunities provided by interactive and dynamic internet platforms contributed to a an ongoing trend on visual narratives of the Covid-19 pandemic.
+Graphical displays are used as weapons for information, but also for misinformation.
+Despite a growing number of guidelines for communicating statistical results transportation of the intended messages is not always straightforward and remains a challenge.
+The deciphering of visual icons and best practices of visual communication require furhter studies on perception and graphics.
+
 %Broad conclusions: 

### 2022-01-24 JDSSV_manuscript_knitr.tex
updated section titles; shortened intro

@@
 \definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\makeatletter
-\@ifundefined{AddToHook}{}{\AddToHook{package/xcolor/after}{\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}}}
-\makeatother
 \newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
@@
 \definecolor{errorcolor}{rgb}{1, 0, 0}
-\makeatletter
-\@ifundefined{AddToHook}{}{\AddToHook{package/xcolor/after}{
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-}}
-\makeatother
 \newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

### 2022-01-24 JDSSV_manuscript_knitr.Rnw
Updated conclusion and data section

@@
 As in any developing situation, choices made at some points in the pandemic did not always persist - many news outlets refined their charts and approach to the design of graphics over the course of the pandemic.
+This evolution of graphical forms has provided us with an opportunity to evaluate what worked and did not -- and why -- in a context of broad general interest.
 
-Ubiquity of the data, persisting general interest on the topic, the rise of data journalism, and the opportunities provided by interactive and dynamic internet platforms contributed to a an ongoing trend on visual narratives of the Covid-19 pandemic.
+The visual narratives of the pandemic which have persisted over time are due in part to the rise of data journalism and interactive graphics platforms which ensure that every outlet has the ability to host engaging and visually appealing graphics; however, while these graphics are nice to look at, not all are equally functional or useful for communication purposes.
+As in any situation, good tools can be used and mis-used freely; we have seen charts and graphics used to perpetuate misinformation and misleading claims, even though it is far more common for charts and graphs to be used for good - to educate and inform the public.
+
+The pandemic and graphics used to show the ebb and flow of COVID cases highlight the challenges of data visualization in the modern era: even with numerous guidelines and standards for graphics, it is not straightforward to communicate about complex data, and creating charts which fully represent the intended message or messages is a challenge.
+It is essential that we continue to study how graphics are perceived and interpreted, supporting the guidelines and standards which exist with user studies that ground these rules in science.
+
+% Ubiquity of the data, persisting general interest on the topic, the rise of data journalism, and the opportunities provided by interactive and dynamic internet platforms contributed to a an ongoing trend on visual narratives of the Covid-19 pandemic.
 Graphical displays are used as weapons for information, but also for misinformation.
@@
 \section*{Computational Details}
-
-If necessary or useful, information about certain computational details
-such as version numbers, operating systems, or compilers could be included
-in an unnumbered section. Also, auxiliary packages (say, for visualisations,
-maps, tables, \dots) that are not cited in the main text can be credited here.
-
-The results in this paper were obtained using
-\proglang{R}~3.5.1. \proglang{R} itself
-and all packages used are available from the Comprehensive
-\proglang{R} Archive Network (CRAN) at
-\url{https://CRAN.R-project.org/}.
+Graphics included in this paper not sourced from media outlets were created using data published by the New York Times at \url{https://github.com/nytimes/covid-19-data/}.
+Graphics were created using ggplot2\citep{ggplot2} and \proglang{R}~4.1.2.
+\proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.
 

### 2022-01-24 JDSSV_manuscript_knitr.tex
Updated conclusion and data section

@@
 As in any developing situation, choices made at some points in the pandemic did not always persist - many news outlets refined their charts and approach to the design of graphics over the course of the pandemic.
+This evolution of graphical forms has provided us with an opportunity to evaluate what worked and did not -- and why -- in a context of broad general interest.
+
+The visual narratives of the pandemic which have persisted over time are due in part to the rise of data journalism and interactive graphics platforms which ensure that every outlet has the ability to host engaging and visually appealing graphics; however, while these graphics are nice to look at, not all are equally functional or useful for communication purposes.
+As in any situation, good tools can be used and mis-used freely; we have seen charts and graphics used to perpetuate misinformation and misleading claims, even though it is far more common for charts and graphs to be used for good - to educate and inform the public.
+
+The pandemic and graphics used to show the ebb and flow of COVID cases highlight the challenges of data visualization in the modern era: even with numerous guidelines and standards for graphics, it is not straightforward to communicate about complex data, and creating charts which fully represent the intended message is a challenge.
+It is essential that we continue to study how graphics are perceived and interpreted, supporting the guidelines and standards which exist with user studies that ground these rules in science.
+
+% Ubiquity of the data, persisting general interest on the topic, the rise of data journalism, and the opportunities provided by interactive and dynamic internet platforms contributed to a an ongoing trend on visual narratives of the Covid-19 pandemic.
+Graphical displays are used as weapons for information, but also for misinformation.
+Despite a growing number of guidelines for communicating statistical results transportation of the intended messages is not always straightforward and remains a challenge.
+The deciphering of visual icons and best practices of visual communication require furhter studies on perception and graphics.
 
@@
 \section*{Computational Details}
-
-If necessary or useful, information about certain computational details
-such as version numbers, operating systems, or compilers could be included
-in an unnumbered section. Also, auxiliary packages (say, for visualisations,
-maps, tables, \dots) that are not cited in the main text can be credited here.
-
-The results in this paper were obtained using
-\proglang{R}~3.5.1. \proglang{R} itself
+Graphics included in this paper not sourced from media outlets were created using data published by the New York Times at \url{https://github.com/nytimes/covid-19-data/}.
+Graphics were created using ggplot2\citep{ggplot2} and \proglang{R}~4.1.2.
+\proglang{R} itself
 and all packages used are available from the Comprehensive

### 2022-01-24 JDSSV_manuscript_knitr.Rnw
Remove orphaned figures

@@
 Choropleth maps based on raw numbers of cases might look convincing and fit to the purpose, but neglect a number of well-know caveats for statistical reporting and visualisation:
+
 \begin{enumerate}
@@
 The symbol, most commonly a disk or a square, is scaled such that its area corresponds to the data value of the region.
+Clearly, the visual impression of any map based illustration depends on the projection chosen for the underlying map as the examples quite often demonstrate a western-centric viewpoint.
 
@@
 
-And clearly, the visual impression of any map based illustration depends on the projection chosen for the underlying map as the examples quite often demonstrate a western-centric viewpoint.
-
 The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe.
@@
 
-\begin{figure}
-\centering
-\begin{subfigure}[t]{.45\textwidth}
-\includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
-\caption{Hexagonal Cartogram of Covid-19 cases in the United States by state, per million people.
+% \begin{figure}
+% \centering
+% \begin{subfigure}[t]{.45\textwidth}
+% \includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
+% \caption{Hexagonal Cartogram of Covid-19 cases in the United States by state, per million people.
 This somewhat mitigates the area bias issue, but is still subject to other perceptual biases.
 Source: The COVID Tracking Project at The Atlantic \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
-\end{subfigure}\hfill
-\begin{subfigure}[t]{.45\textwidth}
-\includegraphics[width=\textwidth]{nyt-covid-scaled-choro}
-\caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
-\end{subfigure}
-\end{figure}
+% \end{subfigure}\hfill
+% \begin{subfigure}[t]{.45\textwidth}
+% \includegraphics[width=\textwidth]{nyt-covid-scaled-choro}
+% \caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
+% \end{subfigure}
+% \caption{Modifications of choropleth maps which attempt to address the area bias of choropleths in different ways.}
+% \end{figure}
 

### 2022-01-24 JDSSV_manuscript_knitr.tex
Remove orphaned figures

@@
 
-The pandemic and graphics used to show the ebb and flow of COVID cases highlight the challenges of data visualization in the modern era: even with numerous guidelines and standards for graphics, it is not straightforward to communicate about complex data, and creating charts which fully represent the intended message is a challenge.
+The pandemic and graphics used to show the ebb and flow of COVID cases highlight the challenges of data visualization in the modern era: even with numerous guidelines and standards for graphics, it is not straightforward to communicate about complex data, and creating charts which fully represent the intended message or messages is a challenge.
 It is essential that we continue to study how graphics are perceived and interpreted, supporting the guidelines and standards which exist with user studies that ground these rules in science.
@@
 Graphics were created using ggplot2\citep{ggplot2} and \proglang{R}~4.1.2.
-\proglang{R} itself
-and all packages used are available from the Comprehensive
-\proglang{R} Archive Network (CRAN) at
-\url{https://CRAN.R-project.org/}.
+\proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.
 

### 2022-01-24 JDSSV_manuscript_knitr.Rnw
Cleaning up files

@@
 \newcommand{\ma}[1]{\ensuremath{\mathbf{#1}}}
+\graphicspath{{Figures_R/}}
+\graphicspath{{Figures_Web/}}
 
@@
 
-download.file("https://www2.census.gov/programs-surveys/popest/tables/2010-2019/state/totals/nst-est2019-01.xlsx", destfile = "census_pop_state.xlsx", quiet = T, mode = "wb")
-pop_data <- readxl::read_xlsx("census_pop_state.xlsx", col_names = c("state", "pop2019"), col_types = c("text", rep("skip", 11), "numeric"), skip = 9) %>%
+download.file("https://www2.census.gov/programs-surveys/popest/tables/2010-2019/state/totals/nst-est2019-01.xlsx", destfile = "Data/census_pop_state.xlsx", quiet = T, mode = "wb")
+pop_data <- readxl::read_xlsx("Data/census_pop_state.xlsx", col_names = c("state", "pop2019"), col_types = c("text", rep("skip", 11), "numeric"), skip = 9) %>%
   mutate(state = str_remove(state, "[[:punct:]]"))
@@
 The Covid 19 pandemic has amply demonstrated that policies are only effectively implemented and followed by the population if they are accepted by a large majority of the population.
-To achieve this goal, effective communication of quantitative evidence and statistical results between scientists, governments, the media and citizens is essential to justify the appropriateness, usefulness and relevance of the measures.
+To achieve this goal, it is essential that scientists, governments, the media, and citizens all effectively communicate in order to justify the appropriateness, usefulness and relevance of the measures.
+This is as true for guidelines about the use and creation of charts as it is for policies like masks and vaccines.
+% Making it a bit more explicit
 

### 2022-01-24 JDSSV_manuscript_knitr.Rnw
Readme and make code portable and a bit more future-proof
submitted version

@@
 cvstate <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")
+
+if (!dir.exists("Data")) dir.create("Data")
+write_csv(cvstate, file = "Data/covid-data-nyt.csv") # Save data just in case
 
@@
 Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
-\item Color-scheme obstructions: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997}.
+\item Color-scheme obstructions: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997, color2021}.
 The choice of a continuous scale or a categorical scale, the choice of a scale that promotes the recognition of patterns or a scale that supports the filtering out of specific map details, influences the quality of choropleth maps.
@@
 
-All acknowledgments should be collected in this
-unnumbered section before the references. It may contain the usual information
-about funding and feedback from colleagues/reviewers/etc. Furthermore,
-information such as relative contributions of the authors may be added here
-(if any).
+First ideas have been presented at DSSV 2020. We thank our colleagues and the editors for valuable feedback and encouragement to complete this manuscript.
 

### 2022-01-24 JDSSV_manuscript_submitted.Rnw
blinded manuscript included

--- 2022-01-24 JDSSV_manuscript_knitr.Rnw
@@
 %% - \AND starts a new line, \And does not.
-\author{Susan Vanderplas\\University of Nebraska–Lincoln
-   \And Adalbert F.X. Wilhelm\\Jacobs University Bremen}
-\Plainauthor{Susan Vanderplas, Adalbert F.X. Wilhelm}
+\author{
+   \And }
+\Plainauthor{}
 
@@
 \Address{
-    email: \email{susan.vanderplas@unl.edu}
-  Susan Vanderplas\\
-  Department of Statistics\\
-  University of Nebraska–Lincoln\\
-  349A Hardin Hall \\
-  Lincoln, NE 68583-0963, USA\\
-  E-mail: \email{susan.vanderplas@unl.edu}\\
-  URL: \url{https://statistics.unl.edu/susan-vanderplas}\\
-\ \\
-%}
-%
-%\Address{
-  Adalbert F.X. Wilhelm\\
-  Department of Psychology and Methods\\
-  Jacobs University Bremen gGmbH\\
-  Campus Ring 1\\
-  28759 Bremen, Germany\\
-  E-mail: \email{a.wilhelm@jacobs-university.de}\\
-  URL: \url{http://www.jacobs-university.de/directory/wilhelm}
 }

### 2022-01-24 JDSSV_manuscript_submitted.R

@@ -1,0 +1,181 @@
+## ----get-nyt-data, include = F----------------------------------------------------------------------------------------------------------------------------
+
+# Read in the data from NYT
+library(tidyverse)
+library(lubridate)
+library(zoo)
+library(ggrepel)
+library(gridExtra)
+cvstate <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")
+
+if (!dir.exists("Data")) dir.create("Data")
+write_csv(cvstate, file = "Data/covid-data-nyt.csv") # Save data just in case
+
+cvstate <- cvstate %>%
+  group_by(state, fips) %>%
+  mutate(date_100th_case = min(date[cases >= 100]),
+         days_100th_case = date - date_100th_case,
+         daily_cases = cases - lag(cases, 1, default = 0),
+         week_avg_cases = rollmean(cases, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
+         daily_avg_cases = rollmean(daily_cases, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
+         daily_deaths = deaths - lag(deaths, 1, default = 0),
+         week_avg_deaths = rollmean(deaths, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
+         daily_avg_deaths = rollmean(daily_deaths, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T)) %>%
+  ungroup() %>%
+  mutate(state = factor(state))
+
+download.file("https://www2.census.gov/programs-surveys/popest/tables/2010-2019/state/totals/nst-est2019-01.xlsx", destfile = "Data/census_pop_state.xlsx", quiet = T, mode = "wb")
+pop_data <- readxl::read_xlsx("Data/census_pop_state.xlsx", col_names = c("state", "pop2019"), col_types = c("text", rep("skip", 11), "numeric"), skip = 9) %>%
+  mutate(state = str_remove(state, "[[:punct:]]"))
+
+cvstate <- left_join(cvstate, pop_data, by = "state")
+
+my_theme <- theme_bw() + 
+  theme(plot.background = element_rect(fill = "transparent", colour = NA), 
+        panel.background = element_rect(fill = "transparent", colour = NA), 
+        legend.background = element_rect(fill = "transparent", colour = NA),
[…]

### 2022-08-31 Review/Response.qmd
Updates

---
title: "Reviewer Response"
output: html_document
---

# Main Issues

## Reviewer 1

::: callout
### How was the sample of publications selected? Under what criteria?

As it stands, the selection seems arbitrary and very limited. Why the Financial Times and The New York Times and not others? Why not analyzing visualizations from non-Western countries? If the authors want to talk about a “global narrative”, the selection of graphics should also be global. A comparative cross-cultural study would be pretty productive, as it might reveal similarities and differences of graphics from different regions.
:::

The authors are from the United States and Germany; while we participated in many discussions on global social media platforms such as Twitter, we do not have the global perspective (or even facility with sufficient languages) to undertake a review of a global sample of media relating to COVID.

::: callout
### Principles

The authors say that Edward Tufte’s optimization of data-to-ink ratio is a “general principle of visualization”. That’s quite an overstatement. Tufte’s data-to-ink ratio is more of a very subjective, highly flexible guideline that is extremely dependent on factors such as the type of audience, culture, publication, etc.

To begin with, empirical evidence doesn’t support at all that highly stripped-down, high data-to-ink graphics support understanding in any way. As long as the non-data ink in a visualization doesn’t grossly obscure the representation of the data itself, it’s harmless. Moreover, it can even be beneficial, as it might help attract a reader’s attention to the information —in some studies, readers expressed their dislike for the highly sanitized, modernist/minimalist visual style Tufte is an advocate for.

I’d eliminate the reference to this non-principle and I’d stick to the other principle, which is truly advisable and normative; from the paper itself: “ensuring clear understanding by organising the graphics in such a way that the story of the data is told most effectively.”

This would need to be extended, though, and more clearly explained. Making a visualization understandable doesn’t depend just on organization. It also involves making choices about encodings, types of charts and maps and, more importantly (something that I think this paper overlooks) actually testing that the graphic actually conveys the intended information to the audience it’s supposed to inform. Anyone who works in visualization will tell you that simply applying “good design practices” to a visualization will never guarantee for certain that it is understood.

:::

We have modified the principles section to remove the reference to Tufte and the data/ink ratio. We have slightly reorganized the introduction section to add subsections, and have added an additional paragraph which discusses the series of choices made in the process of designing a data visualization. While the general guidelines for making these choices are somewhat outside the scope of this paper, as this is an article, and not a book, and proper treatment of these decisions in a broader context would require at least a book, we hope that highlighting the more general context surrounding the creation of data visualizations and explicitly defining the scope for this article provides some additional clarity.

::: callout
### Graphics Commentary

I believe that the commentary on the graphics the authors chose would need to be more extensive and detailed. Just to give an example, they compare choropleth maps to proportional symbol maps, but they don’t reflect on the many shortcomings of the latter, such as the usual overlap between the scaled symbols. This might be related to space limits in the journal, but I think that this is a paper that deserves to be at least double its length to be truly comprehensive and detailed enough.

The authors also make too many assumptions about how a graphic will be interpreted. This is a direct quote: “The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data”. Where is the evidence for such an assertion? The fact that the authors —likely professionals who are very familiar with statistics and visualization— find a graphic intuitive doesn’t make that graphic intuitive to other audiences. (The authors themselves actually address this in the following section, about log scales, arguably the best one in the paper.)

:::

While the editor has cautioned us not to double the length of the paper, where possible we have attempted to address some obvious shortcomings in our initial discussion without unduly increasing the paper's length.

## Reviewer 2

::: callout
### Phrasing

This is clearly an article by researchers for whom English is likely a second language. While fluent, the wording is often obtuse and clunky. I have made a number of changes in the text to show where it might be improved.
:::

We appreciate the edits.

:::
### Organization

More importantly, there is a mass of data here and a failure of organization. The whole requires a rewrite that distinguishes with one example for each the use of chloropleth, dot maps, tables, annimations, etc. In the chloropleth section and dot map one can distiguish between raw numbers and those that are, for example, cases/per 1000 persons. The use of sliders and annimation would be a separate section.

The goal is simple. A better abstract would say:  Maps and graphs have been extensively used in presenting the evolving state of the Covid-19 pandemic globally, nationally, and locally. This article seeks to review the types of maps and graphs employed by major news media and to assess the degree to which they served as communicative arguments useful both as a basis of public understanding of the pandemic and, at different stages, the rationale behind policies implimented to reduce its impact.

As is, it reads badly and the organization is such that its points get lost. Keep the illustrations to the points you wish to make. Distinguish betwen map types and their uses. Use A, B-heads as you reorder your argument.

And in the discussion ... what have we learned? Well, there are static and dynamic maps, the former most common in news presentations, the latter in web-based. A trivial but useful point. Sliders are a good example how the latter expands what is in the traditional news map. 
:::

The organization of this paper is certainly not the same as that which is proposed, in part because we have taken a narrative view that explores the different communication narratives of visualizations during the pandemic. Because we focus on the narratives and look at the chart types employed in service of those narratives, organization necessarily deviates from a straightforward analysis of each chart type in sequence, but this is by design.

### 2022-08-31 JDSSV_manuscript_submitted.Rnw
Updates

@@
 
-Over the past two years, several waves of Covid-19 infections with different mutants of the SARS-CoV-2 virus have swept across the globe, claiming many lives, causing numerous health damages and affecting our personal lives in many ways.
-According to the WHO, over 304 million confirmed cases and over 5.4 million deaths have been reported as of early January 2022 \footnote{\url{https://www.who.int/publications/m/item/weekly-epidemiological-update-on-covid-19---11-january-2022}}.
+Over the past two years, several waves of Covid-19 infections caused by different mutations of the SARS-CoV-2 virus have swept across the globe, claiming lives, stressing healthcare systems, and changing many different facets of our day-to-day experience.
+According to the WHO, almost 600 million confirmed cases and over 6.45 million deaths have been reported as of late August 2022 \footnote{\url{https://www.who.int/docs/default-source/coronaviruse/who_mou_30_august-2022.pdf}}.
 The pandemic has generated enormous interest in epidemiological data, its analysis and visualisation.
@@
 Media, politicians and individuals use this data to build their narratives about the pandemic, discuss its evolution, justify the measures taken and discuss different prevention strategies against the spread of the virus.
-So there are several goals to be achieved by visualising Covid-19 data.
-These goals and their priorities were adapted as dynamically as the virus mutated and the pandemic changed pace.
-But as with any other data visualisation two general principles remain the same: ensuring clear vision by optimising the data-to-ink ratio \citep{Tufte2001} (or signal-to noise ratio) and ensuring clear understanding by organising the graphics in such a way that the story of the data is told most effectively.
+%So there are several goals to be achieved by visualising Covid-19 data.
+These goals and their priorities were adapted as dynamically as the virus mutated and the pandemic changed pace, but as with any other data visualisation the general principles remain the same: ensuring clear understanding by organising graphics in such a way that the story of the data is told most effectively.
 
-In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic as they have been showcasing their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods.
-Data journalism will certainly be one of the beneficiaries of the Covid-19 pandemic and it has become an innovative part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards.
+\subsection{Data Journalism and the Media}
+
+In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic to showcase their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods.
+COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles but also Data journalism will certainly be one of the beneficiaries of the Covid-19 pandemic and it has become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards.
 
@@
 These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org).
-Numerous publications, initiatives, and ideas to improve the communication of quantitative and statistical information have been prepared, see for example \cite{Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.
+Efforts to improve the communication of quantitative and statistical information date from the beginning mass media and the ability to create charts and tables from data systematically, and while there continue to be efforts made in this area, historical papers outline many of the same issues that we still struggle with today \citep{ASA-standards,haemerPresentationProblemsArea1949,kovermanPLANNINGVISUALPRESENTATION1961,fienbergGraphicalMethodsStatistics1979,Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.
+In other words, while our **technology** for recording, visually summarizing, and communicating information distilled from data has vastly improved over the past 100 years, we still face many of the same problems when **designing** charts which communicate data accurately and effectively.
 
@@
 
-A best practice for effective communication is to present information in an understandable way \citep{gigerenzerHelpingDoctorsPatients2007,gigerenzerWhatAreNatural2011}, for example by saying "one in ten" instead of 10\%.
-Using absolute rather than relative numbers, presenting information in an appealing graphic form and summarising the most important facts in "fact boxes" are some of the methods that have been developed and advocated to increase transparency in communication between stakeholders such as healthcare providers and patients (see for example: \url{https:// www.hardingcenter.de}).
+\subsection{Effective Data Communication}
+A best practice for effective communication (graphical or otherwise) is to present information in an understandable way \citep{gigerenzerHelpingDoctorsPatients2007,gigerenzerWhatAreNatural2011}, for example by saying "one in ten" instead of 10\%.
+Using absolute rather than relative numbers.
+Presenting information in an appealing graphic form and summarising the most important facts in "fact boxes" are some of the methods that have been developed and advocated to increase transparency in communication between stakeholders such as healthcare providers and patients (see for example: \url{https:// www.hardingcenter.de}).
 Such "fact boxes" combined with the dynamic layer of the internet allowed for illustrative simulations of the spread of the epidemic, for example as presented already on 14 March 2020 in the Washington Post\footnote{\url{https://www.washingtonpost.com/graphics/2020/world/corona-simulator/}} with the title "Why outbreaks like coronavirus spread exponentially, and how to flatten the curve".
@@
 
+\subsection{Data Visualization Design and Testing}
+Having briefly discussed the media landscape surrounding the COVID pandemic and goals for communicating about data effectively, we should also briefly discuss the choices that go into creating visual representations of data intended to convey quantitative information.
+While the overall goal of a data visualization is to tell the story of the data clearly and effectively, how this story is conveyed involves a series of design choices: which variables are shown, what limits are imposed, what forms are used to represent the data?
+How will the chart (and any surrounding story, in journalism) be presented to the viewer? 
+As the designer makes these choices, they have in mind an audience, and the audience's assumed characteristics affect the design choices as well: language, education level, likely medium used to access the visualization (computer, physical newspaper, tablet, cell phone) - all of these considerations also impact the design of the chart and what ideas the designer considers during the creative process.
+
+Once the basic form of the chart is determined, there are additional considerations that may be specified by the stylistic conventions of different news organizations, but may also be deliberately chosen for visual effect: color palettes, fonts, spatial layouts, and annotations may be used to highlight information, convey additional points of a narrative, or provide helpful information about how to read an unfamiliar chart.
+While each of these design choices could be (and have been) the subject of  additional paper(s), it is important to acknowledge that the graphics we analyze in this paper did not simply pop into being - they are the product of a series of decisions, some consciously made and others imposed by convention or technological capabilities.
+We will highlight particular choices which were effective (or not) throughout this paper, but we will also leave some undiscussed for the sake of brevity.
+
+Of course, the creative process does not only involve the creator.
+Most graphics that are created for a purpose are tested, either formally or informally, before they are published, even if the test results are never themselves published.
+Because so many of the charts we highlight in this paper were discussed and disseminated via social media, in some cases we can provide contemporary discussions of the design process and reactions to the charts we highlight in this paper.
+Where such examples are not provided, that should in no way be taken as evidence that the design and testing process was less formal; it may only have been less public.
+
+\subsection{Scope}
+Our focus in this paper is to highlight, analyze, and discuss a selection of charts and graphics which caught our attention during the initial, real-time experience of the pandemic.
+As we primarily consume media from western Europe and the United States, our assessment is primarily limited to news outlets catering to these audiences; a retrospective global analysis of COVID graphics would certainly be interesting, but is beyond the scope of this work.
+While the charts we have selected were of personal interest during the fear and uncertainty of the initial stages of the COVID pandemic, in this paper we assess these same charts retrospectively, from a world where COVID is an acknowledged fact of life.
+While it still has significant impacts on our day-to-day lives, the existence of treatments, vaccines, and other preventative measures, as well as the human tendency to adapt to changing circumstances mean that the novel coronavirus which emerged in 2019 no longer dominates our consciousness in the same way.
+Because of this perspective difference, we have enough distance from the initial cataclysm to be able to assess the emotional impact of the graphics at the time and to assess their utility retrospectively.
+
 \section{The global narrative}
@@
 On 11 March 2020, WHO declared the outbreak of the novel coronavirus disease (COVID-19) a pandemic, and since that date at the latest, the global distribution of the disease has been in the public eye.
-The spatial spread of the virus and the resulting cases and deaths are commonly visualized by choropleth maps, see for example \Cref{fig:choro1} showing the total number of infections reported in each country as of January 14, 2022.
-For the pandemic perspective, a central element of the narrative is the ubiquity of the disease and the accompanying global impact.
-Choropleth maps based on raw numbers of cases might look convincing and fit to the purpose, but neglect a number of well-know caveats for statistical reporting and visualisation:
+The spatial spread of the virus and the resulting cases and deaths during this initial period were commonly visualized by choropleth maps, see for example \Cref{fig:choro1} showing the total number of infections reported in each country as of January 14, 2022.
+A central element of the narrative during March 2020 was the ubiquity of the disease and the accompanying global impact.
+While choropleth maps based on raw numbers of cases might look convincing and are an obvious choice under the circumstances, the use of these charts ignores a number of well-know caveats for statistical reporting and visualisation:
 
@@
 \item Absolute value unsuitability: As explained in \citep{monmonier2005, slocum2008, speckmann2010} among others, choropleth maps are fundamentally unsuitable for the representation of absolute numbers.
-Especially in the case of similarly coloured areas of the regions, viewers tend to integrate them unconsciously and perceive choropleths as representations of density.
+Viewers tend to integrate similarly coloured areas of the map unconsciously, and perceive choropleths as representations of density.
 They also do not help to convey the desired message as the absolute numbers of Covid-19 cases are strongly influenced by the population size of the country, but also by the number of tests performed and the accuracy of the recording and reporting system.
@@
 Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
-\item Color-scheme obstructions: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997, color2021}.
+\item Colour-scheme selection: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997, color2021}.
 The choice of a continuous scale or a categorical scale, the choice of a scale that promotes the recognition of patterns or a scale that supports the filtering out of specific map details, influences the quality of choropleth maps.
@@
 
+The issues with choropleth maps were highlighted in trade publications during the initial stages of the pandemic \citep{fieldMappingCoronavirusResponsibly}, but these publications did not penetrate the bureaucracy of public health officials suddenly tasked with data visualisation in realtime at scale.
 Choropleth maps are quite commonly used by media companies and governmental organisations and it is easy to find good and bad examples of their usage.
@@
 The symbol, most commonly a disk or a square, is scaled such that its area corresponds to the data value of the region. 
-Clearly, the visual impression of any map based illustration depends on the projection chosen for the underlying map as the examples quite often demonstrate a western-centric viewpoint.
+Proportional symbol maps sacrifice familiar shapes and topological connectedness for equal visual weight\citep{gaoUsabilityValuebyalphaMaps2019}; while this representation avoids the area bias, proportional symbol maps are subject to under-estimation biases \citep{shimAnalysisMiddleSchool2008} that may be of particular concern when visualizing epidemic disease spread.
+In addition, the visual impression of any map based illustration depends on the projection chosen for the underlying map; the examples in \Cref{fig:choro1} and \Cref{fig:propsymb} have a western-centric bias.
 
@@
 
-The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe.
-None of the above viusalisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them.
-The use of maps in this context more often focuses on the comparative aspects: how are we doing as opposed to our neighbors, which strategy to fight Covid-19 is better?
+Another modification of the standard choropleth to address the area-bias problem is to relax the spatial borders of a region in order to show magnitude using area (as well as color).
+Various algorithms can be used to modify geographic boundaries to maintain spatial continuity, or shapes such as hexagons or squares can be used to represent geographic entities (as in \Cref{fig:hex-cartogram}); with either choice, geometry, topology, or both must be sacrificed.
+While these methods are more popular in the academic visualisation literature, including analysis of the spread of COVID-19\citep{yalcinMappingGlobalSpatiotemporal2022}, they were not overly common in the media for global representations of COVID information.
+This may be because cartograms require more visual effort on the part of the viewer; in addition, news organizations may be reluctant to assume their readers have the geographic sophistication necessary to translate between the distorted map of the globe as drawn on the cartogram and the undistorted mental map of the globe used for mental comparison purposes.
+A spatial dot density map might have allowed for analysis of data across geographic boundaries, but this approach was not commonly used to show global case counts, though it was highlighted as a useful option at the national level \citep{fieldMappingCoronavirusResponsibly}, as shown in \Cref{fig:china-dot-map}.
+
+The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe: none of the above viusalisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them.
+The use of maps to assess the global pandemic more often focuses on the comparative aspects: how are we doing as opposed to our neighbors? Which strategy to fight Covid-19 is better? 
 We will examine these topics more closely in Section~\ref{sec:rankings}.
+
+\begin{figure}
+\centering
+\begin{subfigure}[c]{.45\textwidth}
+\includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
+\caption{Hexagonal Cartogram of Covid-19 cases in the United States by state. The COVID Tracking Project \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
+\end{subfigure}\hfill
+\begin{subfigure}[c]{.45\textwidth}
+\includegraphics[width=\textwidth]{china-dotmap}
+\caption{A dot-density map of COVID cases in China \citep{fieldMappingCoronavirusResponsibly}.}\label{fig:china-dot-map}
+\end{subfigure}
+% \begin{subfigure}[t]{.45\textwidth}
+% \includegraphics[width=\textwidth]{nyt-covid-scaled-choro}
+% \caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
+% \end{subfigure}
+\caption{Modifications of choropleth maps which attempt to address the area bias of choropleths in different ways.
+Both methods are applied at the national scale rather than at global scale because of limitations with the visual form or data quality.}
+\end{figure}
 
@@
 \includegraphics[width=.99\linewidth]{ft-off-charts}
-\caption{Chart showing the increase in cases as a result of the B.1.617.2 variant in regions of the UK.
+\caption{Chart showing the increase in cases as a result of the B.1.617.2 variant in regions of the UK, May 27, 2021.
 Notably, this chart arranges areas to accommodate certain regions whose cases are ``off the chart" in order to emphasize the exceptional situation.
-Source: Tweet by John Burn-Murdoch (@jburnmurdoch), designer for the Financial Times.
-\url{https://twitter.com/jburnmurdoch/status/1397995388267810818}.}
+\citep{burn-murdochNEW1617Fuelling2021}}
 \label{fig:jbm-tweet}
 \end{figure}
+
+Early in the pandemic, as case numbers grew exponentially without appropriate public health measures to restrain them, the scale problem for choropleth maps was particularly complex: over time either the colour mapping changes or the map becomes useless due to compression against the upper limits of the scale, as in \citet{panchadsaramIfWeAre2020}.
+Interestingly, media outlets and municipal dashboards were subject to criticism when the color scales changed in response to increasing case counts\citep{boeckDataVizDecisionmakingAction2020} as well as when the color scale did not change\citep{sandalowChiTribGraphicsPleaseUpdate2020}, illustrating the difficulty of designing visualizations amid competing constraints and a constantly evolving pandemic.
 
@@
 Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time, as there are new developments on a daily basis.
-Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis.
+Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis, as in \Cref{fig-jbm-tweet} and \Cref{fig:wapo-covid-time-series}. 
 This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state. 
@@
 In addition, by hiding these details from the user, it is very difficult to identify when the representation changes over time, as it did several times during the summer of 2020.
-The use of a similar color scheme as well as similar representations made it extremely difficult for users to identify that the value represented had changed.
+The use of a similar color scheme as well as similar geometric representations made it extremely difficult for users to identify that the value represented had changed.
 Clearly, the purpose of these charts is not to provide numerical precision, but rather a comparative assessment of the status of one region relative to others as well as relative to previous time points.
@@
 
-While traditionally, time-series information has been limited to charts with time on the x-axis, there have been many attempts to make use of the capabilities of web graphics and animation in order to convey temporal information in less traditional time-series forms.
-One controversial chart appeared in the New York Times; the x-axis was wrapped around an Archimedean spiral to provide a sense of periodicity in the year-by-year evolution of case counts, as shown in \Cref{fig:nyt-spiral}.
+While traditionally, time-series information has been limited to charts with time on a linear x-axis, many charts manipulated time to highlight different aspects of the pandemic.
+In some cases, these modifications provided large amounts of utility for a relatively small increase in complexity: for example, John Burn-Murdoch's decision to index Financial Times covid graphs to the days since the $N$th case or death.
+This decision is a small modification of the basic time-series plot, but has important consequences in the chart's readability as well as the visual emphasis for users: the graphs in the Financial Times were designed to emphasize "...
+that there's an inevitability about how coronavirus spreads...
+even if there are only a few cases in your country today, based on all the data we have, you will end up going along that path, the same path that the likes of Italy and Spain have been on so tragically."\citep{hannenCoronavirusTrajectoryTracker2020a} By placing all countries on an even footing divorced from the temporal randomness of when the first case appeared, the viewer can effectively compare the effect of government policies and other interventions on the trajectory over time\citep{burn-murdochAllCountriesDeath2020,burn-murdochLatestCaseTrajectories2020}.
+
+The New York Times intentionally violated typical design guidelines in a controversial chart; the x-axis was wrapped around an Archimedean spiral to provide a sense of periodicity in the year-by-year evolution of case counts, as shown in \Cref{fig:nyt-spiral}.
 The original form of this chart is prone to the line-width illusion \citep{vanderplasSignsSineIllusion2015}, in addition to the well-known problems with polar charts \citep{hofmannGraphicalTestsPower2012,waldnerComparisonRadialLinear2020}.
-The re-envisioned version, which aligns counts on the outside of the spiral, and displays them as radial lines, mitigates some of the issues with the line-width illusion by directly showing each line as its own entity, facilitating direct comparisons.
+A re-envisioned version created by Sourya Shrestha, aligns counts on the outside of the spiral, and displays them as radial lines, mitigating some of the issues with the line-width illusion by directly showing each line as its own entity, facilitating direct comparisons.
 In addition, the re-envisioned chart includes a reference line at 100K cases/day, which allows the reader to compare the severity of different waves directly.
 Still, it is more work to interpret and compare peaks on this chart than on a similar linear time series chart with the same data, as the reader must assess line width and then do a mental rotation and shift operation in order to compare with any other time period.
-In addition, this type of chart devotes less area to previous years of the pandemic, which may decrease the amount of focus given to that data; whether this is an advantage or disadvantage depends on your perspective and goals when using the chart.
+By design, this type of chart devotes less area to previous years of the pandemic, which may decrease the amount of focus given to that data; whether this is an advantage or disadvantage depends on your perspective and goals when using the chart.
 
@@
 
-Another attempt to creatively show time-series data that has been extremely common online throughout the pandemic is the bar chart race: an animated series of bar charts over time that show how case counts in each country are increasing.
+Other designers used the capabilities of web graphics to show the time component in the data through animation.
+In some cases, this took the form of animated bubble plots XXX add XXX, while others dispensed with the traditional form of a line graph altogether, showing rising case counts using a "bar chart race": an animated series of bar charts over time that show how case counts in each country are increasing.
 A typical example can be found on YouTube at \url{https://www.youtube.com/watch?v=8WH-bTH6GqI}; a screenshot is shown in \Cref{fig:barchart-race}.
@@
 As people grappled with the scale of the pandemic and the lives lost as a result, a different set of data displays attempted to provide context to this loss over time.
-In the New York times, this took the form of a story showing the accumulation of the first 100,000 deaths due to covid, with occasional short quotes from individuals' obituaries to humanize the names\citep{barryRemembering1000002020}.
+In the New York times, this took the form of a story showing the accumulation of the first 100,000 deaths due to COVID in the US, with occasional short quotes from individuals' obituaries to humanize the names\citep{barryRemembering1000002020}.
 While this is not a typical time-series chart, as the user scrolled through the page the pace of the names increases, showing the scale of the pandemic in a visceral way.
+A less sentimental dot-density timeline by the New York Times about 9 months later chronicled the nation reaching 450,000 deaths; the difference between the two visualizations brings to mind the quote oft attributed to Stalin: "One death is a tragedy; a million deaths is a statistic" \citep{investigatorSingleDeathTragedy2016}.
 

### 2022-09-01 JDSSV_manuscript_submitted.Rnw
Updates

@@
 In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic to showcase their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods.
-COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles but also Data journalism will certainly be one of the beneficiaries of the Covid-19 pandemic and it has become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards.
+COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles but also Data journalism will certainly be one of the beneficiaries of the Covid-19 pandemic and it has become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards\citep{kochWelcomeRevolutionCOVID192021}.
 
@@
 
-At the same time, we still see a lot of defective graphics disseminated and shared: some that violate fundamental visualisation and statistical reporting principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility.
+At the same time, we still see a lot of defective graphics disseminated and shared: some that violate fundamental statistical and visualisation design principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility.
 These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org).
@@
 
-The Covid 19 pandemic has amply demonstrated that policies are only effectively implemented and followed by the population if they are accepted by a large majority of the population.
-To achieve this goal, it is essential that scientists, governments, the media, and citizens all effectively communicate in order to justify the appropriateness, usefulness and relevance of the measures.
-This is as true for guidelines about the use and creation of charts as it is for policies like masks and vaccines.
-% Making it a bit more explicit
-
 \subsection{Effective Data Communication}
-A best practice for effective communication (graphical or otherwise) is to present information in an understandable way \citep{gigerenzerHelpingDoctorsPatients2007,gigerenzerWhatAreNatural2011}, for example by saying "one in ten" instead of 10\%.
-Using absolute rather than relative numbers.
-Presenting information in an appealing graphic form and summarising the most important facts in "fact boxes" are some of the methods that have been developed and advocated to increase transparency in communication between stakeholders such as healthcare providers and patients (see for example: \url{https:// www.hardingcenter.de}).
-Such "fact boxes" combined with the dynamic layer of the internet allowed for illustrative simulations of the spread of the epidemic, for example as presented already on 14 March 2020 in the Washington Post\footnote{\url{https://www.washingtonpost.com/graphics/2020/world/corona-simulator/}} with the title "Why outbreaks like coronavirus spread exponentially, and how to flatten the curve".
+A best practice for effective communication (graphical or otherwise) is to present information in an understandable way \citep{gigerenzerHelpingDoctorsPatients2007,gigerenzerWhatAreNatural2011}, for example by saying "one in ten" instead of 10\%, or using absolute rather than relative numbers.
+Presenting information in a visually appealing graphic form and highlighting the most important points in "fact boxes" is one extremely effective way to engage readers while focusing on key takeaway points.
+This approach was developed to increase effective communication between stakeholders such as healthcare providers and patients (see for example: \url{https://www.hardingcenter.de}, archived in ). 
+Such "fact boxes" combined with the dynamic layer of the internet allowed for illustrative simulations of the spread of the epidemic, for example as presented on 14 March 2020 in the Washington Post\footnote{\url{https://www.washingtonpost.com/graphics/2020/world/corona-simulator/}} with the title "Why outbreaks like coronavirus spread exponentially, and how to flatten the curve".
 Nevertheless, both the complexity of phenomena and the "bipolarity" of statistical thinking remain a challenge.
-While human thinking tends to simplify patterns and political communication also prefers a simple cause-effect relationship, real phenomena are often multivariate. 
-Thus, in studying COVID-19 and predicting its spread, it is not only important to consider symptomatology, disease incidence and geographic distribution, population behaviour patterns, government policies, and impacts on the economy, schools, people in nursing homes, and society at large, but also to incorporate these into data analyses and communication of results.
+While human thinking tends to simplify patterns, and political communication also prefers a simple cause-effect relationship, real phenomena are often multivariate. 
+Thus, in studying COVID-19 and predicting its spread, it is important not only to consider symptomatology, disease incidence and geographic distribution, population behaviour patterns, government policies, and impacts on the economy, schools, people in nursing homes, and society at large, but also to incorporate these into data analyses and communication of results.
 Associations observed in the data can often be caused by third party variables (confounders). 
@@
 A central element of the narrative during March 2020 was the ubiquity of the disease and the accompanying global impact.
+
+Choropleth maps are quite commonly used by media companies and governmental organisations and it is easy to find good and bad examples of their usage. 
 While choropleth maps based on raw numbers of cases might look convincing and are an obvious choice under the circumstances, the use of these charts ignores a number of well-know caveats for statistical reporting and visualisation:
@@
 Viewers tend to integrate similarly coloured areas of the map unconsciously, and perceive choropleths as representations of density.
-They also do not help to convey the desired message as the absolute numbers of Covid-19 cases are strongly influenced by the population size of the country, but also by the number of tests performed and the accuracy of the recording and reporting system.
+They also do not help to convey the desired message as the absolute numbers of Covid-19 cases are strongly influenced by the population size of the country as well as by the number of tests performed and the accuracy of the recording and reporting system.
 \item The area-bias: The visual impression is determined more by the colour and the geographical area of the individual countries than by the number of Covid cases.
@@
 
-The issues with choropleth maps were highlighted in trade publications during the initial stages of the pandemic \citep{fieldMappingCoronavirusResponsibly}, but these publications did not penetrate the bureaucracy of public health officials suddenly tasked with data visualisation in realtime at scale.
-Choropleth maps are quite commonly used by media companies and governmental organisations and it is easy to find good and bad examples of their usage.
-Proportional symbol maps, see Fig.~\ref{fig:propsymb} or graduated symbol maps place scaled symbols or diagrams directly on the input map, often on the centroid of the regions. 
+The issues with choropleth maps were highlighted in trade publications during the initial stages of the pandemic \citep{fieldMappingCoronavirusResponsibly}, but these publications did not penetrate the bureaucracy of public health officials suddenly tasked with data visualisation in real time at scale.
+
+Proportional symbol maps, see \Cref{fig:propsymb} or graduated symbol maps place scaled symbols or diagrams directly on the input map, often on the centroid of the regions. 
 The symbol, most commonly a disk or a square, is scaled such that its area corresponds to the data value of the region. 
@@
 Throughout the pandemic, both in choropleth maps and time-series graphs, creators have had to grapple with scale.
-When scales automatically adjust, mental comparisons across time can become misleading, because e.g.
-the colour scale from yesterday does not mean the same thing as it did today, as the range has expanded and the colours have stayed the same.
-This problem is more noticeable in choropleths than in time-series charts, where we are more likely to expect and notice a scale change, but it is still a fundamental issue in many different types of charts.
-Some publications leaned into this by arranging their graphics to highlight the fact that cases were ``off the charts", as in \Cref{fig:jbm-tweet}, implicitly cuing readers that the situation was exceptional and worthy of their attention.
+When scales automatically adjust, mental comparisons across time can become misleading, because the colour scale from yesterday does not mean the same thing as it did today, as the range has expanded and the colours have stayed the same.
+It is not that often that day-to-day design changes (or lack thereof) cause fundamental problems with the interpretation of data visualisation, but during the COVID 19 pandemic this situation was both extremely common and had real consequences: people would glance at the current COVID risk map and assume that "red" meant the same thing that it did yesterday, despite the fact that the underlying values had changed\citep{abiadColorDiagramWHOLebanon2020,rubelHeyUpshotNYTCurious2020,matthewsLessonFutzingData2020}.
+One example of this issue over two months of NY Times graphics is shown in \Cref{fig:nyt-colors}; the overall color scheme used changes once, but the variable displayed changes (at least) 4 times over the course of a two month period.
+
+\begin{figure}
+\centering
+\begin{subfigure}[c]{.25\textwidth}
+\includegraphics[width=\textwidth]{nyt-wayback-20200616}
+\caption{June 16, 2020}\label{fig:nyt-colors1}
+\end{subfigure}\hfill
+\begin{subfigure}[c]{.25\textwidth}
+\includegraphics[width=\textwidth]{nyt-wayback-20200716}
+\caption{July 16, 2020}\label{fig:nyt-colors2}
+\end{subfigure}\hfill
+\begin{subfigure}[c]{.25\textwidth}
+\includegraphics[width=\textwidth]{nyt-wayback-20200718}
+\caption{July 18, 2020}\label{fig:nyt-colors3}
+\end{subfigure}\hfill
+\begin{subfigure}[c]{.25\textwidth}
+\includegraphics[width=\textwidth]{nyt-wayback-20200814}
+\caption{August 1, 2020}\label{fig:nyt-colors4}
+\end{subfigure}\hfill
+\caption{A series of screenshots of the NY Times state-by-state display showing evolving color schemes and even metrics; initially, a discrete scale of two-week change in cases is shown; then, a similar color scheme is used with a different metric.
+Around July 17, the color scheme and variable shown changed to weekly cases per capita (from "fewer" to "more"), and by August 14, the variable shown changed back again ("falling" to "rising") while the color scheme stayed the same.}\label{fig:nyt-colors}
+\end{figure}
+
+This problem is more noticeable in choropleths and other charts which show case counts based on color scale alone than in time-series charts, where we are more likely to expect and notice a scale change, but it is still a fundamental issue across chart types and of particular concern when there is no natural upper bound on the variable displayed.
+Some designers leaned into this by arranging their graphics to highlight the fact that cases were ``off the charts", as in \Cref{fig:jbm-tweet}, implicitly cuing readers that the situation was exceptional and worthy of their attention.
 
@@
 Early in the pandemic, as case numbers grew exponentially without appropriate public health measures to restrain them, the scale problem for choropleth maps was particularly complex: over time either the colour mapping changes or the map becomes useless due to compression against the upper limits of the scale, as in \citet{panchadsaramIfWeAre2020}.
+The magnitude of the problem also depends on whether the scale uses a set of bins (a discrete mapping to a continuous variable) or uses a continuous color mapping; in some cases, outlets switched between the two options over time or changed the underlying unit entirely.
 Interestingly, media outlets and municipal dashboards were subject to criticism when the color scales changed in response to increasing case counts\citep{boeckDataVizDecisionmakingAction2020} as well as when the color scale did not change\citep{sandalowChiTribGraphicsPleaseUpdate2020}, illustrating the difficulty of designing visualizations amid competing constraints and a constantly evolving pandemic.
@@
 The visual narratives of the pandemic which have persisted over time are due in part to the rise of data journalism and interactive graphics platforms which ensure that every outlet has the ability to host engaging and visually appealing graphics; however, while these graphics are nice to look at, not all are equally functional or useful for communication purposes.
-As in any situation, good tools can be used and mis-used freely; we have seen charts and graphics used to perpetuate misinformation and misleading claims, even though it is far more common for charts and graphs to be used for good - to educate and inform the public.
+As in any situation, good tools can be used and misused freely; we have seen charts and graphics used to perpetuate misinformation and misleading claims, even though it is far more common for charts and graphs to be used for good - to educate and inform the public.
+
+The Covid 19 pandemic has amply demonstrated that policies are only effectively implemented and followed by the population if they are accepted by a large majority of the population.
+To achieve this goal, it is essential that scientists, governments, the media, and citizens all effectively communicate in order to justify the appropriateness, usefulness and relevance of the measures.
+This is as true for guidelines about the use and creation of charts as it is for policies like masks and vaccines.
+% Making it a bit more explicit
 
@@
 %% - With proper section titles and _not_ just "Appendix".
-
+% 
 % \newpage

### 2022-09-02 Review/Response.qmd
Updated comments and edits

@@
 
-# Main Issues
+# AE
+
+::: callout
+## Overall Evaluation
+Overall largely
+This paper provides an interesting overview of graphics used to portray data on the COVID 19 pandemic, largely from popular media.
+It attempts to use ideas and principles of graphics design and human factors
+studies of graphical perception to evaluate the effectiveness of these displays for the public and policy
+makers.
+
+All in all, I think it largely succeeds in these aims and would be a suitable and interesting article for JDSSV readers.
+However, the two reviewers make substantial suggestions and comments which need to be considered before it is ready for publication.
+I recommend acceptance, following suitable revision (together with a reply from the authors to these reviews).
+:::
+
+Specific reviewer comments are addressed below, here, we address additional discussions brought up by the AE in response to the reviewers' comments.
+
+::: callout
+## Reviewer 2 - Principles
+
+To the reviewer’s comment, I add that there needs to some clear statement regarding the match
+between the design of a graphic and the capabilities of the intended audience in determining whether a
+graphic display is successful for its intended purpose. This is more important than simple design rules.
+:::
+
+We have added a short paragraph addressing this in the data communication subsection of the introduction, and used the opportunity to state that we have primarily focused on charts intended for mass consumption rather than experts in public health or epidemiology.
+
+::: callout
+## Other comments: Proportional symbol maps
+
+I was very surprised to see no mention or illustration of anamorphic maps or cartograms, where the area of each geographical unit is transformed to represent the count (cases/deaths).
+I strongly suggest including something on this.
+:::
+
+We have added a cartogram to the discussion of proportional symbol maps and choropleths. Thank you for catching an obvious omission on our part.
+
+::: callout
+### Other comments: Color scheme obstructions
+
+Color scheme is mentioned only in regard to Fig 6, but there are other examples where the color scheme used could be noted, either positively or negatively.
+
+:::
+
+We have added additional analysis of color where we feel it is useful; Figure 12 is one case where we chose not to mention color in part because I (Susan) can't tell if it's effective or not - I'm colorblind, and while I have trouble with streamgraphs in general, that is particularly true if color is actually meaningful.
+
+::: callout
+### Other comments: Uncertainty
+
+Uncertainty: In the list of aspects mentioned in the middle of P 2, I am struck by the absence of mention
+of uncertainty in visual representations of COVID. Clearly, the data are complex as they are, but
+understanding their uncertainty is also a topic that could be mentioned or discussed.
+
+:::
+
+We wholeheartedly agree - uncertainty should be mentioned in the introduction, though I can't think of any specific examples of addressing the uncertainty through visualization (as opposed to footnotes about data quality) in mainstream graphics which were circulated at the time.
+Still, that is a rather glaring omission from both our discussion and the collection of graphs we've accumulated as interesting.
+
+::: callout
+### Other comments: Temporal narrative
+
+In Section 3, The temporal narrative, there could usefully be some discussion, and perhaps an illustration
+of the use of dynamic charts like the moving bubble chart, to liberate the X axis from time, using
+animation or an interactive slider.
+
+:::
+
+We've added a moving bubble chart - again, this is an obvious omission.
+Thank you for bringing it to our attention.
+Interestingly, it was somewhat difficult to find a chart like this in a mainstream media outlet when searching through retrospectively.
+Part of that may be the fact that the web isn't great at keeping archival content, and many of these dashboards have been overwritten through successive modification.
+I suspect another part, however, is that moving bubble charts require a lot of effort to read and understand, and possibly were more common on social media and in the data vis community writ large than in media intended for the general public.
+
+::: callout
+### Other comments: standardization
+
+More prominence could be given to the thoughtful standardization that went into some of these
+displays to provide comparability of representation across global and local geographies and over
+multiple waves of the pandemic. For example, in the early days cases and deaths were tied to calendar
+dates giving time series charts with different origins. The simple device (I believe from Burn-Murdoch of
+the FT) of making these all start when N cases had been observed was a game changer. There are
+probably other examples.
+
+:::
+
+There may be other examples, but this is a good one.
+We've added a comment to that effect in the discussion of time series data, but we also mention standardization w.r.t.
+area and population in the choropleth/cartogram section.
+While the topic probably deserves a more thorough treatment, it's not easy to slide it in and maintain the current structure of the paper, so we hope that this suffices.
 
@@
 The authors are from the United States and Germany; while we participated in many discussions on global social media platforms such as Twitter, we do not have the global perspective (or even facility with sufficient languages) to undertake a review of a global sample of media relating to COVID.
+We have added a section which explicitly states how charts were selected and what our individual biases are with respect to geography and culture.
 
@@
 While the editor has cautioned us not to double the length of the paper, where possible we have attempted to address some obvious shortcomings in our initial discussion without unduly increasing the paper's length.
+This paper is in part a result of discussing COVID graphics in several groups and seminars with data visualization professionals and also with laypeople; as a result, while we cannot directly cite a source saying that a specific graphic is simple and intuitive, we can directly assess the visual complexity of the chart and we can distill down the series of discussions we've had about these topics over the course of the pandemic to arrive at what we hope are fairly useful conclusions.
+In an attempt to be a bit more precise, however, where we could find relevant discussions on social media, we have included citations of those discussions to provide some support for claims that we have made summarizing the views of others.
 
@@
 We appreciate the edits.
+We have addressed many of the stylistic suggestions where appropriate.
+In other cases, we have modified the paper in ways that rendered the suggested edits moot, and we hope the reviewer will appreciate the final result in any case.
 
-:::
+::: callout
 ### Organization
@@
 Sliders are a good example how the latter expands what is in the traditional news map.
+
 :::

### 2022-09-02 JDSSV_manuscript_submitted.Rnw
tweak figure and add pdf
Updated comments and edits

@@
 
+In other areas, such as visualization of uncertainty, the pandemic presents new challenges for scientific communication.
+How do we effectively show the uncertainty in the data amid flawed tests, new strains, changing diagnostic criteria, and incredible variability in global public health infrastructure?
+Moreover, how do we do this without further decreasing public trust in institutions responsible for shepharding us through the pandemic?
+While most designers chose to side-step these questions in the moment, as we assess the pandemic in graphical form, we should also consider what was omitted from our visual renderings.
+
 \subsection{Effective Data Communication}
@@
 However, statisticians who point out these limitations are at risk of having their statements pulled out one-sidedly in a polarised debate \citep{McConway2021}.
+
+Finally, it is critical that data communication take into account the intended audience.
+Graphics created to communicate with public health officials should be different from those created for the purpose of informing the public.
+Designers must carefully consider the visual capabilities, attention constraints, and mathematical sophistication of the audience when creating a chart, but in this era of instantaneous global communication, charts designed for specific audiences may still reach a large population of viewers with different capabilities.
+In this paper, we focus primarily on graphics which are intended for mass consumption, if only to reduce the scope to something which can be (non-exhaustively) covered in a paper.
 
@@
 It is not that often that day-to-day design changes (or lack thereof) cause fundamental problems with the interpretation of data visualisation, but during the COVID 19 pandemic this situation was both extremely common and had real consequences: people would glance at the current COVID risk map and assume that "red" meant the same thing that it did yesterday, despite the fact that the underlying values had changed\citep{abiadColorDiagramWHOLebanon2020,rubelHeyUpshotNYTCurious2020,matthewsLessonFutzingData2020}.
-One example of this issue over two months of NY Times graphics is shown in \Cref{fig:nyt-colors}; the overall color scheme used changes once, but the variable displayed changes (at least) 4 times over the course of a two month period.
-
-\begin{figure}
-\centering
-\begin{subfigure}[c]{.25\textwidth}
-\includegraphics[width=\textwidth]{nyt-wayback-20200616}
-\caption{June 16, 2020}\label{fig:nyt-colors1}
-\end{subfigure}\hfill
-\begin{subfigure}[c]{.25\textwidth}
-\includegraphics[width=\textwidth]{nyt-wayback-20200716}
-\caption{July 16, 2020}\label{fig:nyt-colors2}
-\end{subfigure}\hfill
-\begin{subfigure}[c]{.25\textwidth}
-\includegraphics[width=\textwidth]{nyt-wayback-20200718}
-\caption{July 18, 2020}\label{fig:nyt-colors3}
-\end{subfigure}\hfill
-\begin{subfigure}[c]{.25\textwidth}
-\includegraphics[width=\textwidth]{nyt-wayback-20200814}
-\caption{August 1, 2020}\label{fig:nyt-colors4}
-\end{subfigure}\hfill
-\caption{A series of screenshots of the NY Times state-by-state display showing evolving color schemes and even metrics; initially, a discrete scale of two-week change in cases is shown; then, a similar color scheme is used with a different metric.
-Around July 17, the color scheme and variable shown changed to weekly cases per capita (from "fewer" to "more"), and by August 14, the variable shown changed back again ("falling" to "rising") while the color scheme stayed the same.}\label{fig:nyt-colors}
-\end{figure}
+One example of this issue over two months of NY Times graphics is shown in \Cref{fig:sparklines-heatmap-nyt}; the overall color scheme used changes once, but the variable displayed changes (at least) 4 times over the course of a two month period.
 
@@
 We know that slope judgments are much less accurate than position \citep{clevelandGraphicalPerceptionVisual1987}, even in relatively simple situations; when we add in our ability to assess exponential growth, our perceptual accuracy is even more suspect.
+This issue will be discussed further in \Cref{sec:logs} when we discuss log scales, which are one solution to our misperception of exponential growth.
 
-Other time-series representations were more abstract, sacrificing exact data representations for a higher-level summary overview that allowed viewers to make comparisons between regions and points in time without the burden of numerical calculations.
+Yet another approach to showing the temporal component of the pandemic was to leverage web graphics to show the temporal component through animation; this freed up the $x$ axis which in the default configuration would have been devoted to time and allowed another variable to be displayed.
+While this approach was common on social media, it was reserved for specific use-cases and more carefully selected variables by professional media outlets.
+\Cref{fig:bubble-ft} shows an example from the Financial Times which uses the x-axis to show GDP per capita and the y-axis to show vaccination rate, bubble size indicates a country's population.
+With so many variables and ratios of variables, these charts are complex to read and use, even though they are visually appealing.
+
+\begin{figure}
+\centering
+\includegraphics{ft_vaccine_bubble}
+\caption{Animated bubble chart showing vaccination rate relative to per capita GDP for each country, from the Financial Times (Source: \url{https://ig.ft.com/coronavirus-vaccine-tracker})}\label{fig:bubble-ft}
+\end{figure}
+
+Other time-series representations were more abstract, sacrificing exact data representations for a higher-level summary comparing regions and points in time without the burden of numerical calculations.
 Several versions from the New York Times are shown in \Cref{fig:sparklines-heatmap-nyt}.
-The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data, however, they do not present precise numerical information and even the color scale is fairly opaque -- it would be extremely difficult to translate the visualization into any sort of accurate numerical estimate.
-In addition, by hiding these details from the user, it is very difficult to identify when the representation changes over time, as it did several times during the summer of 2020.
+The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data, however, they do not present precise numerical information and even the color scale is (perhaps intentionally) opaque -- it would be extremely difficult to translate the visualization into any sort of accurate numerical estimate.
+In addition, by hiding these details from the user, it is very difficult to identify when the underlying mathematical representation changes over time, as it did several times during the summer of 2020.
 The use of a similar color scheme as well as similar geometric representations made it extremely difficult for users to identify that the value represented had changed.
@@
 
+% 
+% \begin{figure}
+% \centering
+% \includegraphics[width=.3\linewidth]{NYT-heatmap-crop}\hfill
+% \includegraphics[width=.3\linewidth]{NYT-heatmap-crop3}\hfill
+% \includegraphics[width=.3\linewidth]{NYT-heatmap-crop2}
+% \caption{Sparklines style heatmaps embedded in a table of coronavirus data by state.
+Over the course of Summer 2020, the specific measures shown changed, even though the form of the data remained highly similar.
+Source: \url{https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html} (May 6, 2020; June 15, 2020; and July 24, 2020, respectively) }
+% 
+% \label{fig:sparklines-heatmap-nyt}
+% \end{figure}
+
 \begin{figure}
 \centering
-\includegraphics[width=.3\linewidth]{NYT-heatmap-crop}\hfill
-\includegraphics[width=.3\linewidth]{NYT-heatmap-crop3}\hfill
-\includegraphics[width=.3\linewidth]{NYT-heatmap-crop2}
-\caption{Sparklines style heatmaps embedded in a table of coronavirus data by state.
-Over the course of Summer 2020, the specific measures shown changed, even though the form of the data remained highly similar.
-Source: \url{https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html} (May 6, 2020; June 15, 2020; and July 24, 2020, respectively) }
-
-\label{fig:sparklines-heatmap-nyt}
+\begin{subfigure}[t]{.25\textwidth}
+\includegraphics[width=\textwidth]{nyt-wayback-20200616}
+\caption{June 16, 2020}\label{fig:nyt-colors1}
+\end{subfigure}\hfill
+\begin{subfigure}[t]{.25\textwidth}
+\includegraphics[width=\textwidth]{nyt-wayback-20200716}
+\caption{July 16, 2020}\label{fig:nyt-colors2}
+\end{subfigure}\hfill
+\begin{subfigure}[t]{.25\textwidth}
+\includegraphics[width=\textwidth]{nyt-wayback-20200718}
+\caption{July 18, 2020}\label{fig:nyt-colors3}
+\end{subfigure}\hfill
+\begin{subfigure}[t]{.25\textwidth}
+\includegraphics[width=\textwidth]{nyt-wayback-20200814}
+\caption{August 1, 2020}\label{fig:nyt-colors4}
+\end{subfigure}\hfill
+\caption{A series of screenshots of the NY Times state-by-state display showing evolving color schemes and even metrics; initially, a discrete scale of two-week change in cases is shown; then, a similar color scheme is used with a different metric.
+Around July 17, the color scheme and variable shown changed to weekly cases per capita (from "fewer" to "more"), and by August 14, the variable shown changed back again ("falling" to "rising") while the color scheme stayed the same.}\label{fig:sparklines-heatmap-nyt}
 \end{figure}
@@
 This chart provides context for the Omicron wave, showing that while infections are occurring at a much faster rate than in previous waves, hospital admissions seem to be approximately following previous trends.
+In addition, it is clear that the chart is highlighting the Omicron variant from the colour selection: the red used to show the Omicron numbers is a sharp contrast from the green and blue shades used to show the first three waves of the pandemic.
+Notably, even though red and green in combination are usually to be avoided for the sake of colourblind viewers, the shades of blue/green variants which are chosen are light enough that the red stands out even to someone with colour vision deficiency.
 
@@
 \caption{Covid Cases, positivity rate, and hospital admissions from South Africa's Gauteng province, on a log scale (top) and linear scale (bottom).
-Successive waves are overlaid, showing that the increase in cases due to the Omicron variant was sharper than any previously measured increase, while hospitalizations were much more comparable to previous waves.
-Source: John Burn-Murdoch, \url{https://twitter.com/jburnmurdoch/status/1466480113487392769}}
+Successive waves are overlaid, showing that the increase in cases due to the Omicron variant was sharper than any previously measured increase, while hospitalizations were much more comparable to previous waves \citep{burn-murdoch_new_2021-1}.}
 \label{fig:ft-waves}
@@
 A re-envisioned version created by Sourya Shrestha, aligns counts on the outside of the spiral, and displays them as radial lines, mitigating some of the issues with the line-width illusion by directly showing each line as its own entity, facilitating direct comparisons.
+
 In addition, the re-envisioned chart includes a reference line at 100K cases/day, which allows the reader to compare the severity of different waves directly.
 Still, it is more work to interpret and compare peaks on this chart than on a similar linear time series chart with the same data, as the reader must assess line width and then do a mental rotation and shift operation in order to compare with any other time period.
-By design, this type of chart devotes less area to previous years of the pandemic, which may decrease the amount of focus given to that data; whether this is an advantage or disadvantage depends on your perspective and goals when using the chart.
+By design, this type of chart devotes less area to previous years of the pandemic, which may decrease the amount of focus given to that data.%; whether this is an advantage or disadvantage depends on your perspective and goals when using the chart.
 
@@
 A slightly different approach by the BBC's Visual and Data Journalism team, shown in \Cref{fig:bbc-flower}, displayed the Covid cases and deaths as a growing flower, with the stem proportional to the number of cases and the flower petals showing the deaths over time\citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}.
-This visualization is intended to evoke emotion, with the flower representation as an explicitly recognized symbol of grief; it is clearly not about showing specific case counts and deaths numerically.
 The visualization comes with sound as well, an attempt to make the pandemic an auditory experience as well as a visual one.
+The form, animation, and auditory experience are all indicative of an emotional appeal \citep{kostelnick_re-emergence_2016,dignazio_data_2020}; with the flower representation as an explicitly recognized symbol of grief; it is clearly not about showing specific case counts and deaths numerically.
 
@@
 The plague is always the others: This is the short formula for dealing with infectious diseases from a historical perspective \citep{thiessen2021}.
-Sociologists have coined the term ``othering'' for such attributions of the others \citep{mountz2009}.
-This refers to the observation that above all new, unknown threats are projected onto ``strangers'' and ``the others''.
 Closing the borders and restricting access to the country became a popular means of controlling the spread of the disease at several different points during the pandemic.
@@
 
-The bar charts used for this purpose (see top image in Fig.~\ref{fig:ft-death-ranking}) are straightforward and provide a good way to show the ``leading'' countries or regions.
+The bar charts used for this purpose (see top image in \Cref{fig:ft-death-ranking}) are straightforward and provide a good way to show the ``leading'' countries or regions.
 Emphasis on specific countries is either preconfigured by the creator of the chart or can be modified interactively in the online version by check boxes and drop-down menus.
 These bar charts also allow easy adjustment of the count data to population size or other meaningful standard.
-Streamgraphs (see lower image in Fig.~\ref{fig:ft-death-ranking}) are a special type of stacked area graphs and became quite popular around 2008 when they were used by the New York Times to visualise box office results of movies \citep{streamgraph}.
-Their usefulness depends heavily on the clarity of the pattern.
-Various design choices such as the ordering of the different groups have a strong impact on the visual impression and message conveyed by the graph.
+Streamgraphs (see lower image in \Cref{fig:ft-death-ranking}) are a special type of stacked area graphs and became quite popular around 2008 when they were used by the New York Times to visualise box office results of movies \citep{streamgraph}.
+Their usefulness depends heavily on the clarity of the pattern, as well as design choices such as the ordering of the different groups.
 Small variations in the proportions of the groups are almost impossible to detect, but streamgraphs give a good indication of which group is predominant at any given time, even though they are subject to the line-width or sine illusion \citep{vanderplasSignsSineIllusion2015}.
+When paired with the bar charts showing excess mortality, the streamgraph provides temporal context to the instantaneous information in the bar graph.
 
@@
 
-\section{To log or not to log}
+\section{To log or not to log}\label{sec:logs}
 As COVID cases grow quasi-exponentially while there are susceptible members of the population (subject to the effectiveness of mitigation measures and testing availability), it seems natural to use log scales to allow for more effective comparisons of slight changes in case counts over time.
@@
 
-One issue with assessing the use of log scales is that their effectiveness changes with the stage of the pandemic and the amount (and varieties) of data shown.
+One issue with assessing the use of log scales is that their effectiveness changes with the stage of the pandemic and the amount and variety of data shown.
 Initially, log scales were incredibly useful at showing case counts, because minimal mitigation measures were in place and the growth of case counts (or presumptive positive cases, in absence of available testing) was fairly close to exponential.

### 2022-09-02 Review/Edits/JDSSV_manuscript_submitted.tex
Edits and original submission

--- 2022-09-02 JDSSV_manuscript_submitted.Rnw
@@
-\documentclass[article]{jdssv}
+\documentclass[article]{jdssv}\usepackage[]{graphicx}\usepackage[]{xcolor}
+% maxwidth is the original width if it is less than linewidth
+% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 
@@
 }
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
 % \SweaveOpts{concordance=TRUE}
-<<include=FALSE, cache=FALSE, echo=FALSE, purl=FALSE>>=
-library(knitr)
-opts_chunk$set(comment = NA, size = 'footnotesize', prompt = TRUE, highlight = FALSE, 
-               cache = TRUE, crop = TRUE, concordance = FALSE, fig.align='center',
-               fig.path='Figures_R/fig-',out.width="0.4\\textwidth", 
-               message = F, error = F, warning = F, echo = F)
-opts_knit$set(out.format = "latex")
-knit_hooks$set(crop = hook_pdfcrop)
-options(width = 80, prompt = "R> ", continue = "    ", digits = 3)
-thm <- knit_theme$get('default')
-Sys.setenv(TEXINPUTS=getwd(),
-           BIBINPUTS=getwd(),
-           BSTINPUTS=getwd())
-knit_theme$set(thm)
-@
 
-<<get-nyt-data, include = F>>=
-
-# Read in the data from NYT
-library(tidyverse)
-library(lubridate)
-library(zoo)
-library(ggrepel)
-library(gridExtra)
-cvstate <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")
-
-if (!dir.exists("Data")) dir.create("Data")
-write_csv(cvstate, file = "Data/covid-data-nyt.csv") # Save data just in case
-
-cvstate <- cvstate %>%
-  group_by(state, fips) %>%
-  mutate(date_100th_case = min(date[cases >= 100]),
-         days_100th_case = date - date_100th_case,
-         daily_cases = cases - lag(cases, 1, default = 0),
-         week_avg_cases = rollmean(cases, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
-         daily_avg_cases = rollmean(daily_cases, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
-         daily_deaths = deaths - lag(deaths, 1, default = 0),
-         week_avg_deaths = rollmean(deaths, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
-         daily_avg_deaths = rollmean(daily_deaths, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T)) %>%
-  ungroup() %>%
-  mutate(state = factor(state))
-
-download.file("https://www2.census.gov/programs-surveys/popest/tables/2010-2019/state/totals/nst-est2019-01.xlsx", destfile = "Data/census_pop_state.xlsx", quiet = T, mode = "wb")
-pop_data <- readxl::read_xlsx("Data/census_pop_state.xlsx", col_names = c("state", "pop2019"), col_types = c("text", rep("skip", 11), "numeric"), skip = 9) %>%
-  mutate(state = str_remove(state, "[[:punct:]]"))
-
-cvstate <- left_join(cvstate, pop_data, by = "state")
-
-my_theme <- theme_bw() + 
-  theme(plot.background = element_rect(fill = "transparent", colour = NA), 
-        panel.background = element_rect(fill = "transparent", colour = NA), 
-        legend.background = element_rect(fill = "transparent", colour = NA),
-        axis.text.y = element_text(angle = 90, hjust = 0.5)) 
-theme_set(my_theme)
-@
-
-<<colourpal, include = F>>=
-# Including this for temporary compatibility - we may want to use *shudder* viridis?
-main_colours <- c("#d00000", "#f5f1e7", "#c7c8ca")
-
-# use as accents
-secondary_colours <- c("#001226", "#249ab5")
-
-# These are only to be used for infographics/charts
-tertiary_colours <- c("#bccb2a", "#f58a1f", "#005d84", "#ffd74f", "#a5228d")
-
-colour_palette <- c("#d00000", "#249ab5",  "#1B8500", "#f58a1f", "#a5228d", "#001226", "#B1EB66", "#005d84", "#c25700")
-
-heat_colour_palette <- c("#ffd74f", "#f58a1f", "#c25700", "#d00000", "#9D0000", "#6A0000")
-
-blueseqpal <- c("001226", "06293E", "0C3F56", "12566E", "186D85", "1E839D", 
-                "249AB5", 
-                "39ABC1", "4EBCCE", "64CDDA", "79DDE6", "8EEEF3", "A3FFFF") %>% 
-  paste0("#", .) %>% rev()
-
-blueseqpal_lt <- rev(c("#B2FFFF", "#C2FFFF", "#D1FFFF", "#E0FFFF", "#F0FFFF"))
-
-darkblueseqpal <- colorspace::darken(blueseqpal, .2)
-lightblueseqpal <- colorspace::lighten(blueseqpal, .2)
-@
 %% -- Introduction -------------------------------------------------------------

### 2022-09-06 Review/Response.qmd
Revision 1

@@
 
-Specific reviewer comments are addressed below, here, we address additional discussions brought up by the AE in response to the reviewers' comments.
+**Authors' response:** Specific reviewer comments are addressed below, here, we address additional discussions brought up by the AE in response to the reviewers' comments.
 
@@
 
-We have added a short paragraph addressing this in the data communication subsection of the introduction, and used the opportunity to state that we have primarily focused on charts intended for mass consumption rather than experts in public health or epidemiology.
+**Authors' response:** We have added a short paragraph addressing this in the data communication subsection of the introduction, and used the opportunity to state that we have primarily focused on charts intended for mass consumption rather than experts in public health or epidemiology.
 
 ::: callout
-## Other comments: Proportional symbol maps
+### Other comments: Proportional symbol maps
 
@@
 
-We have added a cartogram to the discussion of proportional symbol maps and choropleths. Thank you for catching an obvious omission on our part.
+**Authors' response:** We have added a cartogram to the discussion of proportional symbol maps and choropleths. Thank you for catching an obvious omission on our part.
 
@@
 
-We have added additional analysis of color where we feel it is useful; Figure 12 is one case where we chose not to mention color in part because I (Susan) can't tell if it's effective or not - I'm colorblind, and while I have trouble with streamgraphs in general, that is particularly true if color is actually meaningful.
+**Authors' response:** We have added additional analysis of color where we feel it is useful; Figure 12 is one case where we chose not to mention color in part because one of the authors (Susan) can't tell if it's effective or not - she is colorblind, and while we have trouble with streamgraphs in general, that is particularly true if color is actually meaningful.
 
@@
 
-We wholeheartedly agree - uncertainty should be mentioned in the introduction, though I can't think of any specific examples of addressing the uncertainty through visualization (as opposed to footnotes about data quality) in mainstream graphics which were circulated at the time.
+**Authors' response:** We wholeheartedly agree - uncertainty should be mentioned in the introduction, though I can't think of any specific examples of addressing the uncertainty through visualization (as opposed to footnotes about data quality) in mainstream graphics which were circulated at the time.
 Still, that is a rather glaring omission from both our discussion and the collection of graphs we've accumulated as interesting.
@@
 
-We've added a moving bubble chart - again, this is an obvious omission.
+**Authors' response:** We've added a moving bubble chart - again, this is an obvious omission.
 Thank you for bringing it to our attention.
@@
 
-There may be other examples, but this is a good one.
+**Authors' response:** There may be other examples, but this is a good one.
 We've added a comment to that effect in the discussion of time series data, but we also mention standardization w.r.t.
@@
 
-## Reviewer 1
+# Reviewer 1
 
@@
 
-The authors are from the United States and Germany; while we participated in many discussions on global social media platforms such as Twitter, we do not have the global perspective (or even facility with sufficient languages) to undertake a review of a global sample of media relating to COVID.
+**Authors' response:** The authors are from the United States and Germany; while we participated in many discussions on global social media platforms such as Twitter, we do not have the global perspective (or even facility with sufficient languages) to undertake a review of a global sample of media relating to COVID.
 We have added a section which explicitly states how charts were selected and what our individual biases are with respect to geography and culture.
@@
 
-We have modified the principles section to remove the reference to Tufte and the data/ink ratio.
+**Authors' response:** We have modified the principles section to remove the reference to Tufte and the data/ink ratio.
 We have slightly reorganized the introduction section to add subsections, and have added an additional paragraph which discusses the series of choices made in the process of designing a data visualization.
@@
 
-While the editor has cautioned us not to double the length of the paper, where possible we have attempted to address some obvious shortcomings in our initial discussion without unduly increasing the paper's length.
+**Authors' response:** While the editor has cautioned us not to double the length of the paper, where possible we have attempted to address some obvious shortcomings in our initial discussion without unduly increasing the paper's length.
 This paper is in part a result of discussing COVID graphics in several groups and seminars with data visualization professionals and also with laypeople; as a result, while we cannot directly cite a source saying that a specific graphic is simple and intuitive, we can directly assess the visual complexity of the chart and we can distill down the series of discussions we've had about these topics over the course of the pandemic to arrive at what we hope are fairly useful conclusions.
@@
 
-## Reviewer 2
+# Reviewer 2
 
@@
 
-We appreciate the edits.
+**Authors' response:** We appreciate the edits.
 We have addressed many of the stylistic suggestions where appropriate.
@@
 
-The organization of this paper is certainly not the same as that which is proposed, in part because we have taken a narrative view that explores the different communication narratives of visualizations during the pandemic.
+**Authors' response:** The organization of this paper is certainly not the same as that which is proposed, in part because we have taken a narrative view that explores the different communication narratives of visualizations during the pandemic.
 Because we focus on the narratives and look at the chart types employed in service of those narratives, organization necessarily deviates from a straightforward analysis of each chart type in sequence, but this is by design.

### 2022-09-18 JDSSV_manuscript_submitted.Rnw
Revision 1
Updates - final round of revisions?

@@
 %% - \Shorttitle{} with LaTeX markup (if any), used as running title
-\title{Visual narratives of the Covid-19 pandemic}
-\Plaintitle{Visual narratives of the Covid-19 pandemic}
-\Shorttitle{Visual narratives of Covid-19}
+\title{Visual narratives of the COVID-19 pandemic}
+\Plaintitle{Visual narratives of the COVID-19 pandemic}
+\Shorttitle{Visual narratives of COVID-19}
 
@@
 \Abstract{
-  Covid-19 has sparked a worldwide interest in understanding the dynamic evolution of a pandemic and tracking the effectiveness of preventive measures and rules.
+  COVID-19 has sparked a worldwide interest in understanding the dynamic evolution of a pandemic and tracking the effectiveness of preventive measures and rules.
 For this reason, numerous media and research groups have produced comprehensive data visualisations to illustrate the relevant trends and figures.
-In this paper, we will look at a selection of Covid 19 data visualisations to evaluate and discuss the currently established visualisation tools in terms of their ability to provide a communication channel both within the data science team and between data analysts, domain experts and a general interested audience.
+In this paper, we will look at a selection of COVID 19 data visualisations to evaluate and discuss the currently established visualisation tools in terms of their ability to provide a communication channel both within the data science team and between data analysts, domain experts and a general interested audience.
 Although there is no set catalogue of evaluation criteria for data visualisations, we will try to give an overview of the different core aspects of visualisation evaluation and their competing principles.
@@
 
-Over the past two years, several waves of Covid-19 infections caused by different mutations of the SARS-CoV-2 virus have swept across the globe, claiming lives, stressing healthcare systems, and changing many different facets of our day-to-day experience.
+Over the past two years, several waves of COVID-19 infections caused by different mutations of the SARS-CoV-2 virus have swept across the globe, claiming lives, stressing healthcare systems, and changing many different facets of our day-to-day experience.
 According to the WHO, almost 600 million confirmed cases and over 6.45 million deaths have been reported as of late August 2022 \footnote{\url{https://www.who.int/docs/default-source/coronaviruse/who_mou_30_august-2022.pdf}}.
 The pandemic has generated enormous interest in epidemiological data, its analysis and visualisation.
-From the beginning of the pandemic, data on the number of infections and covid-related deaths has been published daily and made available to the public.
+From the beginning of the pandemic, data on the number of infections and COVID-related deaths has been published daily and made available to the public.
 Media, politicians and individuals use this data to build their narratives about the pandemic, discuss its evolution, justify the measures taken and discuss different prevention strategies against the spread of the virus.
-%So there are several goals to be achieved by visualising Covid-19 data.
+%So there are several goals to be achieved by visualising COVID-19 data.
 These goals and their priorities were adapted as dynamically as the virus mutated and the pandemic changed pace, but as with any other data visualisation the general principles remain the same: ensuring clear understanding by organising graphics in such a way that the story of the data is told most effectively.
@@
 In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic to showcase their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods.
-COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles but also Data journalism will certainly be one of the beneficiaries of the Covid-19 pandemic and it has become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards\citep{kochWelcomeRevolutionCOVID192021}.
+COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles but also Data journalism will certainly be one of the beneficiaries of the COVID-19 pandemic and it has become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards\citep{kochWelcomeRevolutionCOVID192021}.
 
@@
 At the same time, we still see a lot of defective graphics disseminated and shared: some that violate fundamental statistical and visualisation design principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility.
-These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org).
+These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others \citep{EQUATORNetworkEnhancing}.
 Efforts to improve the communication of quantitative and statistical information date from the beginning mass media and the ability to create charts and tables from data systematically, and while there continue to be efforts made in this area, historical papers outline many of the same issues that we still struggle with today \citep{ASA-standards,haemerPresentationProblemsArea1949,kovermanPLANNINGVISUALPRESENTATION1961,fienbergGraphicalMethodsStatistics1979,Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.
@@
 Presenting information in a visually appealing graphic form and highlighting the most important points in "fact boxes" is one extremely effective way to engage readers while focusing on key takeaway points.
-This approach was developed to increase effective communication between stakeholders such as healthcare providers and patients (see for example: \url{https://www.hardingcenter.de}, archived in ). 
-Such "fact boxes" combined with the dynamic layer of the internet allowed for illustrative simulations of the spread of the epidemic, for example as presented on 14 March 2020 in the Washington Post\footnote{\url{https://www.washingtonpost.com/graphics/2020/world/corona-simulator/}} with the title "Why outbreaks like coronavirus spread exponentially, and how to flatten the curve".
+This approach was developed to increase effective communication between stakeholders such as healthcare providers and patients \citep{FaktenboxenZurMRNASchutzimpfung}. 
+Such "fact boxes" combined with the dynamic layer of the internet allowed for illustrative simulations of the spread of the epidemic, for example as presented on 14 March 2020 in the Washington Post with the title "Why outbreaks like coronavirus spread exponentially, and how to flatten the curve" \citep{stevensTheseSimulationsShow2020}.
 Nevertheless, both the complexity of phenomena and the "bipolarity" of statistical thinking remain a challenge.
@@
 Thus, in studying COVID-19 and predicting its spread, it is important not only to consider symptomatology, disease incidence and geographic distribution, population behaviour patterns, government policies, and impacts on the economy, schools, people in nursing homes, and society at large, but also to incorporate these into data analyses and communication of results.
-Associations observed in the data can often be caused by third party variables (confounders). 
+Associations observed in the data can often be caused by confounding variables. 
 In addition, much of the data comes from observational studies, which usually makes robust causal attribution problematic. 
-However, statisticians who point out these limitations are at risk of having their statements pulled out one-sidedly in a polarised debate \citep{McConway2021}.
+However, statisticians who point out these limitations are at risk of having their statements pulled out to support one side in a polarised debate \citep{McConway2021}.
 
@@
 Having briefly discussed the media landscape surrounding the COVID pandemic and goals for communicating about data effectively, we should also briefly discuss the choices that go into creating visual representations of data intended to convey quantitative information.
-While the overall goal of a data visualization is to tell the story of the data clearly and effectively, how this story is conveyed involves a series of design choices: which variables are shown, what limits are imposed, what forms are used to represent the data?
+While the overall goal of a data visualization is to tell the story of the data clearly and effectively, how this story is conveyed involves a series of design choices: which variables are shown, what limits are imposed, what visual forms are used to represent the data?
 How will the chart (and any surrounding story, in journalism) be presented to the viewer? 
@@
 Viewers tend to integrate similarly coloured areas of the map unconsciously, and perceive choropleths as representations of density.
-They also do not help to convey the desired message as the absolute numbers of Covid-19 cases are strongly influenced by the population size of the country as well as by the number of tests performed and the accuracy of the recording and reporting system.
-\item The area-bias: The visual impression is determined more by the colour and the geographical area of the individual countries than by the number of Covid cases.
+They also do not help to convey the desired message as the absolute numbers of COVID-19 cases are strongly influenced by the population size of the country as well as by the number of tests performed and the accuracy of the recording and reporting system.
+\item The area-bias: The visual impression is determined more by the colour and the geographical area of the individual countries than by the number of COVID cases.
 Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
@@
 	\includegraphics[width = 0.98\textwidth]{Figures_Web/who_totalcases_choro.png}
-	\caption{Choropleth map of the Covid-19 cases world-wide by country. Source: WHO \url{https://covid19.who.int}}
+	\caption{Choropleth map of the COVID-19 cases world-wide by country \citep{WHOCoronavirusCOVID19}.}
 	\label{fig:choro1}
@@
 	\includegraphics[width = 0.98\textwidth]{Figures_Web/wp_totaldeaths_propsymb.png}
-	\caption{Proportional symbol map showing the cumulated reported Covid-19 related deaths adjusted for population world-wide by country.
-Source: Washington Post \url{https://www.washingtonpost.com/graphics/2020/world/mapping-spread-new-coronavirus/}}
+	\caption{Proportional symbol map showing the cumulated reported COVID-19 related deaths adjusted for population world-wide by country \citep{alcantaraMappingWorldwideSpread}.}
 	\label{fig:propsymb}
@@
 The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe: none of the above viusalisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them.
-The use of maps to assess the global pandemic more often focuses on the comparative aspects: how are we doing as opposed to our neighbors? Which strategy to fight Covid-19 is better? 
-We will examine these topics more closely in Section~\ref{sec:rankings}.
+The use of maps to assess the global pandemic more often focuses on the comparative aspects: how are we doing as opposed to our neighbors? Which strategy to fight COVID-19 is better? 
+We will examine these topics more closely in \Cref{sec:rankings}.
 
@@
 \centering
-\begin{subfigure}[c]{.45\textwidth}
-\includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
-\caption{Hexagonal Cartogram of Covid-19 cases in the United States by state. The COVID Tracking Project \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
-\end{subfigure}\hfill
-\begin{subfigure}[c]{.45\textwidth}
+\begin{subfigure}[c]{\textwidth}
+\centering
+\includegraphics[width=.65\textwidth]{COVID-tracking-hex-cartogram}
+\caption{Hexagonal Cartogram of COVID-19 cases in the United States by state \citep{StateCases1M2021}.}\label{fig:hex-cartogram}
+\end{subfigure}
+
+\begin{subfigure}[c]{\textwidth}
+\centering
 \includegraphics[width=\textwidth]{china-dotmap}
@@
 % \begin{subfigure}[t]{.45\textwidth}
-% \includegraphics[width=\textwidth]{nyt-covid-scaled-choro}
-% \caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
+% \includegraphics[width=\textwidth]{nyt-COVID-scaled-choro}
+% \caption{Choropleth Map of COVID cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/COVID-cases.html}.}\label{fig:scaled-choro}
 % \end{subfigure}
@@
 \end{figure}
+% 
+% \begin{minipage}{\linewidth}
+% \begin{minipage}{.7\linewidth}
+% \includegraphics[width=\textwidth]{COVID-tracking-hex-cartogram}
+% \end{minipage}
+% \begin{minipage}{.29\linewidth}
+% \captionof{figure}{Hexagonal Cartogram of COVID-19 cases in the United States by state.
+The COVID Tracking Project \url{https://COVIDtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
+% \end{minipage}
+% \end{minipage}
+% 
+% \begin{minipage}{\linewidth}
+% \begin{minipage}{.7\linewidth}
+% \includegraphics[width=\textwidth]{china-dotmap}
+% \end{minipage}
+% \begin{minipage}{.29\linewidth}
+% \captionof{figure}{A dot-density map of COVID cases in China \citep{fieldMappingCoronavirusResponsibly}.}\label{fig:china-dot-map}
+% \end{minipage}
+% \end{minipage}
 
@@
 % \begin{subfigure}[t]{.45\textwidth}
-% \includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
-% \caption{Hexagonal Cartogram of Covid-19 cases in the United States by state, per million people.
+% \includegraphics[width=\textwidth]{COVID-tracking-hex-cartogram}
+% \caption{Hexagonal Cartogram of COVID-19 cases in the United States by state, per million people.
 This somewhat mitigates the area bias issue, but is still subject to other perceptual biases.
-Source: The COVID Tracking Project at The Atlantic \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
+Source: The COVID Tracking Project at The Atlantic \url{https://COVIDtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
 % \end{subfigure}\hfill
 % \begin{subfigure}[t]{.45\textwidth}
-% \includegraphics[width=\textwidth]{nyt-covid-scaled-choro}
-% \caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
+% \includegraphics[width=\textwidth]{nyt-COVID-scaled-choro}
+% \caption{Choropleth Map of COVID cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/COVID-cases.html}.}\label{fig:scaled-choro}
 % \end{subfigure}
@@
 % Is it worth discussing e.g.
-https://www.nytimes.com/interactive/2021/us/covid-risk-map.html that does not make this distinction and runs into the "boy who cried wolf" problem?
+https://www.nytimes.com/interactive/2021/us/COVID-risk-map.html that does not make this distinction and runs into the "boy who cried wolf" problem?
 When everything is always extremely high risk we just tend to tune things out.
@@
 Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time, as there are new developments on a daily basis.
-Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis, as in \Cref{fig-jbm-tweet} and \Cref{fig:wapo-covid-time-series}. 
+Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis, as in \Cref{fig:jbm-tweet} and \Cref{fig:wapo-COVID-time-series}. 
 This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state. 
@@
 \centering
-\includegraphics[width=.99\linewidth]{wapo-covid-monitoring}
-\caption{COVID monitoring summary charts on the Washington Post's Coronavirus page. Source: \url{https://www.washingtonpost.com/coronavirus/}, 2022-01-17.}
-\label{fig:wapo-covid-time-series}
+\includegraphics[width=.99\linewidth]{wapo-COVID-monitoring}
+\caption{COVID monitoring summary charts on the Washington Post's Coronavirus page \citep{Coronavirus2022}.}
+\label{fig:wapo-COVID-time-series}
 \end{figure}
 
-The most obvious time-series plot is also the most common: plotting cases against date, often with a 7 or 14-day smooth to handle periodicity in case reporting due to the work week, as shown in \Cref{fig:wapo-covid-time-series}.
+The most obvious time-series plot is also the most common: plotting cases against date, often with a 7 or 14-day smooth to handle periodicity in case reporting due to the work week, as shown in \Cref{fig:wapo-COVID-time-series}.
 This allows for individuals to assess the current situation and easily compare the current number of cases to past times for which the individual has a direct reference for what to expect in impact to daily life.
-Often, as in \Cref{fig:wapo-covid-time-series}, these time series show multiple variables: cases, deaths, and hospitalizations (and sometimes vaccinations) to provide a more comprehensive view of the situation; this is particularly useful as there has been some decoupling of case numbers and hospitalizations/deaths with the emergence of effective vaccinations and strains which are more contagious but potentially also less severe.
+Often, as in \Cref{fig:wapo-COVID-time-series}, these time series show multiple variables: cases, deaths, and hospitalizations (and sometimes vaccinations) to provide a more comprehensive view of the situation; this is particularly useful as there has been some decoupling of case numbers and hospitalizations/deaths with the emergence of effective vaccinations and strains which are more contagious but potentially also less severe.
 Many outlets also provide additional charts designed to provide some immediate context to readers, as shown in \Cref{fig:wapo-context}.
@@
 \centering
-\includegraphics[width=.5\linewidth]{wapo-covid-context}
-\caption{Some outlets provide contextual information to assist readers with interpreting time-series plots.
-Source: Washington Post Coronavirus Tracking, \url{https://www.washingtonpost.com/graphics/2020/national/coronavirus-us-cases-deaths/}.}
+\includegraphics[width=.5\linewidth]{wapo-COVID-context}
+\caption{Some outlets provide contextual information to assist readers with interpreting time-series plots \citep{CoronavirusCasesTracking2022}.}
 \label{fig:wapo-context}
@@
 \includegraphics{ft_vaccine_bubble}
-\caption{Animated bubble chart showing vaccination rate relative to per capita GDP for each country, from the Financial Times (Source: \url{https://ig.ft.com/coronavirus-vaccine-tracker})}\label{fig:bubble-ft}
+\caption{Animated bubble chart showing vaccination rate relative to per capita GDP for each country \citep{visualCovid19VaccineTracker}.}\label{fig:bubble-ft}
 \end{figure}
@@
 This provides viewers with a sense of the pace of the epidemic, rather than the raw case counts in more standard time-series representations.
-This representation can also be shown in the auditory domain, providing access to those who are typically excluded from visualizations due to vision loss.
-Source: Romain Vuillemot, \url{https://observablehq.com/@romsson/sonification-of-covid-19-data}.}
+This representation can also be shown in the auditory domain, providing access to those who are typically excluded from visualizations due to vision loss \citep{vuillemotSonificationCOVID19Data2020}.}
 \label{fig:freq-domain}
@@
 There are many different ways that designers can use time-series data to provide additional contextual information and facilitate comparisons.
-\Cref{fig:ft-waves} shows covid cases, positivity rate, and hospital admissions from South Africa's Gauteng province, which was one of the first areas to experience a large wave of the Omicron variant.
+\Cref{fig:ft-waves} shows COVID cases, positivity rate, and hospital admissions from South Africa's Gauteng province, which was one of the first areas to experience a large wave of the Omicron variant.
 This chart provides context for the Omicron wave, showing that while infections are occurring at a much faster rate than in previous waves, hospital admissions seem to be approximately following previous trends.
@@
 \includegraphics[width=\linewidth]{ft-waves-linear}
-\caption{Covid Cases, positivity rate, and hospital admissions from South Africa's Gauteng province, on a log scale (top) and linear scale (bottom).
+\caption{COVID Cases, positivity rate, and hospital admissions from South Africa's Gauteng province, on a log scale (top) and linear scale (bottom).
 Successive waves are overlaid, showing that the increase in cases due to the Omicron variant was sharper than any previously measured increase, while hospitalizations were much more comparable to previous waves \citep{burn-murdoch_new_2021-1}.}
@@
 While traditionally, time-series information has been limited to charts with time on a linear x-axis, many charts manipulated time to highlight different aspects of the pandemic.
-In some cases, these modifications provided large amounts of utility for a relatively small increase in complexity: for example, John Burn-Murdoch's decision to index Financial Times covid graphs to the days since the $N$th case or death.
+In some cases, these modifications provided large amounts of utility for a relatively small increase in complexity: for example, John Burn-Murdoch's decision to index Financial Times COVID graphs to the days since the $N$th case or death.
 This decision is a small modification of the basic time-series plot, but has important consequences in the chart's readability as well as the visual emphasis for users: the graphs in the Financial Times were designed to emphasize "...
@@
 \includegraphics[width=.45\linewidth]{nyt-spiral}\hfill\includegraphics[width=.45\linewidth]{spiral-rework}
-\caption{Covid cases in the United States, 2020-2022.
-The original graph from the New York Times is on the left, and a reenvisioned version which is more perceptually friendly is on the right.
-Source: (Left) New York Times \url{https://www.nytimes.com/2022/01/06/opinion/omicron-covid-us.html}, (Right) Covid Spiral, by Sourya Shrestha, \url{https://github.com/souryashrestha/covid_spiral}.}
+\caption{COVID cases in the United States, 2020-2022.
+The original graph from the New York Times is on the left \citep{shamanOpinionHereWhen2022}, and a reenvisioned version which is more perceptually friendly is on the right \citep{shresthaCovidSpiral2022}.}
 \label{fig:nyt-spiral}
@@
 In some cases, this took the form of animated bubble plots XXX add XXX, while others dispensed with the traditional form of a line graph altogether, showing rising case counts using a "bar chart race": an animated series of bar charts over time that show how case counts in each country are increasing.
-A typical example can be found on YouTube at \url{https://www.youtube.com/watch?v=8WH-bTH6GqI}; a screenshot is shown in \Cref{fig:barchart-race}.
+A typical example can be found on YouTube\citep{clockCoronavirusCasesBar2022}; a screenshot is shown in \Cref{fig:barchart-race}.
 This type of chart is effective in showing instantaneous changes in counts over days for countries with an extremely high number of cases, but may make it difficult for the viewer to process how cases are changing in countries that do not dominate case counts.
@@
 \includegraphics[width=.8\linewidth]{Figures_Web/Bar-chart-race-screenshot}
-\caption{A screenshot from Stats on Clock's bar chart race showing counts of covid cases over time throughout the pandemic.
-Source: \url{https://www.youtube.com/watch?v=8WH-bTH6GqI}}\label{fig:barchart-race}
+\caption{A screenshot from Stats on Clock's bar chart race showing counts of COVID cases over time throughout the pandemic \citep{clockCoronavirusCasesBar2022}.}\label{fig:barchart-race}
 \end{figure}
@@
 
-A slightly different approach by the BBC's Visual and Data Journalism team, shown in \Cref{fig:bbc-flower}, displayed the Covid cases and deaths as a growing flower, with the stem proportional to the number of cases and the flower petals showing the deaths over time\citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}.
+A slightly different approach by the BBC's Visual and Data Journalism team, shown in \Cref{fig:bbc-flower}, displayed the COVID cases and deaths as a growing flower, with the stem proportional to the number of cases and the flower petals showing the deaths over time\citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}.
 The visualization comes with sound as well, an attempt to make the pandemic an auditory experience as well as a visual one.
@@
 \includegraphics[width=.8\linewidth]{Figures_Web/BBC_flower_time_series}
-\caption{The BBC Visual and Data Journalism Team's Covid flower, showing the cases and deaths over time.
-Source: \url{https://www.bbc.co.uk/news/resources/idt-7464500a-6368-4029-aa41-ab94e0ee09fb}}\label{fig:bbc-flower}
+\caption{The BBC Visual and Data Journalism Team's COVID flower, showing the cases and deaths over time \citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}.}\label{fig:bbc-flower}
 \end{figure}
@@
 Unsurprisingly, the prevailing visual narrative focused on comparisons often fueled by political rivalry, historical dependencies, recent withdrawal from supranational institutions or regional competitions.
-An ongoing debate about the true extent of the dangers of covid-19 and how best to combat it, combined with daily availability and public access to data across administrative levels, fostered ongoing competition and the use of leaderboards to show how bad case counts were somewhere else.
+An ongoing debate about the true extent of the dangers of COVID-19 and how best to combat it, combined with daily availability and public access to data across administrative levels, fostered ongoing competition and the use of leaderboards to show how bad case counts were somewhere else.
 
@@
 \includegraphics[width=\linewidth]{Figures_Web/ft_deathcounts_area}
-\caption{Ordered death rates in countries and the temporal evolution of the death toll share in major world regions.
-Source: Financial Times, \url{https://www.ft.com/content/a2901ce8-5eb7-4633-b89c-cbdf5b386938}}
+\caption{Ordered death rates in countries and the temporal evolution of the death toll share in major world regions \citep{CoronavirusTrackerLatest2021}.}
 \label{fig:ft-death-ranking}
@@
 Initially, log scales were incredibly useful at showing case counts, because minimal mitigation measures were in place and the growth of case counts (or presumptive positive cases, in absence of available testing) was fairly close to exponential.
-In addition, the use of log scales allowed for the comparison of nominal cases across entities with large population differences: in the US, we could compare cases in New York and California with cases in Michigan and Washington, even though the population of Michigan and Washington are much lower than the population of either New York or California.
+In addition, the use of log scales allowed for the comparison of total cases across entities with large population differences: in the US, we could compare cases in New York and California with cases in Michigan and Washington, even though the populations (and corresponding case counts) of Michigan and Washington are much lower than the populations (and case counts) of either New York or California.
 
@@
 \centering
-\includegraphics[width=.8\linewidth]{ft-covid-19-deaths}
+\includegraphics[width=.8\linewidth]{ft-COVID-19-deaths}
 \caption{Reference lines to compare exponential growth rates of deaths in different countries.
@@
 Linear scales do not have this problem - the exponential source is noticeable very quickly in the total line, but on the log scale it is much harder to discern when the exponential source causes the total line to diverge from the background.
-In the top-right corner, it is difficult to identify that there is an exponential increase in cases amid the baseline, even though the exponential source makes up approximately 50\\% of the cases at the end of the time period shown." >>=
+In the top-right corner, it is difficult to identify that there is an exponential increase in cases amid the baseline of a uniform level of spread over time, even though the exponential source makes up approximately 50\\% of the cases at the end of the time period shown." >>=
 # Set up labeller to tweak things 
@@
 \caption{An annotated time series from the Financial Times which appeared on Twitter on November 12, 2020.
-This chart has annotations which show the decreasing $R_{eff}$ in successive peaks of the pandemic, with periods of $R_{eff}<1$ between surges in cases.
-These annotations assist the reader with drawing complex conclusions from the chart, but are difficult to automate and thus tend to be manually curated.
-Source: Mark Gubrud (@mgubrud), \url{https://twitter.com/mgubrud/status/1326784082399944704}.}
+This chart has annotations which show the decreasing $R_{eff}$, the effective rate of exponential spread, in successive peaks of the pandemic, with periods of $R_{eff}<1$ between surges in cases.
+These annotations assist the reader with drawing complex conclusions from the chart, but are difficult to automate and thus tend to be manually curated \citep{gubrudDrEricDingWeVe2020}.}
 \label{fig:log-scale-ref-multiple}
@@
 
-\section{Summary and discussion} \label{sec:summary}
+\section{Summary and Discussion} \label{sec:summary}
 Throughout the pandemic (thus far), visual representations of data have been an integral part of scientific communication.
@@
 The visual narratives of the pandemic which have persisted over time are due in part to the rise of data journalism and interactive graphics platforms which ensure that every outlet has the ability to host engaging and visually appealing graphics; however, while these graphics are nice to look at, not all are equally functional or useful for communication purposes.
+Thus, analyzing the rich collection of visualizations produced as part of the pandemic provides a unique opportunity to examine how these graphics were (and were not) effective, as well as the design decisions made during their creation.
 As in any situation, good tools can be used and misused freely; we have seen charts and graphics used to perpetuate misinformation and misleading claims, even though it is far more common for charts and graphs to be used for good - to educate and inform the public.
 
-The Covid 19 pandemic has amply demonstrated that policies are only effectively implemented and followed by the population if they are accepted by a large majority of the population.
+The COVID 19 pandemic has amply demonstrated that policies are only effectively implemented and followed by the population if they are accepted by a large majority of the population.
 To achieve this goal, it is essential that scientists, governments, the media, and citizens all effectively communicate in order to justify the appropriateness, usefulness and relevance of the measures.
@@
 
-% Ubiquity of the data, persisting general interest on the topic, the rise of data journalism, and the opportunities provided by interactive and dynamic internet platforms contributed to a an ongoing trend on visual narratives of the Covid-19 pandemic.
+% Ubiquity of the data, persisting general interest on the topic, the rise of data journalism, and the opportunities provided by interactive and dynamic internet platforms contributed to a an ongoing trend on visual narratives of the COVID-19 pandemic.
 Graphical displays are used as weapons for information, but also for misinformation.
@@
 
+Archived data and code to create the figures in this paper can be found at \url{https://github.com/srvanderplas/covid-visualization-dssv}.
+
+\section*{Archived Information}
+This paper references many news articles which contain interactive graphics.
+In order to preserve these graphics (which depend on a great many external files), the authors have preserved many of the interactive graphics as self-contained html files \citep{news-file-archive}.
+The authors have also made extensive use of the Internet Wayback Machine, which aims to preserve the state of important web pages as they change over time.
+
 \section*{Acknowledgments}
 
-First ideas have been presented at DSSV 2020. We thank our colleagues and the editors for valuable feedback and encouragement to complete this manuscript.
+First ideas have been presented at DSSV 2020. We thank our colleagues, the reviewers, and the editors for valuable feedback and encouragement to complete and revise this manuscript.
 

### 2022-09-18 JDSSV_manuscript_submitted.tex
Updates - final round of revisions?

--- 2022-09-02 Review/Edits/JDSSV_manuscript_submitted.tex
@@
 	\includegraphics[width = 0.98\textwidth]{Figures_Web/who_totalcases_choro.png}
-	\caption{Choropleth map of the Covid-19 cases world-wide by country. Source: WHO \url{https://covid19.who.int}}
+	\caption{Choropleth map of the Covid-19 cases world-wide by country \citep{WHOCoronavirusCOVID19}.}
 	\label{fig:choro1}
@@
 	\includegraphics[width = 0.98\textwidth]{Figures_Web/wp_totaldeaths_propsymb.png}
-	\caption{Proportional symbol map showing the cumulated reported Covid-19 related deaths adjusted for population world-wide by country.
-Source: Washington Post \url{https://www.washingtonpost.com/graphics/2020/world/mapping-spread-new-coronavirus/}}
+	\caption{Proportional symbol map showing the cumulated reported Covid-19 related deaths adjusted for population world-wide by country \citep{alcantaraMappingWorldwideSpread}.}
 	\label{fig:propsymb}
@@
 The use of maps to assess the global pandemic more often focuses on the comparative aspects: how are we doing as opposed to our neighbors? Which strategy to fight Covid-19 is better? 
-We will examine these topics more closely in Section~\ref{sec:rankings}.
+We will examine these topics more closely in \Cref{sec:rankings}.
 
@@
 \centering
-\begin{subfigure}[c]{.45\textwidth}
-\includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
-\caption{Hexagonal Cartogram of Covid-19 cases in the United States by state. The COVID Tracking Project \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
-\end{subfigure}\hfill
-\begin{subfigure}[c]{.45\textwidth}
+\begin{subfigure}[c]{\textwidth}
+\centering
+\includegraphics[width=.65\textwidth]{covid-tracking-hex-cartogram}
+\caption{Hexagonal Cartogram of Covid-19 cases in the United States by state \citep{StateCases1M2021}.}\label{fig:hex-cartogram}
+\end{subfigure}
+
+\begin{subfigure}[c]{\textwidth}
+\centering
 \includegraphics[width=\textwidth]{china-dotmap}
@@
 \end{figure}
+% 
+% \begin{minipage}{\linewidth}
+% \begin{minipage}{.7\linewidth}
+% \includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
+% \end{minipage}
+% \begin{minipage}{.29\linewidth}
+% \captionof{figure}{Hexagonal Cartogram of Covid-19 cases in the United States by state.
+The COVID Tracking Project \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
+% \end{minipage}
+% \end{minipage}
+% 
+% \begin{minipage}{\linewidth}
+% \begin{minipage}{.7\linewidth}
+% \includegraphics[width=\textwidth]{china-dotmap}
+% \end{minipage}
+% \begin{minipage}{.29\linewidth}
+% \captionof{figure}{A dot-density map of COVID cases in China \citep{fieldMappingCoronavirusResponsibly}.}\label{fig:china-dot-map}
+% \end{minipage}
+% \end{minipage}
 
@@
 Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time, as there are new developments on a daily basis.
-Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis, as in \Cref{fig-jbm-tweet} and \Cref{fig:wapo-covid-time-series}. 
+Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis, as in \Cref{fig:jbm-tweet} and \Cref{fig:wapo-covid-time-series}. 
 This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state. 
@@
 \includegraphics[width=.99\linewidth]{wapo-covid-monitoring}
-\caption{COVID monitoring summary charts on the Washington Post's Coronavirus page. Source: \url{https://www.washingtonpost.com/coronavirus/}, 2022-01-17.}
+\caption{COVID monitoring summary charts on the Washington Post's Coronavirus page \citep{Coronavirus2022}.}
 \label{fig:wapo-covid-time-series}
@@
 \includegraphics[width=.5\linewidth]{wapo-covid-context}
-\caption{Some outlets provide contextual information to assist readers with interpreting time-series plots.
-Source: Washington Post Coronavirus Tracking, \url{https://www.washingtonpost.com/graphics/2020/national/coronavirus-us-cases-deaths/}.}
+\caption{Some outlets provide contextual information to assist readers with interpreting time-series plots \citep{CoronavirusCasesTracking2022}.}
 \label{fig:wapo-context}
@@
 \caption{Covid cases in the United States, 2020-2022.
-The original graph from the New York Times is on the left, and a reenvisioned version which is more perceptually friendly is on the right.
-Source: (Left) New York Times \url{https://www.nytimes.com/2022/01/06/opinion/omicron-covid-us.html}, (Right) Covid Spiral, by Sourya Shrestha, \url{https://github.com/souryashrestha/covid_spiral}.}
+The original graph from the New York Times is on the left \citep{shamanOpinionHereWhen2022}, and a reenvisioned version which is more perceptually friendly is on the right \citep{shresthaCovidSpiral2022}.}
 \label{fig:nyt-spiral}
@@
 \includegraphics[width=.8\linewidth]{Figures_Web/Bar-chart-race-screenshot}
-\caption{A screenshot from Stats on Clock's bar chart race showing counts of covid cases over time throughout the pandemic.
-Source: \url{https://www.youtube.com/watch?v=8WH-bTH6GqI}}\label{fig:barchart-race}
+\caption{A screenshot from Stats on Clock's bar chart race showing counts of covid cases over time throughout the pandemic \citep{clockCoronavirusCasesBar2022}.}\label{fig:barchart-race}
 \end{figure}
@@
 \includegraphics[width=.8\linewidth]{Figures_Web/BBC_flower_time_series}
-\caption{The BBC Visual and Data Journalism Team's Covid flower, showing the cases and deaths over time.
-Source: \url{https://www.bbc.co.uk/news/resources/idt-7464500a-6368-4029-aa41-ab94e0ee09fb}}\label{fig:bbc-flower}
+\caption{The BBC Visual and Data Journalism Team's Covid flower, showing the cases and deaths over time \citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}.}\label{fig:bbc-flower}
 \end{figure}
@@
 Linear scales do not have this problem - the exponential source is noticeable very quickly in the total line, but on the log scale it is much harder to discern when the exponential source causes the total line to diverge from the background.
-In the top-right corner, it is difficult to identify that there is an exponential increase in cases amid the baseline, even though the exponential source makes up approximately 50\% of the cases at the end of the time period shown.}\label{fig:log-scale-failures}
+In the top-right corner, it is difficult to identify that there is an exponential increase in cases amid the baseline of a uniform level of spread over time, even though the exponential source makes up approximately 50\% of the cases at the end of the time period shown.}\label{fig:log-scale-failures}
 \end{figure}

### 2022-09-18 JDSSV_manuscript_submitted.tex
Updates

@@
 	\includegraphics[width = 0.98\textwidth]{Figures_Web/who_totalcases_choro.png}
-	\caption{Choropleth map of the Covid-19 cases world-wide by country \citep{WHOCoronavirusCOVID19}.}
+	\caption{Choropleth map of the COVID-19 cases world-wide by country \citep{WHOCoronavirusCOVID19}.}
 	\label{fig:choro1}
@@
 	\includegraphics[width = 0.98\textwidth]{Figures_Web/wp_totaldeaths_propsymb.png}
-	\caption{Proportional symbol map showing the cumulated reported Covid-19 related deaths adjusted for population world-wide by country \citep{alcantaraMappingWorldwideSpread}.}
+	\caption{Proportional symbol map showing the cumulated reported COVID-19 related deaths adjusted for population world-wide by country \citep{alcantaraMappingWorldwideSpread}.}
 	\label{fig:propsymb}
@@
 The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe: none of the above viusalisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them.
-The use of maps to assess the global pandemic more often focuses on the comparative aspects: how are we doing as opposed to our neighbors? Which strategy to fight Covid-19 is better? 
+The use of maps to assess the global pandemic more often focuses on the comparative aspects: how are we doing as opposed to our neighbors? Which strategy to fight COVID-19 is better? 
 We will examine these topics more closely in \Cref{sec:rankings}.
@@
 \centering
-\includegraphics[width=.65\textwidth]{covid-tracking-hex-cartogram}
-\caption{Hexagonal Cartogram of Covid-19 cases in the United States by state \citep{StateCases1M2021}.}\label{fig:hex-cartogram}
+\includegraphics[width=.65\textwidth]{COVID-tracking-hex-cartogram}
+\caption{Hexagonal Cartogram of COVID-19 cases in the United States by state \citep{StateCases1M2021}.}\label{fig:hex-cartogram}
 \end{subfigure}
@@
 % \begin{minipage}{.7\linewidth}
-% \includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
+% \includegraphics[width=\textwidth]{COVID-tracking-hex-cartogram}
 % \end{minipage}
 % \begin{minipage}{.29\linewidth}
-% \captionof{figure}{Hexagonal Cartogram of Covid-19 cases in the United States by state.
-The COVID Tracking Project \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
+% \captionof{figure}{Hexagonal Cartogram of COVID-19 cases in the United States by state.
+The COVID Tracking Project \url{https://COVIDtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
 % \end{minipage}
@@
 Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time, as there are new developments on a daily basis.
-Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis, as in \Cref{fig:jbm-tweet} and \Cref{fig:wapo-covid-time-series}. 
+Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis, as in \Cref{fig:jbm-tweet} and \Cref{fig:wapo-COVID-time-series}. 
 This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state. 
@@
 \centering
-\includegraphics[width=.99\linewidth]{wapo-covid-monitoring}
+\includegraphics[width=.99\linewidth]{wapo-COVID-monitoring}
 \caption{COVID monitoring summary charts on the Washington Post's Coronavirus page \citep{Coronavirus2022}.}
-\label{fig:wapo-covid-time-series}
+\label{fig:wapo-COVID-time-series}
 \end{figure}
 
-The most obvious time-series plot is also the most common: plotting cases against date, often with a 7 or 14-day smooth to handle periodicity in case reporting due to the work week, as shown in \Cref{fig:wapo-covid-time-series}.
+The most obvious time-series plot is also the most common: plotting cases against date, often with a 7 or 14-day smooth to handle periodicity in case reporting due to the work week, as shown in \Cref{fig:wapo-COVID-time-series}.
 This allows for individuals to assess the current situation and easily compare the current number of cases to past times for which the individual has a direct reference for what to expect in impact to daily life.
-Often, as in \Cref{fig:wapo-covid-time-series}, these time series show multiple variables: cases, deaths, and hospitalizations (and sometimes vaccinations) to provide a more comprehensive view of the situation; this is particularly useful as there has been some decoupling of case numbers and hospitalizations/deaths with the emergence of effective vaccinations and strains which are more contagious but potentially also less severe.
+Often, as in \Cref{fig:wapo-COVID-time-series}, these time series show multiple variables: cases, deaths, and hospitalizations (and sometimes vaccinations) to provide a more comprehensive view of the situation; this is particularly useful as there has been some decoupling of case numbers and hospitalizations/deaths with the emergence of effective vaccinations and strains which are more contagious but potentially also less severe.
 Many outlets also provide additional charts designed to provide some immediate context to readers, as shown in \Cref{fig:wapo-context}.
@@
 \centering
-\includegraphics[width=.5\linewidth]{wapo-covid-context}
+\includegraphics[width=.5\linewidth]{wapo-COVID-context}
 \caption{Some outlets provide contextual information to assist readers with interpreting time-series plots \citep{CoronavirusCasesTracking2022}.}
@@
 \includegraphics[width=.45\linewidth]{nyt-spiral}\hfill\includegraphics[width=.45\linewidth]{spiral-rework}
-\caption{Covid cases in the United States, 2020-2022.
+\caption{COVID cases in the United States, 2020-2022.
 The original graph from the New York Times is on the left \citep{shamanOpinionHereWhen2022}, and a reenvisioned version which is more perceptually friendly is on the right \citep{shresthaCovidSpiral2022}.}
@@
 \includegraphics[width=.8\linewidth]{Figures_Web/Bar-chart-race-screenshot}
-\caption{A screenshot from Stats on Clock's bar chart race showing counts of covid cases over time throughout the pandemic \citep{clockCoronavirusCasesBar2022}.}\label{fig:barchart-race}
+\caption{A screenshot from Stats on Clock's bar chart race showing counts of COVID cases over time throughout the pandemic \citep{clockCoronavirusCasesBar2022}.}\label{fig:barchart-race}
 \end{figure}
@@
 \includegraphics[width=.8\linewidth]{Figures_Web/BBC_flower_time_series}
-\caption{The BBC Visual and Data Journalism Team's Covid flower, showing the cases and deaths over time \citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}.}\label{fig:bbc-flower}
+\caption{The BBC Visual and Data Journalism Team's COVID flower, showing the cases and deaths over time \citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}.}\label{fig:bbc-flower}
 \end{figure}

### 2022-10-27 JDSSV_manuscript_submitted.Rnw
copyediting changes

@@
 
-In other areas, such as visualization of uncertainty, the pandemic presents new challenges for scientific communication.
+In other areas, such as visualisation of uncertainty, the pandemic presents new challenges for scientific communication.
 How do we effectively show the uncertainty in the data amid flawed tests, new strains, changing diagnostic criteria, and incredible variability in global public health infrastructure?
@@
 
-\subsection{Data Visualization Design and Testing}
+\subsection{Data visualisation Design and Testing}
 Having briefly discussed the media landscape surrounding the COVID pandemic and goals for communicating about data effectively, we should also briefly discuss the choices that go into creating visual representations of data intended to convey quantitative information.
-While the overall goal of a data visualization is to tell the story of the data clearly and effectively, how this story is conveyed involves a series of design choices: which variables are shown, what limits are imposed, what visual forms are used to represent the data?
+While the overall goal of a data visualisation is to tell the story of the data clearly and effectively, how this story is conveyed involves a series of design choices: which variables are shown, what limits are imposed, what visual forms are used to represent the data?
 How will the chart (and any surrounding story, in journalism) be presented to the viewer? 
-As the designer makes these choices, they have in mind an audience, and the audience's assumed characteristics affect the design choices as well: language, education level, likely medium used to access the visualization (computer, physical newspaper, tablet, cell phone) - all of these considerations also impact the design of the chart and what ideas the designer considers during the creative process.
+As the designer makes these choices, they have in mind an audience, and the audience's assumed characteristics affect the design choices as well: language, education level, likely medium used to access the visualisation (computer, physical newspaper, tablet, cell phone) - all of these considerations also impact the design of the chart and what ideas the designer considers during the creative process.
 
-Once the basic form of the chart is determined, there are additional considerations that may be specified by the stylistic conventions of different news organizations, but may also be deliberately chosen for visual effect: color palettes, fonts, spatial layouts, and annotations may be used to highlight information, convey additional points of a narrative, or provide helpful information about how to read an unfamiliar chart.
+Once the basic form of the chart is determined, there are additional considerations that may be specified by the stylistic conventions of different news organizations, but may also be deliberately chosen for visual effect: colour palettes, fonts, spatial layouts, and annotations may be used to highlight information, convey additional points of a narrative, or provide helpful information about how to read an unfamiliar chart.
 While each of these design choices could be (and have been) the subject of  additional paper(s), it is important to acknowledge that the graphics we analyze in this paper did not simply pop into being - they are the product of a series of decisions, some consciously made and others imposed by convention or technological capabilities.
@@
 Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
-\item Colour-scheme selection: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997, color2021}.
+\item colour-scheme selection: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997, color2021}.
 The choice of a continuous scale or a categorical scale, the choice of a scale that promotes the recognition of patterns or a scale that supports the filtering out of specific map details, influences the quality of choropleth maps.
@@
 
-Another modification of the standard choropleth to address the area-bias problem is to relax the spatial borders of a region in order to show magnitude using area (as well as color).
+Another modification of the standard choropleth to address the area-bias problem is to relax the spatial borders of a region in order to show magnitude using area (as well as colour).
 Various algorithms can be used to modify geographic boundaries to maintain spatial continuity, or shapes such as hexagons or squares can be used to represent geographic entities (as in \Cref{fig:hex-cartogram}); with either choice, geometry, topology, or both must be sacrificed.
@@
 
-Throughout the pandemic, both in choropleth maps and time-series graphs, creators have had to grapple with scale.
+Throughout the pandemic, both in choropleth maps and time series graphs, creators have had to grapple with scale.
 When scales automatically adjust, mental comparisons across time can become misleading, because the colour scale from yesterday does not mean the same thing as it did today, as the range has expanded and the colours have stayed the same.
 It is not that often that day-to-day design changes (or lack thereof) cause fundamental problems with the interpretation of data visualisation, but during the COVID 19 pandemic this situation was both extremely common and had real consequences: people would glance at the current COVID risk map and assume that "red" meant the same thing that it did yesterday, despite the fact that the underlying values had changed\citep{abiadColorDiagramWHOLebanon2020,rubelHeyUpshotNYTCurious2020,matthewsLessonFutzingData2020}.
-One example of this issue over two months of NY Times graphics is shown in \Cref{fig:sparklines-heatmap-nyt}; the overall color scheme used changes once, but the variable displayed changes (at least) 4 times over the course of a two month period.
+One example of this issue over two months of NY Times graphics is shown in \Cref{fig:sparklines-heatmap-nyt}; the overall colour scheme used changes once, but the variable displayed changes (at least) 4 times over the course of a two month period.
 
-This problem is more noticeable in choropleths and other charts which show case counts based on color scale alone than in time-series charts, where we are more likely to expect and notice a scale change, but it is still a fundamental issue across chart types and of particular concern when there is no natural upper bound on the variable displayed.
+This problem is more noticeable in choropleths and other charts which show case counts based on colour scale alone than in time series charts, where we are more likely to expect and notice a scale change, but it is still a fundamental issue across chart types and of particular concern when there is no natural upper bound on the variable displayed.
 Some designers leaned into this by arranging their graphics to highlight the fact that cases were ``off the charts", as in \Cref{fig:jbm-tweet}, implicitly cuing readers that the situation was exceptional and worthy of their attention.
@@
 Early in the pandemic, as case numbers grew exponentially without appropriate public health measures to restrain them, the scale problem for choropleth maps was particularly complex: over time either the colour mapping changes or the map becomes useless due to compression against the upper limits of the scale, as in \citet{panchadsaramIfWeAre2020}.
-The magnitude of the problem also depends on whether the scale uses a set of bins (a discrete mapping to a continuous variable) or uses a continuous color mapping; in some cases, outlets switched between the two options over time or changed the underlying unit entirely.
-Interestingly, media outlets and municipal dashboards were subject to criticism when the color scales changed in response to increasing case counts\citep{boeckDataVizDecisionmakingAction2020} as well as when the color scale did not change\citep{sandalowChiTribGraphicsPleaseUpdate2020}, illustrating the difficulty of designing visualizations amid competing constraints and a constantly evolving pandemic.
+The magnitude of the problem also depends on whether the scale uses a set of bins (a discrete mapping to a continuous variable) or uses a continuous colour mapping; in some cases, outlets switched between the two options over time or changed the underlying unit entirely.
+Interestingly, media outlets and municipal dashboards were subject to criticism when the colour scales changed in response to increasing case counts\citep{boeckDataVizDecisionmakingAction2020} as well as when the colour scale did not change\citep{sandalowChiTribGraphicsPleaseUpdate2020}, illustrating the difficulty of designing visualisations amid competing constraints and a constantly evolving pandemic.
 
@@
 
-The most obvious time-series plot is also the most common: plotting cases against date, often with a 7 or 14-day smooth to handle periodicity in case reporting due to the work week, as shown in \Cref{fig:wapo-COVID-time-series}.
+The most obvious time series plot is also the most common: plotting cases against date, often with a 7 or 14-day smooth to handle periodicity in case reporting due to the work week, as shown in \Cref{fig:wapo-COVID-time-series}.
 This allows for individuals to assess the current situation and easily compare the current number of cases to past times for which the individual has a direct reference for what to expect in impact to daily life.
@@
 \includegraphics[width=.5\linewidth]{wapo-COVID-context}
-\caption{Some outlets provide contextual information to assist readers with interpreting time-series plots \citep{CoronavirusCasesTracking2022}.}
+\caption{Some outlets provide contextual information to assist readers with interpreting time series plots \citep{CoronavirusCasesTracking2022}.}
 \label{fig:wapo-context}
@@
 
-However, there are some limitations with this presentation: if we are interested in showing the change in counts over time, viewers must assess the slope of the line, rather than its position.
+However, there are some limitations with this presentation; if we are interested in showing the change in counts over time, viewers must assess the slope of the line, rather than its position.
 We know that slope judgments are much less accurate than position \citep{clevelandGraphicalPerceptionVisual1987}, even in relatively simple situations; when we add in our ability to assess exponential growth, our perceptual accuracy is even more suspect.
@@
 
-Other time-series representations were more abstract, sacrificing exact data representations for a higher-level summary comparing regions and points in time without the burden of numerical calculations.
+Other time series representations were more abstract, sacrificing exact data representations for a higher-level summary comparing regions and points in time without the burden of numerical calculations.
 Several versions from the New York Times are shown in \Cref{fig:sparklines-heatmap-nyt}.
-The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data, however, they do not present precise numerical information and even the color scale is (perhaps intentionally) opaque -- it would be extremely difficult to translate the visualization into any sort of accurate numerical estimate.
+The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data, however, they do not present precise numerical information and even the colour scale is (perhaps intentionally) opaque -- it would be extremely difficult to translate the visualisation into any sort of accurate numerical estimate.
 In addition, by hiding these details from the user, it is very difficult to identify when the underlying mathematical representation changes over time, as it did several times during the summer of 2020.
-The use of a similar color scheme as well as similar geometric representations made it extremely difficult for users to identify that the value represented had changed.
+The use of a similar colour scheme as well as similar geometric representations made it extremely difficult for users to identify that the value represented had changed.
 Clearly, the purpose of these charts is not to provide numerical precision, but rather a comparative assessment of the status of one region relative to others as well as relative to previous time points.
@@
 \end{subfigure}\hfill
-\caption{A series of screenshots of the NY Times state-by-state display showing evolving color schemes and even metrics; initially, a discrete scale of two-week change in cases is shown; then, a similar color scheme is used with a different metric.
-Around July 17, the color scheme and variable shown changed to weekly cases per capita (from "fewer" to "more"), and by August 14, the variable shown changed back again ("falling" to "rising") while the color scheme stayed the same.}\label{fig:sparklines-heatmap-nyt}
+\caption{A series of screenshots of the NY Times state-by-state display showing evolving colour schemes and even metrics; initially, a discrete scale of two-week change in cases is shown; then, a similar colour scheme is used with a different metric.
+Around July 17, the colour scheme and variable shown changed to weekly cases per capita (from "fewer" to "more"), and by August 14, the variable shown changed back again ("falling" to "rising") while the colour scheme stayed the same.}\label{fig:sparklines-heatmap-nyt}
 \end{figure}
 
-One unique time-series plot created a representation of the case counts in the frequency domain: instead of plotting cases as they occurred, the chart instead shows line segments from 0 to N in $y$, where $N$ is a predefined number of cases; then the chart resets to show the next angle.
+One unique time series plot created a representation of the case counts in the frequency domain; instead of plotting cases as they occurred, the chart instead shows line segments from 0 to N in $y$, where $N$ is a predefined number of cases; then the chart resets to show the next angle.
 This produces a sense of intensity of the waves of the pandemic over time as shown in \Cref{fig:freq-domain}.
@@
 Here, each 1000 deaths are represented by a single line segment whose slope represents the time taken to reach the next 1000 deaths.
-This provides viewers with a sense of the pace of the epidemic, rather than the raw case counts in more standard time-series representations.
-This representation can also be shown in the auditory domain, providing access to those who are typically excluded from visualizations due to vision loss \citep{vuillemotSonificationCOVID19Data2020}.}
+This provides viewers with a sense of the pace of the epidemic, rather than the raw case counts in more standard time series representations.
+This representation can also be shown in the auditory domain, providing access to those who are typically excluded from visualisations due to vision loss \citep{vuillemotSonificationCOVID19Data2020}.}
 \label{fig:freq-domain}
@@
 
-There are many different ways that designers can use time-series data to provide additional contextual information and facilitate comparisons.
+There are many different ways that designers can use time series data to provide additional contextual information and facilitate comparisons.
 \Cref{fig:ft-waves} shows COVID cases, positivity rate, and hospital admissions from South Africa's Gauteng province, which was one of the first areas to experience a large wave of the Omicron variant.
@@
 
-While traditionally, time-series information has been limited to charts with time on a linear x-axis, many charts manipulated time to highlight different aspects of the pandemic.
-In some cases, these modifications provided large amounts of utility for a relatively small increase in complexity: for example, John Burn-Murdoch's decision to index Financial Times COVID graphs to the days since the $N$th case or death.
-This decision is a small modification of the basic time-series plot, but has important consequences in the chart's readability as well as the visual emphasis for users: the graphs in the Financial Times were designed to emphasize "...
+While traditionally, time series information has been limited to charts with time on a linear x-axis, many charts manipulated time to highlight different aspects of the pandemic.
+In some cases, these modifications provided large amounts of utility for a relatively small increase in complexity; for example, John Burn-Murdoch's decision to index Financial Times COVID graphs to the days since the $N$th case or death.
+This decision is a small modification of the basic time series plot, but has important consequences in the chart's readability as well as the visual emphasis for users: the graphs in the Financial Times were designed to emphasize "...
 that there's an inevitability about how coronavirus spreads...
@@
 Other designers used the capabilities of web graphics to show the time component in the data through animation.
-In some cases, this took the form of animated bubble plots XXX add XXX, while others dispensed with the traditional form of a line graph altogether, showing rising case counts using a "bar chart race": an animated series of bar charts over time that show how case counts in each country are increasing.
+In some cases, this took the form of animated bubble plots XXX add XXX, while others dispensed with the traditional form of a line graph altogether, showing rising case counts using a "bar chart race"; an animated series of bar charts over time that show how case counts in each country are increasing.
 A typical example can be found on YouTube\citep{clockCoronavirusCasesBar2022}; a screenshot is shown in \Cref{fig:barchart-race}.
@@
 In the New York times, this took the form of a story showing the accumulation of the first 100,000 deaths due to COVID in the US, with occasional short quotes from individuals' obituaries to humanize the names\citep{barryRemembering1000002020}.
-While this is not a typical time-series chart, as the user scrolled through the page the pace of the names increases, showing the scale of the pandemic in a visceral way.
-A less sentimental dot-density timeline by the New York Times about 9 months later chronicled the nation reaching 450,000 deaths; the difference between the two visualizations brings to mind the quote oft attributed to Stalin: "One death is a tragedy; a million deaths is a statistic" \citep{investigatorSingleDeathTragedy2016}.
+While this is not a typical time series chart, as the user scrolled through the page the pace of the names increases, showing the scale of the pandemic in a visceral way.
+A less sentimental dot-density timeline by the New York Times about 9 months later chronicled the nation reaching 450,000 deaths; the difference between the two visualisations brings to mind the quote oft attributed to Stalin: "One death is a tragedy; a million deaths is a statistic" \citep{investigatorSingleDeathTragedy2016}.
 
 A slightly different approach by the BBC's Visual and Data Journalism team, shown in \Cref{fig:bbc-flower}, displayed the COVID cases and deaths as a growing flower, with the stem proportional to the number of cases and the flower petals showing the deaths over time\citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}.
-The visualization comes with sound as well, an attempt to make the pandemic an auditory experience as well as a visual one.
+The visualisation comes with sound as well, an attempt to make the pandemic an auditory experience as well as a visual one.
 The form, animation, and auditory experience are all indicative of an emotional appeal \citep{kostelnick_re-emergence_2016,dignazio_data_2020}; with the flower representation as an explicitly recognized symbol of grief; it is clearly not about showing specific case counts and deaths numerically.
@@
 
-The plague is always the others: This is the short formula for dealing with infectious diseases from a historical perspective \citep{thiessen2021}.
+"The plague is always the others." This is the short formula for dealing with infectious diseases from a historical perspective \citep{thiessen2021}.
 Closing the borders and restricting access to the country became a popular means of controlling the spread of the disease at several different points during the pandemic.
@@
 \Cref{fig:policy-comparison} shows a chart from an article which appeared in The Federalist, a conservative outlet in the United States, suggesting that mask mandates are ineffective because it is difficult to spot the impact of the mask mandate on the overall trend of cases.
-Of course, mask mandates are but one component of a much broader pandemic management strategy, but the goal of the chart is clear: the reader is supposed to conclude that cases and masks are not associated.
+Of course, mask mandates are but one component of a much broader pandemic management strategy, but the goal of the chart is clear; the reader is supposed to conclude that cases and masks are not associated.
 These charts were so common on social media that media outlets ran stories identifying them as misinformation \citep{reutersstaffFactCheckMask2020}; clearly, comparative charts were effectively deployed to misinform as well as to inform.
@@
 Initially, log scales were incredibly useful at showing case counts, because minimal mitigation measures were in place and the growth of case counts (or presumptive positive cases, in absence of available testing) was fairly close to exponential.
-In addition, the use of log scales allowed for the comparison of total cases across entities with large population differences: in the US, we could compare cases in New York and California with cases in Michigan and Washington, even though the populations (and corresponding case counts) of Michigan and Washington are much lower than the populations (and case counts) of either New York or California.
+In addition, the use of log scales allowed for the comparison of total cases across entities with large population differences; in the US, we could compare cases in New York and California with cases in Michigan and Washington, even though the populations (and corresponding case counts) of Michigan and Washington are much lower than the populations (and case counts) of either New York or California.
 
@@
 % Problems with log scales
-However, after the first wave of COVID, the issues with log scales became more apparent: it was difficult to detect slight increases in case counts that indicated the beginning of a new wave amid a background level of spread, as demonstrated in  \Cref{fig:log-scale-failures}.
+However, after the first wave of COVID, the issues with log scales became more apparent; it was difficult to detect slight increases in case counts that indicated the beginning of a new wave amid a background level of spread, as demonstrated in  \Cref{fig:log-scale-failures}.
 Diagonal reference lines from the origin were also less helpful, as the growth of cases or deaths was no longer approximately exponential and varied over time; for these reference lines to be effective there would need to be a clear idea of when the case counts started to increase exponentially, which is difficult to determine whilst in the thick of a potential COVID wave.
@@
 Guide lines, like those used in the Financial Times, were helpful in the initial phases of the pandemic, but no outlets that we are aware of shifted the lines to show the growth rate of the current peak (as opposed to the initial peak), and due to the issues shown in \Cref{fig:log-scale-failures}, these guide lines may not have been all that successful in any case.
-Those hoping to create successful time-series charts that allow forecasting are thus left with two options: deal with our inability to forecast exponential growth, or deal with most individuals' inability to translate log scales into practical reality.
+Those hoping to create successful time series charts that allow forecasting are thus left with two options: deal with our inability to forecast exponential growth, or deal with most individuals' inability to translate log scales into practical reality.
 
@@
 The visual narratives of the pandemic which have persisted over time are due in part to the rise of data journalism and interactive graphics platforms which ensure that every outlet has the ability to host engaging and visually appealing graphics; however, while these graphics are nice to look at, not all are equally functional or useful for communication purposes.
-Thus, analyzing the rich collection of visualizations produced as part of the pandemic provides a unique opportunity to examine how these graphics were (and were not) effective, as well as the design decisions made during their creation.
+Thus, analyzing the rich collection of visualisations produced as part of the pandemic provides a unique opportunity to examine how these graphics were (and were not) effective, as well as the design decisions made during their creation.
 As in any situation, good tools can be used and misused freely; we have seen charts and graphics used to perpetuate misinformation and misleading claims, even though it is far more common for charts and graphs to be used for good - to educate and inform the public.
@@
 
-The pandemic and graphics used to show the ebb and flow of COVID cases highlight the challenges of data visualization in the modern era: even with numerous guidelines and standards for graphics, it is not straightforward to communicate about complex data, and creating charts which fully represent the intended message or messages is a challenge.
+The pandemic and graphics used to show the ebb and flow of COVID cases highlight the challenges of data visualisation in the modern era.
+Even with numerous guidelines and standards for graphics, it is not straightforward to communicate about complex data, and creating charts which fully represent the intended message or messages is a challenge.
 It is essential that we continue to study how graphics are perceived and interpreted, supporting the guidelines and standards which exist with user studies that ground these rules in science.

### 2022-10-27 JDSSV_manuscript_submitted.tex
copyediting changes

@@
-\documentclass[article]{jdssv}\usepackage[]{graphicx}\usepackage[]{xcolor}
+\documentclass[article]{jdssv}\usepackage[]{graphicx}\usepackage[]{color}
 % maxwidth is the original width if it is less than linewidth

### 2022-10-27 JDSSV_manuscript_submitted.Rnw
copyediting_secondRound

@@
 In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic to showcase their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods.
-COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles but also Data journalism will certainly be one of the beneficiaries of the COVID-19 pandemic and it has become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards\citep{kochWelcomeRevolutionCOVID192021}.
+COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles but also Data journalism will certainly be one of the beneficiaries of the COVID-19 pandemic and it has become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards \citep{kochWelcomeRevolutionCOVID192021}.
 
@@
 The symbol, most commonly a disk or a square, is scaled such that its area corresponds to the data value of the region. 
-Proportional symbol maps sacrifice familiar shapes and topological connectedness for equal visual weight\citep{gaoUsabilityValuebyalphaMaps2019}; while this representation avoids the area bias, proportional symbol maps are subject to under-estimation biases \citep{shimAnalysisMiddleSchool2008} that may be of particular concern when visualizing epidemic disease spread.
+Proportional symbol maps sacrifice familiar shapes and topological connectedness for equal visual weight \citep{gaoUsabilityValuebyalphaMaps2019}; while this representation avoids the area bias, proportional symbol maps are subject to under-estimation biases \citep{shimAnalysisMiddleSchool2008} that may be of particular concern when visualizing epidemic disease spread.
 In addition, the visual impression of any map based illustration depends on the projection chosen for the underlying map; the examples in \Cref{fig:choro1} and \Cref{fig:propsymb} have a western-centric bias.
@@
 Various algorithms can be used to modify geographic boundaries to maintain spatial continuity, or shapes such as hexagons or squares can be used to represent geographic entities (as in \Cref{fig:hex-cartogram}); with either choice, geometry, topology, or both must be sacrificed.
-While these methods are more popular in the academic visualisation literature, including analysis of the spread of COVID-19\citep{yalcinMappingGlobalSpatiotemporal2022}, they were not overly common in the media for global representations of COVID information.
+While these methods are more popular in the academic visualisation literature, including analysis of the spread of COVID-19 \citep{yalcinMappingGlobalSpatiotemporal2022}, they were not overly common in the media for global representations of COVID information.
 This may be because cartograms require more visual effort on the part of the viewer; in addition, news organizations may be reluctant to assume their readers have the geographic sophistication necessary to translate between the distorted map of the globe as drawn on the cartogram and the undistorted mental map of the globe used for mental comparison purposes.
@@
 When scales automatically adjust, mental comparisons across time can become misleading, because the colour scale from yesterday does not mean the same thing as it did today, as the range has expanded and the colours have stayed the same.
-It is not that often that day-to-day design changes (or lack thereof) cause fundamental problems with the interpretation of data visualisation, but during the COVID 19 pandemic this situation was both extremely common and had real consequences: people would glance at the current COVID risk map and assume that "red" meant the same thing that it did yesterday, despite the fact that the underlying values had changed\citep{abiadColorDiagramWHOLebanon2020,rubelHeyUpshotNYTCurious2020,matthewsLessonFutzingData2020}.
+It is not that often that day-to-day design changes (or lack thereof) cause fundamental problems with the interpretation of data visualisation, but during the COVID 19 pandemic this situation was both extremely common and had real consequences: people would glance at the current COVID risk map and assume that "red" meant the same thing that it did yesterday, despite the fact that the underlying values had changed \citep{abiadColorDiagramWHOLebanon2020,rubelHeyUpshotNYTCurious2020,matthewsLessonFutzingData2020}.
 One example of this issue over two months of NY Times graphics is shown in \Cref{fig:sparklines-heatmap-nyt}; the overall colour scheme used changes once, but the variable displayed changes (at least) 4 times over the course of a two month period.
@@
 The magnitude of the problem also depends on whether the scale uses a set of bins (a discrete mapping to a continuous variable) or uses a continuous colour mapping; in some cases, outlets switched between the two options over time or changed the underlying unit entirely.
-Interestingly, media outlets and municipal dashboards were subject to criticism when the colour scales changed in response to increasing case counts\citep{boeckDataVizDecisionmakingAction2020} as well as when the colour scale did not change\citep{sandalowChiTribGraphicsPleaseUpdate2020}, illustrating the difficulty of designing visualisations amid competing constraints and a constantly evolving pandemic.
+Interestingly, media outlets and municipal dashboards were subject to criticism when the colour scales changed in response to increasing case counts \citep{boeckDataVizDecisionmakingAction2020} as well as when the colour scale did not change \citep{sandalowChiTribGraphicsPleaseUpdate2020}, illustrating the difficulty of designing visualisations amid competing constraints and a constantly evolving pandemic.
 
@@
 that there's an inevitability about how coronavirus spreads...
-even if there are only a few cases in your country today, based on all the data we have, you will end up going along that path, the same path that the likes of Italy and Spain have been on so tragically."\citep{hannenCoronavirusTrajectoryTracker2020a} By placing all countries on an even footing divorced from the temporal randomness of when the first case appeared, the viewer can effectively compare the effect of government policies and other interventions on the trajectory over time\citep{burn-murdochAllCountriesDeath2020,burn-murdochLatestCaseTrajectories2020}.
+even if there are only a few cases in your country today, based on all the data we have, you will end up going along that path, the same path that the likes of Italy and Spain have been on so tragically" \citep{hannenCoronavirusTrajectoryTracker2020a}.
+By placing all countries on an even footing divorced from the temporal randomness of when the first case appeared, the viewer can effectively compare the effect of government policies and other interventions on the trajectory over time \citep{burn-murdochAllCountriesDeath2020,burn-murdochLatestCaseTrajectories2020}.
 
@@
 In some cases, this took the form of animated bubble plots XXX add XXX, while others dispensed with the traditional form of a line graph altogether, showing rising case counts using a "bar chart race"; an animated series of bar charts over time that show how case counts in each country are increasing.
-A typical example can be found on YouTube\citep{clockCoronavirusCasesBar2022}; a screenshot is shown in \Cref{fig:barchart-race}.
+A typical example can be found on YouTube \citep{clockCoronavirusCasesBar2022}; a screenshot is shown in \Cref{fig:barchart-race}.
 This type of chart is effective in showing instantaneous changes in counts over days for countries with an extremely high number of cases, but may make it difficult for the viewer to process how cases are changing in countries that do not dominate case counts.
@@
 As people grappled with the scale of the pandemic and the lives lost as a result, a different set of data displays attempted to provide context to this loss over time.
-In the New York times, this took the form of a story showing the accumulation of the first 100,000 deaths due to COVID in the US, with occasional short quotes from individuals' obituaries to humanize the names\citep{barryRemembering1000002020}.
+In the New York times, this took the form of a story showing the accumulation of the first 100,000 deaths due to COVID in the US, with occasional short quotes from individuals' obituaries to humanize the names \citep{barryRemembering1000002020}.
 While this is not a typical time series chart, as the user scrolled through the page the pace of the names increases, showing the scale of the pandemic in a visceral way.
@@
 
-A slightly different approach by the BBC's Visual and Data Journalism team, shown in \Cref{fig:bbc-flower}, displayed the COVID cases and deaths as a growing flower, with the stem proportional to the number of cases and the flower petals showing the deaths over time\citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}.
+A slightly different approach by the BBC's Visual and Data Journalism team, shown in \Cref{fig:bbc-flower}, displayed the COVID cases and deaths as a growing flower, with the stem proportional to the number of cases and the flower petals showing the deaths over time \citep{thebbcvisualanddatajournalismteamCoronavirusHowCan2020}.
 The visualisation comes with sound as well, an attempt to make the pandemic an auditory experience as well as a visual one.
@@
 
-Often, these comparative charts were combined with policy discussions, with individuals challenged to spot certain policy interventions on the case-count graphs of different localities with different policies\citep{weissThese12Graphs2020}.
+Often, these comparative charts were combined with policy discussions, with individuals challenged to spot certain policy interventions on the case-count graphs of different localities with different policies \citep{weissThese12Graphs2020}.
 \Cref{fig:policy-comparison} shows a chart from an article which appeared in The Federalist, a conservative outlet in the United States, suggesting that mask mandates are ineffective because it is difficult to spot the impact of the mask mandate on the overall trend of cases.
@@
 As noted previously, however, interpreting log scales requires levels of numerical sophistication that may not be appropriate for the general public.
-Even researchers do not always read and interpret log scales correctly\citep{mengeLogarithmicScalesEcological2018}; expecting the general public to do so is difficult under normal circumstances\citep{hecklerStudentAccuracyReading2013} is difficult.
+Even researchers do not always read and interpret log scales correctly \citep{mengeLogarithmicScalesEcological2018}; expecting the general public to do so is difficult under normal circumstances \citep{hecklerStudentAccuracyReading2013} is difficult.
 When panic, fear, uncertainty, and doubt about the situation are added to the mix, it is easy to imagine that we become even worse when interpreting graphics.
@@
 Graphics included in this paper not sourced from media outlets were created using data published by the New York Times at \url{https://github.com/nytimes/covid-19-data/}.
-Graphics were created using ggplot2\citep{ggplot2} and \proglang{R}~4.1.2.
+Graphics were created using ggplot2 \citep{ggplot2} and \proglang{R}~4.1.2.
 \proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.

### 2022-10-27 JDSSV_manuscript_submitted.Rnw
final copyediting

@@
 %% - \AND starts a new line, \And does not.
-\author{
-   \And }
-\Plainauthor{}
+\author{Susan VanderPlas\\
+\affiliation{Statistics Department\\ University of Nebraska Lincoln\\ United States}
+   \And Adalbert F.X. Wilhelm \\ \affiliation{School for Business\\ Social and Decision Sciences\\ Jacobs University Bremen}}
+\Plainauthor{Susan VanderPlas, Adalbert F.X. Wilhelm}
 
@@
 %%   (in the same first line, separated by comma).
-\Address{
+\Address{Adalbert F.X. Wilhelm\\
+School for Business, Social and Decision Sciences\\
+Jacobs University Bremen\\
+Campus Ring 1\\
+28759 Bremen\\
+Germany\\
+e-mail:a.wilhelm@jacobs-university.de
 }
@@
 %So there are several goals to be achieved by visualising COVID-19 data.
-These goals and their priorities were adapted as dynamically as the virus mutated and the pandemic changed pace, but as with any other data visualisation the general principles remain the same: ensuring clear understanding by organising graphics in such a way that the story of the data is told most effectively.
+These goals and their priorities were adapted as dynamically as the virus mutated and the pandemic changed pace, but as with any other data visualisation, the general principles remain the same; ensuring clear understanding by organising graphics in such a way that the story of the data is told most effectively.
 
@@
 In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic to showcase their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods.
-COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles but also Data journalism will certainly be one of the beneficiaries of the COVID-19 pandemic and it has become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards \citep{kochWelcomeRevolutionCOVID192021}.
+COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles.
+Hence data journalism will certainly be one of the beneficiaries of the COVID-19 pandemic and has already become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards \citep{kochWelcomeRevolutionCOVID192021}.
 
@@
 
-At the same time, we still see a lot of defective graphics disseminated and shared: some that violate fundamental statistical and visualisation design principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility.
+At the same time, we still see a lot of defective graphics disseminated and shared; some that violate fundamental statistical and visualisation design principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility.
 These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others \citep{EQUATORNetworkEnhancing}.
-Efforts to improve the communication of quantitative and statistical information date from the beginning mass media and the ability to create charts and tables from data systematically, and while there continue to be efforts made in this area, historical papers outline many of the same issues that we still struggle with today \citep{ASA-standards,haemerPresentationProblemsArea1949,kovermanPLANNINGVISUALPRESENTATION1961,fienbergGraphicalMethodsStatistics1979,Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.
-In other words, while our **technology** for recording, visually summarizing, and communicating information distilled from data has vastly improved over the past 100 years, we still face many of the same problems when **designing** charts which communicate data accurately and effectively.
+The use of numbers, their graphic representation and the systematic creation of charts and tables from data in the mass media have been accompanied from the beginning by efforts to improve the communication of quantitative and statistical information.
+And while efforts continue to be made in this area, historical documents describe many of the same problems we still struggle with today \citep{ASA-standards,haemerPresentationProblemsArea1949,kovermanPLANNINGVISUALPRESENTATION1961,fienbergGraphicalMethodsStatistics1979,Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.
+In other words, while our \textit{technology} for recording, visually summarising, and communicating information distilled from data has vastly improved over the past 100 years, we still face many of the same problems when \textit{designing} charts which communicate data accurately and effectively.
 
@@
 Having briefly discussed the media landscape surrounding the COVID pandemic and goals for communicating about data effectively, we should also briefly discuss the choices that go into creating visual representations of data intended to convey quantitative information.
-While the overall goal of a data visualisation is to tell the story of the data clearly and effectively, how this story is conveyed involves a series of design choices: which variables are shown, what limits are imposed, what visual forms are used to represent the data?
+While the overall goal of a data visualisation is to tell the story of the data clearly and effectively, how this story is conveyed involves a series of design choices: Which variables are shown, What limits are imposed, What visual forms are used to represent the data?
 How will the chart (and any surrounding story, in journalism) be presented to the viewer? 
-As the designer makes these choices, they have in mind an audience, and the audience's assumed characteristics affect the design choices as well: language, education level, likely medium used to access the visualisation (computer, physical newspaper, tablet, cell phone) - all of these considerations also impact the design of the chart and what ideas the designer considers during the creative process.
+As the designer makes these choices, they have in mind an audience, and the audience's assumed characteristics affect the design choices as well: language, education level, likely medium used to access the visualisation (computer, physical newspaper, tablet, cell phone) -- all of these considerations also impact the design of the chart and what ideas the designer considers during the creative process.
 
-Once the basic form of the chart is determined, there are additional considerations that may be specified by the stylistic conventions of different news organizations, but may also be deliberately chosen for visual effect: colour palettes, fonts, spatial layouts, and annotations may be used to highlight information, convey additional points of a narrative, or provide helpful information about how to read an unfamiliar chart.
-While each of these design choices could be (and have been) the subject of  additional paper(s), it is important to acknowledge that the graphics we analyze in this paper did not simply pop into being - they are the product of a series of decisions, some consciously made and others imposed by convention or technological capabilities.
+Once the basic form of the chart is determined, there are additional considerations that may be specified by the stylistic conventions of different news organisations, but may also be deliberately chosen for visual effect: colour palettes, fonts, spatial layouts, and annotations may be used to highlight information, convey additional points of a narrative, or provide helpful information about how to read an unfamiliar chart.
+While each of these design choices could be (and have been) the subject of  additional paper(s), it is important to acknowledge that the graphics we analyse in this paper did not simply pop into being -- they are the product of a series of decisions, some consciously made and others imposed by convention or technological capabilities.
 We will highlight particular choices which were effective (or not) throughout this paper, but we will also leave some undiscussed for the sake of brevity.
@@
 \subsection{Scope}
-Our focus in this paper is to highlight, analyze, and discuss a selection of charts and graphics which caught our attention during the initial, real-time experience of the pandemic.
+Our focus in this paper is to highlight, analyse, and discuss a selection of charts and graphics which caught our attention during the initial, real-time experience of the pandemic.
 As we primarily consume media from western Europe and the United States, our assessment is primarily limited to news outlets catering to these audiences; a retrospective global analysis of COVID graphics would certainly be interesting, but is beyond the scope of this work.
@@
 On 11 March 2020, WHO declared the outbreak of the novel coronavirus disease (COVID-19) a pandemic, and since that date at the latest, the global distribution of the disease has been in the public eye.
-The spatial spread of the virus and the resulting cases and deaths during this initial period were commonly visualized by choropleth maps, see for example \Cref{fig:choro1} showing the total number of infections reported in each country as of January 14, 2022.
+The spatial spread of the virus and the resulting cases and deaths during this initial period were commonly visualised by choropleth maps, see for example \Cref{fig:choro1} showing the total number of infections reported in each country as of January 14, 2022.
 A central element of the narrative during March 2020 was the ubiquity of the disease and the accompanying global impact.
@@
 The symbol, most commonly a disk or a square, is scaled such that its area corresponds to the data value of the region. 
-Proportional symbol maps sacrifice familiar shapes and topological connectedness for equal visual weight \citep{gaoUsabilityValuebyalphaMaps2019}; while this representation avoids the area bias, proportional symbol maps are subject to under-estimation biases \citep{shimAnalysisMiddleSchool2008} that may be of particular concern when visualizing epidemic disease spread.
-In addition, the visual impression of any map based illustration depends on the projection chosen for the underlying map; the examples in \Cref{fig:choro1} and \Cref{fig:propsymb} have a western-centric bias.
+Proportional symbol maps sacrifice familiar shapes and topological connectedness for equal visual weight \citep{gaoUsabilityValuebyalphaMaps2019}; while this representation avoids the area bias, proportional symbol maps are subject to under-estimation biases \citep{shimAnalysisMiddleSchool2008} that may be of particular concern when visualising epidemic disease spread.
+In addition, the visual impression of any map-based illustration depends on the projection chosen for the underlying map; the examples in \Cref{fig:choro1} and \Cref{fig:propsymb} have a western-centric bias.
 
@@
 While these methods are more popular in the academic visualisation literature, including analysis of the spread of COVID-19 \citep{yalcinMappingGlobalSpatiotemporal2022}, they were not overly common in the media for global representations of COVID information.
-This may be because cartograms require more visual effort on the part of the viewer; in addition, news organizations may be reluctant to assume their readers have the geographic sophistication necessary to translate between the distorted map of the globe as drawn on the cartogram and the undistorted mental map of the globe used for mental comparison purposes.
+This may be because cartograms require more visual effort on the part of the viewer; in addition, news organisations may be reluctant to assume their readers have the geographic sophistication necessary to translate between the distorted map of the globe as drawn on the cartogram, and the undistorted mental map of the globe used for mental comparison purposes.
 A spatial dot density map might have allowed for analysis of data across geographic boundaries, but this approach was not commonly used to show global case counts, though it was highlighted as a useful option at the national level \citep{fieldMappingCoronavirusResponsibly}, as shown in \Cref{fig:china-dot-map}.
 
-The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe: none of the above viusalisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them.
-The use of maps to assess the global pandemic more often focuses on the comparative aspects: how are we doing as opposed to our neighbors? Which strategy to fight COVID-19 is better? 
+The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe; none of the above visualisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them.
+The use of maps to assess the global pandemic more often focuses on the comparative aspects: How are we doing as opposed to our neighbors? Which strategy to fight COVID-19 is better? 
 We will examine these topics more closely in \Cref{sec:rankings}.
@@
 When scales automatically adjust, mental comparisons across time can become misleading, because the colour scale from yesterday does not mean the same thing as it did today, as the range has expanded and the colours have stayed the same.
-It is not that often that day-to-day design changes (or lack thereof) cause fundamental problems with the interpretation of data visualisation, but during the COVID 19 pandemic this situation was both extremely common and had real consequences: people would glance at the current COVID risk map and assume that "red" meant the same thing that it did yesterday, despite the fact that the underlying values had changed \citep{abiadColorDiagramWHOLebanon2020,rubelHeyUpshotNYTCurious2020,matthewsLessonFutzingData2020}.
+It is not that often that day-to-day design changes (or lack thereof) cause fundamental problems with the interpretation of data visualisation, but during the COVID 19 pandemic, this situation was both extremely common and had real consequences; people would glance at the current COVID risk map and assume that "red" meant the same thing that it did yesterday, despite the fact that the underlying values had changed \citep{abiadColorDiagramWHOLebanon2020,rubelHeyUpshotNYTCurious2020,matthewsLessonFutzingData2020}.
 One example of this issue over two months of NY Times graphics is shown in \Cref{fig:sparklines-heatmap-nyt}; the overall colour scheme used changes once, but the variable displayed changes (at least) 4 times over the course of a two month period.
@@
 
-Early in the pandemic, as case numbers grew exponentially without appropriate public health measures to restrain them, the scale problem for choropleth maps was particularly complex: over time either the colour mapping changes or the map becomes useless due to compression against the upper limits of the scale, as in \citet{panchadsaramIfWeAre2020}.
+Early in the pandemic, as case numbers grew exponentially without appropriate public health measures to restrain them, the scale problem for choropleth maps was particularly complex; over time either the colour mapping changes or the map becomes useless due to compression against the upper limits of the scale, as in \citet{panchadsaramIfWeAre2020}.
 The magnitude of the problem also depends on whether the scale uses a set of bins (a discrete mapping to a continuous variable) or uses a continuous colour mapping; in some cases, outlets switched between the two options over time or changed the underlying unit entirely.
@@
 Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time, as there are new developments on a daily basis.
-Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis, as in \Cref{fig:jbm-tweet} and \Cref{fig:wapo-COVID-time-series}. 
+Typically, this means that cases, hospitalisations, and/or deaths are plotted on the y-axis, with date shown on the x-axis, as in \Cref{fig:jbm-tweet} and \Cref{fig:wapo-COVID-time-series}. 
 This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state. 
@@
 
-The most obvious time series plot is also the most common: plotting cases against date, often with a 7 or 14-day smooth to handle periodicity in case reporting due to the work week, as shown in \Cref{fig:wapo-COVID-time-series}.
+The most obvious time series plot is also the most common; plotting cases against date, often with a 7 or 14-day smooth to handle periodicity in case reporting due to the work week, as shown in \Cref{fig:wapo-COVID-time-series}.
 This allows for individuals to assess the current situation and easily compare the current number of cases to past times for which the individual has a direct reference for what to expect in impact to daily life.
-Often, as in \Cref{fig:wapo-COVID-time-series}, these time series show multiple variables: cases, deaths, and hospitalizations (and sometimes vaccinations) to provide a more comprehensive view of the situation; this is particularly useful as there has been some decoupling of case numbers and hospitalizations/deaths with the emergence of effective vaccinations and strains which are more contagious but potentially also less severe.
+Often, as in \Cref{fig:wapo-COVID-time-series}, these time series show multiple variables: cases, deaths, and hospitalisations (and sometimes vaccinations), to provide a more comprehensive view of the situation.
+This is particularly useful as there has been some decoupling of case numbers and hospitalisations/deaths with the emergence of effective vaccinations and strains which are more contagious but potentially also less severe.
 Many outlets also provide additional charts designed to provide some immediate context to readers, as shown in \Cref{fig:wapo-context}.
@@
 \caption{COVID Cases, positivity rate, and hospital admissions from South Africa's Gauteng province, on a log scale (top) and linear scale (bottom).
-Successive waves are overlaid, showing that the increase in cases due to the Omicron variant was sharper than any previously measured increase, while hospitalizations were much more comparable to previous waves \citep{burn-murdoch_new_2021-1}.}
+Successive waves are overlaid, showing that the increase in cases due to the Omicron variant was sharper than any previously measured increase, while hospitalisations were much more comparable to previous waves \citep{burn-murdoch_new_2021-1}.}
 \label{fig:ft-waves}
@@
 Other designers used the capabilities of web graphics to show the time component in the data through animation.
-In some cases, this took the form of animated bubble plots XXX add XXX, while others dispensed with the traditional form of a line graph altogether, showing rising case counts using a "bar chart race"; an animated series of bar charts over time that show how case counts in each country are increasing.
+In some cases, this took the form of animated bubble plots, while others dispensed with the traditional form of a line graph altogether, showing rising case counts using a "bar chart race"; an animated series of bar charts over time that show how case counts in each country are increasing.
 A typical example can be found on YouTube \citep{clockCoronavirusCasesBar2022}; a screenshot is shown in \Cref{fig:barchart-race}.
@@
 As people grappled with the scale of the pandemic and the lives lost as a result, a different set of data displays attempted to provide context to this loss over time.
-In the New York times, this took the form of a story showing the accumulation of the first 100,000 deaths due to COVID in the US, with occasional short quotes from individuals' obituaries to humanize the names \citep{barryRemembering1000002020}.
+In the New York times, this took the form of a story showing the accumulation of the first 100,000 deaths due to COVID in the US, with occasional short quotes from individuals' obituaries to humanise the names \citep{barryRemembering1000002020}.
 While this is not a typical time series chart, as the user scrolled through the page the pace of the names increases, showing the scale of the pandemic in a visceral way.
@@
 
-If we evaluate the use of log and linear scales under the more general context of exponential (or near-exponential) growth rates, however, we can gain some clarity as to their use in this particular context.
+If we evaluate the use of log and linear scales under the more general context of exponential (or near-exponential) growth rates,  we can gain some clarity as to their use in this particular context.
 We are abysmal at forecasting exponential growth, vastly under-estimating future growth by using linear or quadratic approximations \citep{wagenaarExtrapolationExponentialTime1978a,lawrenceExploringJudgementalForecasting1992,timmersInverseStatisticsMisperception1977a}.
@@
 While not always optimal, these graphics have attempted to provide meaningful context, to encourage individual and collective action, and to help individuals grapple with the scale of the pandemic in cases, deaths, vaccinations, and interventions.
-As in any developing situation, choices made at some points in the pandemic did not always persist - many news outlets refined their charts and approach to the design of graphics over the course of the pandemic.
+As in any developing situation, choices made at some points in the pandemic did not always persist -- many news outlets refined their charts and approach to the design of graphics over the course of the pandemic.
 This evolution of graphical forms has provided us with an opportunity to evaluate what worked and did not -- and why -- in a context of broad general interest.
@@
 The visual narratives of the pandemic which have persisted over time are due in part to the rise of data journalism and interactive graphics platforms which ensure that every outlet has the ability to host engaging and visually appealing graphics; however, while these graphics are nice to look at, not all are equally functional or useful for communication purposes.
-Thus, analyzing the rich collection of visualisations produced as part of the pandemic provides a unique opportunity to examine how these graphics were (and were not) effective, as well as the design decisions made during their creation.
-As in any situation, good tools can be used and misused freely; we have seen charts and graphics used to perpetuate misinformation and misleading claims, even though it is far more common for charts and graphs to be used for good - to educate and inform the public.
+Thus, analysing the rich collection of visualisations produced as part of the pandemic provides a unique opportunity to examine how these graphics were (and were not) effective, as well as the design decisions made during their creation.
+As in any situation, good tools can be used and misused freely; we have seen charts and graphics used to perpetuate misinformation and misleading claims, even though it is far more common for charts and graphs to be used for good -- to educate and inform the public.
 
@@
 The pandemic and graphics used to show the ebb and flow of COVID cases highlight the challenges of data visualisation in the modern era.
-Even with numerous guidelines and standards for graphics, it is not straightforward to communicate about complex data, and creating charts which fully represent the intended message or messages is a challenge.
+Even with numerous guidelines and standards for graphics, it is not straightforward to communicate complex data; creating charts which fully represent the intended message or messages remains a challenge.
 It is essential that we continue to study how graphics are perceived and interpreted, supporting the guidelines and standards which exist with user studies that ground these rules in science.
@@
 
-First ideas have been presented at DSSV 2020. We thank our colleagues, the reviewers, and the editors for valuable feedback and encouragement to complete and revise this manuscript.
+First ideas have been presented at DSSV 2021. We thank our colleagues, the reviewers, and the editors for valuable feedback and encouragement to complete and revise this manuscript.
 

### 2022-10-28 JDSSV_manuscript_submitted.Rnw
Tweaks for copyediting

@@
 The use of numbers, their graphic representation and the systematic creation of charts and tables from data in the mass media have been accompanied from the beginning by efforts to improve the communication of quantitative and statistical information.
-And while efforts continue to be made in this area, historical documents describe many of the same problems we still struggle with today \citep{ASA-standards,haemerPresentationProblemsArea1949,kovermanPLANNINGVISUALPRESENTATION1961,fienbergGraphicalMethodsStatistics1979,Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.
+While efforts continue to be made in this area, historical documents describe many of the same problems we still struggle with today \citep{ASA-standards,haemerPresentationProblemsArea1949,kovermanPLANNINGVISUALPRESENTATION1961,fienbergGraphicalMethodsStatistics1979,Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.
 In other words, while our \textit{technology} for recording, visually summarising, and communicating information distilled from data has vastly improved over the past 100 years, we still face many of the same problems when \textit{designing} charts which communicate data accurately and effectively.
@@
 Yet another approach to showing the temporal component of the pandemic was to leverage web graphics to show the temporal component through animation; this freed up the $x$ axis which in the default configuration would have been devoted to time and allowed another variable to be displayed.
-While this approach was common on social media, it was reserved for specific use-cases and more carefully selected variables by professional media outlets.
+While this approach was common on social media, it was used more selectively, with carefully chosen variables by professional media outlets.
 \Cref{fig:bubble-ft} shows an example from the Financial Times which uses the x-axis to show GDP per capita and the y-axis to show vaccination rate, bubble size indicates a country's population.
@@
 This chart provides context for the Omicron wave, showing that while infections are occurring at a much faster rate than in previous waves, hospital admissions seem to be approximately following previous trends.
-In addition, it is clear that the chart is highlighting the Omicron variant from the colour selection: the red used to show the Omicron numbers is a sharp contrast from the green and blue shades used to show the first three waves of the pandemic.
+In addition, it is clear that the chart is highlighting the Omicron variant from the colour selection; the red used to show the Omicron numbers is a sharp contrast from the green and blue shades used to show the first three waves of the pandemic.
 Notably, even though red and green in combination are usually to be avoided for the sake of colourblind viewers, the shades of blue/green variants which are chosen are light enough that the red stands out even to someone with colour vision deficiency.
@@
 
-First ideas have been presented at DSSV 2021. We thank our colleagues, the reviewers, and the editors for valuable feedback and encouragement to complete and revise this manuscript.
+These ideas were first presented at DSSV 2020. We thank our colleagues, the reviewers, and the editors for valuable feedback and encouragement to complete and revise this manuscript.
 

### 2022-10-28 JDSSV_manuscript_submitted.tex
FinalCopyedit

@@
 
-"The plague is always the others." This is the short formula for dealing with infectious diseases from a historical perspective \citep{thiessen2021}.
+"The plague is always the others" \citep{thiessen2021}.
+This is the short formula for dealing with infectious diseases from a historical perspective.
 Closing the borders and restricting access to the country became a popular means of controlling the spread of the disease at several different points during the pandemic.

### 2022-10-31 JDSSV_manuscript_submitted.Rnw
FinalCopyedit
FinalFinalCopyedit

@@
 %%   (in the same first line, separated by comma).
-\Address{Adalbert F.X. Wilhelm\\
-School for Business, Social and Decision Sciences\\
-Jacobs University Bremen\\
+\Address{
+  Susan Vanderplas\\
+  Department of Statistics\\
+  University of Nebraska–Lincoln\\
+  349A Hardin Hall \\
+  Lincoln, NE 68583-0963, USA\\
+  E-mail: \email{susan.vanderplas@unl.edu}\\
+  URL: \url{https://statistics.unl.edu/susan-vanderplas}\\
+\And\\
+  Adalbert F.X. Wilhelm\\
+  School of Business, Social and Decision Sciences\\
+  Jacobs University Bremen gGmbH\\
   Campus Ring 1\\
-28759 Bremen\\
-Germany\\
-e-mail:a.wilhelm@jacobs-university.de
+  28759 Bremen, Germany\\
+  E-mail: \email{a.wilhelm@jacobs-university.de}\\
+  URL: \url{http://www.jacobs-university.de/directory/wilhelm}
 }
@@
 COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles.
-Hence data journalism will certainly be one of the beneficiaries of the COVID-19 pandemic and has already become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards \citep{kochWelcomeRevolutionCOVID192021}.
+Hence, data journalism will certainly be one of the beneficiaries of the COVID-19 pandemic and has already become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards \citep{kochWelcomeRevolutionCOVID192021}.
 
@@
 
-Once the basic form of the chart is determined, there are additional considerations that may be specified by the stylistic conventions of different news organisations, but may also be deliberately chosen for visual effect: colour palettes, fonts, spatial layouts, and annotations may be used to highlight information, convey additional points of a narrative, or provide helpful information about how to read an unfamiliar chart.
+Once the basic form of the chart is determined, there are additional considerations that may be specified by the stylistic conventions of different news organisations, but may also be deliberately chosen for visual effect; colour palettes, fonts, spatial layouts, and annotations may be used to highlight information, convey additional points of a narrative, or provide helpful information about how to read an unfamiliar chart.
 While each of these design choices could be (and have been) the subject of  additional paper(s), it is important to acknowledge that the graphics we analyse in this paper did not simply pop into being -- they are the product of a series of decisions, some consciously made and others imposed by convention or technological capabilities.
